The ciliary kinesin KIF7 controls the development of the cerebral cortex by acting differentially on SHH-signaling in dorsal and ventral forebrain

Version 1

Abstract

Abstract

Mutations of KIF7, a key ciliary component of Sonic hedgehog (SHH) pathway, are associated in humans with cerebral cortex malformations and clinical features suggestive of cortical dysfunction. KIF7 regulates the processing of GLI-A and GLI3-R transcription factors in a SHH-dependent manner both in humans and mice. Here, we examine the embryonic cortex development of a mouse model that lacks the expression of KIF7 (Kif7 -/-). The cortex is composed of principal neurons generated locally in the dorsal telencephalon where SHH expression is low and inhibitory interneurons (cIN) generated in the ventral telencephalon where SHH expression is high. We observe a strong impact of Kif7 deletion on the dorsal cortex development whose abnormalities resemble those of GLI3R mutants: subplate cells are absent, the intermediate progenitor layer and cortical plate do not segregate properly, and corticofugal axons do not develop timely, leading to a delayed colonization of the telencephalon by thalamo-cortical axons. These structural defects alter the cortical distribution of cIN, which moreover exhibit intrinsic migration defects resembling those of cyclopamine-treated cIN. Our results show that Kif7 deletion impairs the cortex development in multiple ways, exhibiting opposite effects on SHH pathway activity in the developing principal neurons and inhibitory interneurons.

Introduction

The primary cilium is a tiny microtubule-based organelle present on the surface of nearly all mammalian cell types including neurons, which functions as a signaling hub and transduces several signaling pathways comprising Sonic Hedghog (SHH), Wnt, Delta/Notch and mTOR pathways 1. Primary cilium dysfunction causes pleiotropic diseases named ciliopathies. KIF7 is a ciliary kinesin responsible for the trafficking and the positive and negative regulation of the GLI transcription factors in the primary cilium of mammals 2. Loss of function of KIF7 in the mouse has shown that KIF7 regulates (SHH) signalling by acting downstream of Smoothened (Smo) and upstream of GLI2 and GLI3 3–7. Studies in mice established that KIF7 activity is dependent on the expression level of SHH. In the absence of SHH, KIF7 localizes to the base of the primary cilium 5. GLI factors are phosphorylated and addressed to the proteasome at the base of the primary cilium for degradation, leading to the formation of a cleaved and stable transcriptional repressor (GLI3-R). In the presence of SHH, KIF7 accumulates at the distal tip of the primary cilium 4,5 and associates with full length GLI2/3 that become transcriptional activators (GLI-A) 8.

Since 2011, ten studies have identified patients carrying mutations in the KIF7 gene responsible of ciliopathies classified according to clinical features as hydrolethalus, acrocallosal, Joubert and Greig cephalopolysyndactyly syndromes 9–18. MRI investigations revealed macrocephaly, ventricles enlargement and corpus callosum alterations. They also showed the characteristic hindbrain abnormalities observed in most ciliopathies such as molar tooth sign (MTS) and cerebellar atrophy. Patients also presented with mild but frequent cortical malformations as well as neurodevelopmental delay, intellectual disability and seizures 10,13,16,19 which indicate cortical abnormalities.

In physiological conditions, the proper activity of the cortex relies on the excitatory/inhibitory balance i.e. on the ratio, positioning and connectivity of excitatory glutamatergic (principal neurons, PN) and inhibitory GABAergic interneurons (cIN), generated from progenitors in the dorsal and ventral telencephalon, respectively. The cortical neurons of dorsal and ventral origin each migrate towards the developing cortex, the so-called cortical plate (CP), according to a specific timeline allowing the establishment of proper connections in the CP.

SHH plays a central role in establishing the dorso-ventral patterning of the forebrain, and in regulating the proliferation and differentiation of cortical progenitors of dorsal and ventral origins. At early embryonic stage, the ventral expression of SHH orchestrates the ventro-dorsal 20,21 and medio-lateral 22 regionalization of the mouse forebrain and the differentiation of ventral cell types 20. Shh ablation performed after forebrain patterning alters the specification of distinct subgroups of GABAergic interneurons 23–26. Interestingly, conditional ablation of Shh and Smo in the embryonic cortex reduces the proliferation of dorsal progenitors 27, demonstrating a minimal SHH expression in the dorsal telencephalon, even before birth 28. Beside its role on proliferation and specification, SHH also controls the migration of interneurons to the cortical plate 29–33. In the embryonic telencephalon, GLI transcription factors mediate SHH signals in complex and specific ways. The three GLI factors identified in mammals are expressed in the mouse forebrain: GLI1 along the source of SHH in the ganglionic sulcus, and GLI2 and GLI3 dorsally to the SHH source 26,34,35. GLI1 acts uniquely as a pathway activator whereas GLI2 and GLI3 can be processed in transcriptional activators or inhibitors in the primary cilium. However, both GLI1 and GLI2 function primarily as transcriptional activators in response to SHH activity in the ventral forebrain 35. Nevertheless, GLI1 and GLI2 mutants show mild phenotypes 36,37 and GLI2 seems required to transduce high level SHH signals in mice 38. In contrast, the development of the dorsal cortex, where principal excitatory neurons differentiate, depends mainly on the expression of the GLI3-R repressor, revealing low cortical expression of SHH 39,40. Following patterning, Gli3R/A ratio remains critical for specifying the fate of cortical progenitors and regulating cell cycle kinetics 41,42.

Previous studies investigated the mechanisms underlying the corpus callosum agenesis in patients with KIF7 mutation using Kif7 knock-out (Kif7 -/-) mice 43 and the consequence of KIF7 knock down on cortical neurogenesis in principal cells electroporated with Kif7 shRNA44. Nevertheless, the influence of developmental abnormalities associated with ciliopathies on the cortical cytoarchitecture is poorly understood. Given the crucial role of KIF7 in regulating the GLI pathways and the important role of Gli3-R and GLI2/3-A in regulating the early developmental stages of the dorsal and ventral telencephalon, respectively, we examined here the cortical development, the establishment of long distance projections with the thalamus and the migration of GABAergic interneurons (cIN) in Kif7 -/- mice. We focused at embryonic stage E14.5, a key period in corticogenesis when the cortical plate begins to form and cIN start to invade the developing cortex. The influence of abnormal SHH signaling in Kif7 -/- mice on the migratory behavior of cIN was investigated by comparing the migration of cIN in living Kif7 -/- cortical slices and in control slices treated with pharmacological activator and inhibitor of the SHH pathway. We moreover determined the local distribution of the SHH protein in the embryonic cortex.

Our results show developmental defects leading to permanently displaced neurons, abnormal formation of cortical layers and defective cortical circuits that could be responsible for epilepsy and/or intellectual deficiency in patients carrying KIF7 mutation 9–18.

Results

Kif7 knock-out mice as a model to investigate the structural cortical defects that could lead to clinical feature in patients carrying KIF7 mutation

The KIF7 gene is located on chromosome 15 in human and encodes a 1343 aa protein containing in the N-terminal part, a kinesin motor domain and a GLI-binding domain followed by a Coiled-coil region and in the C-terminal part, a Cargo domain able to bind a diverse set of cargos 47. Table 1 summarizes the clinical features associated with mutations targeting either the kinesin or GLI binding domain in the N-terminal part of the protein, the Coiled-coil domain, or the N-terminal Cargo domain. Interestingly, various mutations are associated with the same clinical picture, suggesting that KIF7 mutations, whatever their nature, could lead to protein loss of function, for example by altering the protein structure and solubility as proposed by Klejnot and Kozielski (2012). All patients carrying mutation in the KIF7 gene have developpemental delay (DD) and intellectual deficit (ID) associated with classical defects of ciliopathies such as ventricle enlargement, macrocephaly, corpus callosum (CC) agenesis and molar tooth sign (MTS). Some patients have anatomical cerebral cortex defects as poor frontal development, atrophy, pachygyria and heterotopia 9,17,18,49 that could participate not only to DD and ID but moreover to seizure as observed in a quarter of patients. In this context, we used a murine model in which the Kif7 gene had been deleted (Kif7 -/-) 3 to investigate the consequence of KIF7 loss of function on the cortex development. Kif7 -/- mice have been previoulsy characterized as dying at birth with severe malformations, including exencephaly in a third of the mutants, skeletal abnormalities (digits and ribs), neural tube patterning defects, microphtalmia, lack of olfactory bulbs and CC agenesis 3,4,15. In the present study, the KIF7 loss of function was analyzed in a mouse strain with a mixed genetic background in which exencephaly at E14.5 was observed at the same frequency in Kif7 -/- and control embryos. As previously reported, E14.5 Kif7 -/- embryos displayed microphtalmia (Fig. 1A, black arrow) and polydactily (not illustrated). Interestingly, skin laxity previously reported in ciliopathic patients 50,51 was observed in all mutants (Fig. 1A, white arrow). Comparison of brains from Kif7 -/- embryos and control littermates at E14.5 revealed that Kif7 -/- could be distinguished by the absence of olfactory bulbs (Fig. 1B, white arrows), a thinning of the dorsal cortex allowing the lateral ventricles to be seen through (Fig. 1B, black arrow and Fig. 1C1) and a prominent diencephalon (Fig. 1C1).

Kif7 deletion alters cortical anatomy at E14.5 but preserves the dorso-ventral patterning of the telencephalon.

(A) Kif7 -/- embryos are microphtalmic (black arrow) and exhibit skin laxity (white arrow). (B) External examination of the brain reveals the thinning of the dorsal telencephalon (black arrow, left panel) and the lack of olfactory bulbs (white arrows, right panel). (C) DAPI staining of rostro-caudal series of coronal sections (C1) illustrates the anatomical defects of Kif7 -/- embryonic brains that are quantified in C2-C5. The ventricles of Kif7 -/- embryos are strongly enlarged (C2), their cortical thickness strongly decreased (C3), resulting in minimal brain width (C4) and height (C5) changes. Statistical significance was tested by Two-way ANOVA or mixed model (GraphPad 8.1.0). In C2 (WT, n=4-6, Kif7 -/-; n=5-6 depending on the rostro-caudal level), C3 (WT, n=3; Kif7 -/-, n= 4), mix model reveals a genotype effect (p<0.0001). No genotype effect was observed in C4 (WT, n=5-6; n=5-6 for Kif7 -/- depending on the rostro-caudal level), C5 (WT, n=5-6; Kif7 -/-, n=4-6 depending on the rostro-caudal level), but a significant effect of the rostro-caudal level on brain width (C4, P=0.0441) and height (C5, P=0.092). D. The pallium-subpallium boundary (PSB) identified by the limit of expression of ventral (GSH2, left panels) and dorsal (TBR2, right panels) telencephalic markers remains well-defined in Kif7 -/- embryos. Graphs in C2-C5 represent the means and S.E.M. PSB, pallium-subpallium boundary; CX, cortex; LGE, lateral ganglionic eminence; MGE, median ganglionic eminence; CGE, lateral ganglionic eminence; Th, thalamus. Scale bars, 250 µm.

Kif7 invalidation alters the development of the cortex at E14.5

Analyses performed on rostro-caudal series of frontal sections (Fig. 1C1) confirmed that the brain of E14.5 Kif7 -/- embryos exhibited enlarged lateral ventricles (Fig. 1C2) and thinned dorsal cortex (Fig. 1C3) resulting however in unchanged total width and height of the telencephalon (Fig. 1C4,5). KIF7 depletion in E10.5-E11.5 mouse embryos was reported to alter the dorso-ventral patterning of the medial forebrain 15, in line with the decreased expression of the transcriptional repressor GLI3R 3,4. GLI3 loss of function 52,53 has been associated with an abnormal dorso-ventral patterning of the telencephalon. We confirmed here the decreased expression of GLI3R in the cortex of Kif7 -/- embryos (Fig. S1). However, the frontier of expression of GSH2 (Fig. 1D1) and TBR2 (Fig. 1D2), two transcription factors expressed in the ventral and dorsal telencephalon respectively, was preserved in the pallium-subpallium boundary (PSB) at E14.5 in Kif7 -/- embryos, suggesting a rather normal dorso-ventral patterning of the telencephalic vesicles in their lateral side.

The thinning of the dorsal cortex of E14.5 Kif7 -/- embryos convinced us to examine the structural organization of the cortex of Kif7 -/- embryos. At E14.5, the mouse cortex can be described as a stack of three specialized domains: i) the upper/superficial cortical layers containing post-mitotic cells generated in the proliferative zones of the cortex, including TBR1(+) post-mitotic neurons located in the cortical plate (CP); ii) the deep proliferative layers, ventricular and subventricular zones (VZ and SVZ, respectively), located along the lateral ventricle; iii) the intermediate zone (IZ) located between the CP and the proliferative layers, which hosts the radially and tangentially migrating neurons and growing axons. In control embryos, TBR1(+) cells formed a rather regular, radially organized CP surrounded by two thin continuous layers of MAP2(+) neurons located in the upper marginal zone (MZ) and the bottom subplate (SP) respectively (Fig. 2A, left). In Kif7 -/- embryos, the TBR1(+) CP cells appeared more clustered and the thickness of the upper MAP2(+) layer was extremly irregular. MAP2(+) cells were no longer observed below the CP in the dorsal cortex (Fig. 2A, arrow on right panel). Moreover, focal heterotopia characterized by an abnormal accumulation of TBR2(+) cells were observed in the dorsal or lateral cortex of Kif7 -/- embryos, either at the ventricular or pial surface (Fig. S2). TBR2 labels the intermediate progenitors (IP) of principal cells, which normally form a dense layer in the SVZ of the cortex (Fig. 2B1, left). Beside sporadic heterotopia, the TBR2(+) layer constantly showed an abnormal positioning in the dorsal cortex of Kif7 -/- embryos. TBR2(+) cells indeed reached the brain surface (Fig. 2B1, arrow) where they mixed up with post-mitotic TBR1(+) cells instead of keeping a minimal distance with TBR1 cells as observed in control brains (Fig. 2B2, arrow) and in the lateral cortex of Kif7 -/- brains (Fig. 2B2). Therefore, intermediate progenitors and post-mitotic neurons no longer segregated in the dorsal cortex of Kif7 -/- embryos and no IZ was observed (Fig. 2B2, white arrow).

Histological alterations in the developing cortex of E14.5 Kif7 -/- embryos.

Coronal sections representative of E14.5 wild type (WT) and E14.5 Kif7 -/- embryos imaged on confocal (A) or epifluorescence (B1) microscopes, and on a macroscope (B2). (A) The TBR1(+) staining (red) of the cortical plate is more clustered in Kif7 -/- than in WT embryos, and the MAP2(+) staining (green) of the subplate is absent in the dorsal cortex of Kif7 -/- embryos (white arrow, right column). (B) The TBR2(+) layer (green) of secondary progenitors appears disorganized in the lateral cortex of the Kif7 -/- embryos (white arrowhead in B1). In the dorsal cortex of Kif7 -/- embryos (white arows in B1, B2), the TBR2(+) cells (green) form a disorganized layer that reaches the brain surface (B1) where it intermingles with TBR1(+) cells (B2, red cells). V, ventricle; VZ, ventricular zone; SVZ, subventricular zone; IZ, intermediate zone; MZ, marginal zone; CP, cortical plate. Scale bars: 100 µm (A), 200 µm (B1, B2).

The loss of Kif7 alters the connectivity between the cortex and the thalamus at E14.5

The intermediate zone (IZ) hosts the radially and tangentially migrating neurons and comprises the growing cortical and thalamic projections 54. Corticofugal axons develop from early stages: the trailing process of post-mitotic neurons migrating radially to the CP first elongates tangentially in the IZ and then differentiates as an axon 55,56. We thus examined whether the defaults observed in the CP and the lack of IZ in the dorsal cortex of Kif7 -/- embryos did associate with developmental abnormalities of the corticofugal projection. We labeled corticofugal axons by inserting small crystals of DiI in the dorsal or lateral cortex in E14.5 paraformaldehyde fixed brains. After DiI had diffused along corticofugal axons, we analyzed labeled axons on coronal sections. In both control and Kif7 -/- brains, CP neurons located in the dorsal or lateral cortex extended axons below the CP toward the PSB (Fig. 3A). DiI injections in the dorsal cortex of control and Kif7 -/- brains (Fig. 3A1) labeled much less axons than injections in the lateral cortex (Fig. 3A2,A3) according to the latero-medial gradient of cortical development. Very remarkably, the cortical bundles labeled from similar cortical regions were always much smaller in Kif7 -/- than in wild type brains (Fig. 3A, compare right and left columns). Whereas dorsal injections labeled large bundles crossing the PSB in control brains, only a few axons reached the PSB in Kif7 -/- brains (Fig. 3A1), showing that corticofugal axons growth was severely hampered in the dorsal cortex of Kif7 -/- embryos. Corticofugal axons labeled from lateral injection sites in control brains crossed the PSB, extended within the embryonic striatum and continued medially in the internal capsule (IC) where they met thalamic axons retrogradely labeled from the lateral cortex (Fig. 3A3, left panel). In Kif7 -/- embryos, lateral injections labeled axons that either spread in the striatum (Fig. 3A2, right panel) or extended ventrally to the cerebral peduncle (Fig. 3A3, right panel, white arrow), a trajectory never observed in control brains. Moreover, no thalamic axons were retrogradely labeled from the lateral cortex. To identify the trajectory of thalamic axons in Kif7 -/- embryos, we immunostained coronal sections of E14.5 brains with antibodies against the Netrin G1a (NG1a), an early marker of thalamocortical axons (TCA, Fig. 3B) 57. At E14.5, in both control and Kif7 -/- brains, NG1a antibodies labeled a large population of dorsal thalamic neurons (Fig. 3B). In control brains, thalamic axons made a right angle turn to join the IC. Then they extended to the PSB and some axons entered the lateral cortex (Fig. 3B, left panel). In Kif7 -/- embryos, most thalamic axons stopped their course at the exit of the diencephalon (Fig. 3B, white arrow in right panel) while some of them formed a short and thick bundle oriented ventrally to the surface of the basal telencephalon (Fig. 3B, white arrow in right panel). No evident structural defect could be observed at the telo-diencephalic junction in Kif7 -/- brain (Fig. S3A). Given the complexity of the trajectory of NG1a thalamic axons, we immunostained thalamic axons in whole brains and imaged them after transparisation using a light-sheet microscope. Reconstruction and three-dimension analysis of labeled projections confirmed that thalamic axons made a sharp turn to reach the IC and then navigated straight to the PBS in control brains (Fig. 3C, Fig. S3B and Movie S1). In contrast, most thalamic axons stopped their course after leaving the diencephalon in Kif7-/- embryos, and formed two large bundles, one oriented to the IC (Fig. 3C, white star and Movie S2) and another one oriented caudally (Fig. 3C, white arrow, Fig. S3B and Movie S2). This last projection was never observed in control brains (Movie S1). The present analysis thus identified impaired trajectories of corticofugal and thalamo-cortical projections at E14.5 in the basal forebrain of Kif7 -/- animals, suggesting impaired axonal guidance mechanisms in the basal forebrain.

Kif7 deletion disrupts of the connectivity between the cortex and the thalamus at E14.5.

(A) Panels illustrate the corticofugal projection labeled by DiI crystal (red dots) positioned in the dorsal (A1) and lateral (A2,A3) cortex of wild type (WT, left) and Kif7 -/- (right) embryos on vibratome sections performed 30 days after DiI placement. In Kif7 -/- embryos, less axons project from the dorsal (A1) and lateral (A2) cortex to the subpallium than in WT. Compare enlarged views of the projections below A1 and A2. Cortical injections in Kif7-/- embryos do not label thalamic axons (compare left and right panels in A3) but label a ventral projection (A3, white arrow). (B) Rostro-caudal series of coronal sections immunostained with anti-Netrin G1a (NG1a) antibodies compare the trajectory of thalamo-cortical axons (TCA) in a WT (left) and in a Kif7 -/- (right) embryo. Thalamo-cortical axons reach the pallium-subpallium boundary (PSB) of the WT embryo whereas they are lost in the ventral forebrain of the mutant (right panel, white arrow). C. Representative three-dimensional reconstruction of WT and Kif7 -/- brains immunostained as a whole with NG1a (red) and TBR1 (green) antibodies before transparization and imaging with a light sheet microscope. Compared to TCA in the WT that form a bundle extending in the internal capsule (IC), TCA in the Kif7 -/- brain organize two short bundles, one directed to the IC that stops soon after entering the IC, and the other one oriented caudally and ventrally. PSB, pallium-subpallium boundary; CX, cortex; LGE, lateral ganglionic eminence; MGE, median ganglionic eminence; Th, thalamus; Hy, hypothalamus. IC, internal capsule. Scale bars: 250 µm.

Kif7 invalidation alters the cortical distribution of cIN at E14.5

Because cIN are born in the basal forebrain and likely depend on contact/functional interactions with pioneer corticofugal projections for the first stages of their migration 58,59, we also examined the distribution of cIN in the developing cortex of Kif7 -/- animals. cIN are generated outside of the cortex, in the medial and caudal ganglonic eminences (MGE, CGE) and preoptic area (POA) of the ventral forebrain. They enter the lateral cortex at E12.5 - E13.5, depending on mouse strains, and colonize the whole cortex by organizing two main tangential migratory streams, a superficial one in the MZ, and a deep and large one in the lower IZ/SVZ 60,61. Intermediate cortical progenitors of the SVZ and meninges strongly express CxCl12 mRNA 62, coding for a chemokine essential for cIN to recognize their tangential pathways of migration in the developing cortex. CXCL12 promotes the tangential progression of cIN in the developing cortex and prevents the premature occurrence from switch to tangential to radial mode of migration that is required for cIN to colonize the CP 62–65. We analyzed the expression of CxCl12 mRNA in the cortex of mutant embryos using in situ hybridization experiments at E14.5, a stage when TBR2(+) SVZ cells show defective positioning in the dorsal cortex of Kif7 -/- embryos. In control brains, CxCl12 mRNA showed a latero-medial gradient of expression in the SVZ, which ended at the frontier with the hippocampus. In contrast, CxCl12 mRNA was no longer expressed in the dorsal cortex of Kif7 -/- embryos (Fig. 4A, black arrow), in the region where TBR2(+) cells starts to position at the brain surface (Fig. 4B1). We thereafter analyzed the cortical distribution of Tomato(+) MGE-derived cIN at E14.5 in control Nkx2.1-Cre;Rosa26-TdTomato transgenic embryos and in Kif7 -/-, Nkx2.1-Cre;Rosa26-TdTomato transgenic mutant embryos (Fig. 4B and see methods). The latero-medial extent of both the superficial and deep tangential migratory streams was significantly shortened in Kif7 -/- brains as compared to control brains (Fig. 4B1, white arrows). Very remarkably, the cIN deep tangential stream stopped in the region where the layer of TBR2(+) cells started its abnormal switch to the surface of the cortex (Fig. 4B1, white arrows) and did not colonize the dorsal cortex (Fig. 4B2). In addition, cIN presented an altered radial distribution in the cortex of Kif7-/- embryos (Fig. 4C). The superficial migratory stream was thinner and denser in the MZ of Kif7 -/- embryos as compared to control (Fig. 4C1, left panel in C2) and the distance between the superficial and deep tangential migratory streams was significantly diminished in the lateral cortex of Kif7 -/- mutants (Fig. 4C2). This decrease is in agreement with the reduced corticofugal projections identified in Kif7 -/- mutants, as these axons navigate tangentially in the IZ. The reduced latero-dorsal distribution of cIN and their abnormal distribution in the cortical thickness reflected structural and molecular abnormalities identified in the cortex of Kif7 -/- embryos. To verify if migratory defaults proper to Kif7 -/- cIN could moreover contribute to their abnormal distribution in the cortex of Kif7 -/- embryos, we analyzed the migration of cIN in living cortical slices.

Altered Cxcl12 transcript expression and abnormal cortical distribution of cIN at E14.5 in Kif7 -/- brains.

(A) Panels compare the distribution of Cxcl12 mRNA in wild type (WT, left panel) and Kif7 -/- (right panel) forebrain coronal sections at E14.5. The WT section shows Cxcl12 transcript enrichment in a deep cortical layer already identified as the SVZ. In the Kif7 -/- cortical section, the expression of Cxcl12 transcripts is reduced to the lateral part of the SVZ. (B) The cortical distribution of cIN is visualized in WT and Kif7 -/- mouse embryos using crosses with the Nkx2.1-Cre/R26R-Tomato strain that expresses the fluorescent marker Tomato in MGE-derived cIN. Panels in B1 compare the distribution of Tomato (+) cIN in WT and Kif7 -/- cortical sections prepared at the same rostro-caudal level and in which SVZ is immunostained with TBR2 antibodies (green). Pictures show that the deep migratory stream of cIN terminates in Kif7 -/- brains in the cortical region where the TBR2(+) layer reaches the cortical surface. Quantitative analysis of the mean length of the deep and superficial migratory streams measured from the entry in the pallium to the last detected cIN in the cortex is illustrated in graph B2. Statistical significance is assessed using Two way ANOVA [layers (WT, n=4; Kif7 -/-, n=3; ***, P=0.001) and genotype (*, P=0.0157)]. (C) Representative pictures (C1) of the deep and superfical tangential migratory streams of cIN in the lateral cortex of WT and Kif7 -/- embryos. Pictures illustrate the decreased thickness of the superficial stream, and the reduced distance between the deep-superficial streams in Kif7 -/- embryos. Graph in C2 (WT, n=4; Kif7 -/-, n=4) compares the distribution of the fluorescence intensity along a ventricle/MZ axis (see grey rectangles in C1) using the plot profile function of FIJI. Curves show no change in the distance between the ventricular wall and the deep cIN and a reduction of the distance between the two streams in the Kif7 -/- cortical sections as quantified on the graph [WT, n=4; Kif7 -/-, n=4; Two way ANOVA reveals a significant interaction between genotype and layer (P=0.0233) and multiple comparisons, a statistical difference between genotype only for the distance between de cIN streams (**, P=0.0051)]. Scale bars: 200 µm.

Kif7 invalidation and SHH pathway activity both affect the migratory behavior of cIN in cortical slices

The forebrain of E14.5 Nkx2.1-Cre;Rosa26-TdTomato transgenic embryos exhibiting or not a Kif7 deletion (control and Kif7 -/-, respectively) were sliced coronally to image the cortical migration of cIN by time-lapse video-microscopy (Fig. 5). In control slices (Fig. 5A1, Supplementary Movie S3), fluorescent cIN migrated tangentially from the PSB to the dorsal cortex in two main pathways, a superficial one in the MZ and a deep one in the lower IZ/SVZ. All along these pathways, some cIN operated a tangential to oblique/radial migration switch to colonize the CP. Since high cell density in the MZ prevented a dynamic monitoring of tangentially migrating cells, cells migrating in the MZ were not analyzed in this study.A large proportion of cells imaged during the recording session (59.3 %), either maintained tangentially oriented trajectories in the deep stream (28.7 %, red trajectories in Fig. 5A1, left column of Fig. 5A2 and Fig. 5A3) or reoriented to the CP along oblique (and less frequently radial) trajectories (30.6 % green trajectories in Fig. 5A1, left column in Fig. 5A2). A majority of cells reached the CP surface and remained there. A smaller proportion of cIN moved from the deep stream to the opposite - ventricular-side of the slice (11.7%, blue trajectories in Fig. 5A1, left column in Fig. 5A2 and Fig. 5A3). In the VZ, deep stream, and CP, a minimal proportion of cIN (3.6 %) moved very short distances or remained immobile (dark colors in the left column of Fig. 5A2). A significant proportion of cIN (15.6 %) migrated radially, traveling all over the cortex thickness and frequently inverting their polarity in the MZ or at the ventricular surface (pink trajectories in Fig. 5A1, left column in Fig. 5A2 and Fig. 5A3). Finally, 8 % (orange) of the recorded cIN moved backward to the ventral brain and 1.7 % of them (grey) did not show any specific directionality or layer specificity and were classified as chaotic (Fig. 5A1, left column in Fig. 5A2 and Fig. 5A3). A large majority of cIN presented a characteristic saltatory behavior, alternating stops and fast moves 66. Analyses of resting and moving phases of the largest cIN population recorded in the deep stream and/or moving to the CP are illustrated in Fig. 5B1-3 (left column of histograms). The trajectories followed by cIN in the cortical slices of Kif7 -/- embryos dramatically differed from those in control slices (Fig. 5A2,A3 and Supplementary Movie S4 compared to S1). The number of cIN migrating tangentially in the deep migratory stream and moving to the CP from the deep stream dropped drastically (17.8 % as compared to 59.3 % in control slices), instead, they were immobile (35.7 % as compared to 3.6 % in control slices) especially in the SVZ/lower IZ. Almost the same proportion of cIN moved to the ventricle and to the CP (10.2 % and 13.9 %, respectively, as compared to 11.7 % and 30.6 % in control slices) suggesting that the radial asymmetry of cortical slice was lost for Kif7 -/- cIN. The proportion of cIN migrating radially across cortical layers remained similar as in control slices (17.2 % versus 15.6 % in control) but the proportion of cIN with chaotic trajectories strongly increased (7.8 % versus 1.7 % in control slices, Fig. 5A2, second column, Fig. 5A3). Another major change observed in the Kif7 -/- cortical slices was a significant decrease of the migration speed (excluding immobile cIN) (Fig. 5B1) and an abnormal saltatory behavior, with less frequent but longer resting phases than in control slices (Fig. 5B2,B3). Of note, trajectories were analyzed at E14.5 in the lateral and latero-dorsal cortex where CxCl12 mRNA was observed in both control and Kif7 -/- embryos (Fig. 4A). Because CXCL12 is a potent chemoattractant for migrating cIN, the very large proportion of immobile cIN recorded in the CXCL12-expressing deep stream of Kif7 -/- slices was suggestive of intrinsic migratory defaults of Kif7 -/- cIN. Kif7 ablation has been shown to induce either transcriptional SHH signaling activation by inhibiting the processing of GLI3 in GLI3-R repressor, or transcriptional SHH signaling inhibition by inhibiting the positive impact of KIF7 on the GLI1/2 dependent SHH signaling 3. We thus examined whether the acute application of drugs able to either promote or inhibit the SHH signaling could induce in wild type cIN migratory defaults resembling those of cIN in Kif7 -/- cortical slices. SHH application on control slices minimaly affected the trajectories of cIN (Fig. 5A2, compare left and right columns, Fig. 5A3), except an increased proportion of cIN migrating radially (29 % compared to 15.6 % in control slices) at the expense of cIN migrating from the deep stream to the CP (20.8 % instead of 30.6 % in control slices). In addition, cIN moving radially across the cortical layers no longer stopped in the CP and frequently moved backward to the VZ (Fig. 5A2,A3, Supplementary Movie S5). The frequency of chaotic trajectories (grey, Fig. 5A2) did not increase, and the migration speed of cIN was slightly decreased compared to control slices (Fig. 5B1). Main dynamic alteration of SHH treated cIN was the increased duration of resting phases (Fig. 5B2,B3). On the contrary, application of the SHH pathway inhibitor cyclopamine strongly altered cIN trajectories (Fig. 5A2,A3). For example, the proportion of immobile cIN and of cIN moving short distance reached 33.7 % as compared to 3.6 % in control slices (Fig. 5A2). Most immobile cells in cyclopamine treated slices were located in the deep stream, and only few cIN switched from the deep stream to the CP at the benefit of cIN migrating backward to the ventral brain. Therefore, inhibiting the SHH pathway mainly affected the ability of cIN to move tangentially in the deep stream and to switch to the CP (Fig. 5A2,A3, Supplementary Movie S6),. Globally, cIN trajectories in cyclopamine treated slices recalled the trajectories of cIN in Kif7 -/- cIN (Fig. 5A2, compare columns 2 and 3), suggesting that Kif7 ablation resembled the cyclopamine-mediated SHH pathway inhibition. Nevertheless, whereas the migration speed of Kif7 -/- cIN was significantly decreased in the deep tangential pathway and not in CP, cyclopamine application decreased the migration speed only in the CP (Fig. 5B1). The strong influence of cyclopamine on the cortical migration of cIN led us to determine the distribution of the endogenous activator SHH in the developing cortex.

Dynamic behavior of migrating cIN in organotypic cortical slices: comparison between Kif7 -/- slices and control slices with an acute treatment to either activate or block the SHH pathway.

(A) Tomato(+) cIN migrating in living cortical slices were tracked manually using the MTrackJ plugin and their trajectories color-coded as shown in legend to characterize their preferred direction (tangential, oblique, radial, immobile) and cortical layer localization (VZ-SVZ, IZ, CP). The picture in A1 illustrates the z-projection of trajectories reconstructed in a control slice, superimposed to the last picture of the movie. Cortical interneurons migrating tangentially in the superficial migratory stream (MZ) could not be tracked because of high density. A2. Graphs compare the percentage of each kind of trajectory recorded in the lateral cortex of control slices (control, see also Suppl. Movie S3), Kif7 -/- slices (Kif7 -/-, see also Suppl. Movie S4), control slices treated acutely with either murine SHH (SHH, see also Suppl. Movie S5) or cyclopamine (Cyclo, see also Suppl. Movie S6). Significance of the differences between the four distributions were assessed by a Chi-square test, X2 (24, n=1224), P=0.0004998, ***). All experimental conditions differed from the control (Fisher test, p =0.0004998, ***). Slices from three WT animals in control condition or treated with drugs and from three Kif7 -/- animals were analyzed; the number of analyzed cells is indicated above bars. A3. Schemes summarize the main results observed in each experimental condition. Trajectories are represented with the same color code as in A1, and line thickness is proportional to the percentage of cells exhibiting each type of trajectory. Immobile cells are figured by a coil. (B) Box and whisker plots indicate the mean speed (B1) and the frequency (B2) and duration (B3) of resting phases for cIN migrating tangentially in the deep stream (red box and whisker plots, left) or to the cortical plate (green box and whisker plots, right). Statistical significance assessed by Krustkal-Wallis tests in each cluster. ****, P<0.0001; ***, B2 left P=0.0005; **, B1 right P=0.0088, B2 right, P=0.0037, B3 left P=0.0034; *, B1 left P=0.0233, B2 right P=0.0470, B3 left P=0.0134, B3 right P=0.0394. Number of analyzed cells is indicated on plots. VZ, ventricular zone; SVZ, subventricular zone; IZ, intermediate zone; CP, cortical plate; MZ, marginal zone. Scale bar: 300 µm.

Distribution of Shh mRNA and SHH protein in the E14.5 cortex

Shh mRNA is strongly expressed by the mantle zone of the ganglionic eminences (MGE, CGE) and POA during the time cIN differenciate 24,56 but its level of expression in cIN after they reach the cortex is controversial 29,31,67. We thus reexamined Shh mRNA expression in forebrain and SHH protein distribution in cortical layers. In situ hybridization (ISH) experiments performed at E13 confirmed the strong expression of Shh mRNA in the ventral forebrain, especially in the SVZ and mantle zone of the MGE (Fig. 6A1,6B1) in agreement with the SHH-dependent Gli1 mRNA expression in a VZ subdomain of the MGE 68. In contrast, Shh transcripts were bearly detectable in the cortex. Using RNAscope, whose sensitivity and spatial resolution is higher compared to standard ISH, we re-examined the expression pattern of Shh mRNA at E14.5. In the ventral forebrain, RNAscope Shh transcript detection showed similar expression pattern as ISH (Fig. 6A2,B2). In addition, the expression domains of Shh mRNA and Lhx-6 mRNA, a cIN marker, overlapped in the mantle zone of the MGE (Fig. 6A2). In the cortex, some cells oriented either radially in the VZ/SVZ or tangentially in the tangential migration pathways expressed Shh mRNA at a very low level (one or two dots versus large cluster in the MGE) (Fig. 6A2). In the tangential migration streams where a large number of cIN expressed Lhx-6 mRNA (Fig. 6C1,C2), cIN that co-expressed Lhx-6 and Shh mRNA represented near 5 % of cIN expressing Lhx-6 transcript (Fig. 6C3).

Expression of Shh transcripts (A-C) and distribution of SHH protein (D) in the developing forebrain.

(A,B) Distribution in median (A) and caudal (B) coronal sections of Shh mRNA detected by in situ hybridization with an antisens Shh probe at E13.5 (A1,B1) and by RNAscope at E14.5 (A2-3,B2). Shh transcripts are strongly expressed in the medial ventral forebrain (SVZ and mantle zone of the MGE and septum, A1, A2), in the mantle zone of the CGE (B1,B2), in the zona limitans intrathalamica (ZLI in B1) and in the ventral midline of the 3rd ventricle (V3 in B1). RNAscope further confirmed the strong expression of Shh mRNA (A2, green) in MGE and septum regions that strongly express Lhx-6 mRNA (A2, red). Confocal observations in the SVZ and mantle zone of the MGE showed that Shh mRNA (green) is co-expressed with the Lhx6 mRNA (red) in a significant number of cIN (yellow cells in A3). (C) Confocal analyses at higher magnification of the double detection by RNAscope of Shh and Lhx-6 mRNA. Cells were identified on stacked images (Δz=1 µm) using Nomarski optic. In the lateral cortex close to the PSB (C1, z projection of 10 confocal planes) and in the dorsal cortex (C2, z projection of 10 confocal planes), a very small proportion of cells expressing Lhx-6 mRNA also express Shh mRNA (white arrows in C1,C2). Counting in the deep stream (SVZ-IZ) and in the MZ is shown in graph C3 (9-17 fields in three sections). A few progenitors in the cortical VZ express Shh mRNA at very low level (arrowhead, C1). (D) E14.5 forebrain coronal sections immunostained with antibodies directed against the N-ter domain of the activated SHH protein are imaged by epifluorescence (D1) and confocal (D2) microscopy. At low magnification (D1), SHH-Nter is slightly enriched in the ventral midline (black arrow) and at the ventricular angle (black arrowhead) near the PSB. Confocal analysis of SHH-Nter immunostaining (D2) shows immuno-detection in blood vessels and the presence of numerous bright dots all over the cortical neuropile. In the cortical neuropile, small bright dots align radially in the VZ (enlarged in D3), tangentially in the SVZ-IZ, and radially in the CP. On the ventricular side of the PSB and of the lateral-most part of the LGE (D3), larger SHH-Nter(+) brigth elements are aligned radially. Confocal images are a merged stacks of 10 images distant of 0.2 µm. CX, cortex; LGE, MGE and CGE, lateral, medial and caudal ganglionic eminence; V3, third ventricle; ZLI, zona intra-thalamica; Th, thalamus; Hyp, hypothalamus; PSB, pallium-subpallium boundary; VZ, ventricular zone; SVZ, subventricular zone; IZ, intermediate zone; CP, cortical plate; MZ, marginal zone. Scale bars: 500 µm (A, B, D1), 20 µm (C), 250 µm (D1), 50 µm (D2).

Antibodies that recognize the N-ter part of SHH [SHH-Nter, e.g. “activated” SHH 69] strongly labeled brain regions known to strongly express SHH 67,70,71, such as zona limitans intrathalamica and the lateral walls of the third ventricle at E12.5 (Fig. S4), and the choroid plexus at E14.5 (Fig. 6D1), validating the SHH-Nter antibodies used here. At E14.5, SHH-Nter antibodies immunostained bright tiny dots aligned radially in the VZ/SVZ, and tangentially in the IZ. In the CP, dots were again radially aligned and some CP cells presented a rather diffuse cytoplasmic staining, which was not observed in the bottom layers (Fig. 6D2). We observed no particular SHH enrichment in the lower IZ/SVZ where cIN migrate, suggesting that their contribution to the cortical pattern of SHH expression was minimal. Moreover, large dots were observed in the subpallium at the ventricular angle (Fig. 6D4). Together, our ISH and immunostaining experiments showed that the cortical SHH protein has probably an extrinsic origin as previously described at the early stage of cerebellum development (Huang et al, 2010).

Ontogeny of Kif7 mutant brain developmental defects

After E14.5, the cortex of Kif7 -/- embryos continued to develop. Analyses performed at E16.5 showed that cIN continued to migrate tangentially and reached the dorsal cortex in Kif7 -/- embryos (Fig. 7A). Nevertheless, the tangential migration of cIN to the dorso-medial cortex remained delayed in Kif7 -/- embryos with regard to the tangential migration of cIN in control embryos. Kif7 -/- cIN failed to reach the hippocampus identified by the presence of TBR2(+) cells, as observed in control brains (compare left and right panels in Fig. 7A). At E16.5, NG1a(+) thalamic axons colonized the whole medial and rostral cortex, in both control and Kif7 -/- embryos (Fig. 7B). Nevertheless, thalamic axons were less dense in Kif7 -/- than in control cortex and exhibited abnormal trajectories (Fig. 7C) toward the cortical plate. Analyses performed in a surviving P0 animal (Fig. S5) revealed an abnormal distribution of cIN in the cortical plate, with a decreased density in the bottom layers, which were moreover thinner than in control brains, as illustrated by CTIP2(+) staining (Fig. S5A). The cortical distribution of thalamic axons was extremely perturbed in Kif7 -/- brains: a dense bundle of NG1a(+) fibers ran below the cortical plate to the dorsal cortex and failed to invade the CP (Fig. S5B).

KIF7 deletion affects the cortex development at E16.5.

(A) Coronal sections at E16.5 of transgenic Nkx2.1-Cre/R26R-Tomato control (left panels) and Kif7 -/- (right panels) brains show that the hippocampus does not form properly in Kif7 -/- animals and the cerebral cortex remains thinner. The migration of Tomato(+) cIN to the dorso-medial cortex had progressed between E14.5 and E16.5, but remains delayed with regard to E16.5 control brains. (B) Coronal view of whole-mount imaging of control and Kif7 -/- brains immunostained with NG1a (red) and TBR1 (green) antibodies shows that thalamo-cortical axons extended to the cortex in both control (upper panel) and Kif7 -/- (bottom panel) brains. An abnormal projection of thalamic axons to the ventral telencephalon is still present in Kif7 -/- animals is (white arrow in B). (C) Immunostaining of thalamo-cortical axons on coronal brain sections reveals hampered colonization of the deep cortical layers above TBR2(+) layer (green) by thalamic axons (NG1a in white) in the Kif7 -/- embryos. Scale bars: 100 µm.

Discussion

Our aim in the present paper was to better understand the developmental origin of functional abnormalities (e.g. epilepsy, cognitive deficits, …) in cortical circuits of human patients with KIF7 mutations by characterizing the developmental abnormalities of the cerebral cortex in a Kif7 -/- mouse model. We show that Kif7 ablation affects the development of the two populations of cortical neurons, cortical plate neurons and GABAergic interneurons, in specific and distinct ways, accordingly with their dorsal and ventral origin in the telencephalon. The alterations of the CP development (lack of most subplate cells in the dorsal cortex, and delayed corticofugal projection) are remiscient of defaults previously described in GLI3 null mutants that lack most subplate neurons and pioneer cortical projections 72. And the abnormal distribution of Kif7 -/- cIN in the developing cortex indeed reflects the structural and functional alterations of cortical layers on which cIN migrate (e.g. abnormal CXCL12 expression in the SVZ). However, we identify moreover migration defaults of Kif7 -/- cIN that recall those of wild type cIN treated with the SHH pathway inhibitor cyclopamine, suggesting that the SHH pathway activity is inhibited in Kif7 -/- cIN. Kif7 ablation thus alters specifically and differently the SHH pathway activity in CP and cIN subpopulations, in agreement with their dorsal or ventral origin.

Kif7 knock-out mice as a model to study the effect of mutation in the KIF7 gene in human

The deletion of Kif7 gene in mice leads to lethality the first day of life, but up to birth, the brain develops and a previous study demonstrated that Kif7 mutant mice was a good model to dissect the mechanism that led to corpus callosum agenesis in patients 15. All patients carrying mutation in the KIF7 gene on both alleles have developmental delay/intellectual disability (ID) and some have seizures. Investigations on cerebral malformation by MRI have shown that some of them have alterations in the cerebral cortex including cortical atrophy 9,10,18 and pachygyria 17. Detecting defects in the cerebral cortex is more demanding in MRI quality images that corpus callosum agenesis or molar tooth sign, we can speculate that the number of patients with cerebral cortex defects was underscored. Malformations of cortical development was recognized as causes of ID 73 and epilepsy 74 and the cellular defects underlying these pathologies could imply defects in cell proliferation, neuronal migration and differenciation that lead to improper excitatory/inhibitory balance, connectivity into the cortex or of the cortex with other brain structures. The mechanism of action of KIF7 to transduce SHH ciliary dependent signaling has been debated. It has been accepted that SHH activates KIF7 binding to microtubules and its accumulation to the tip of the cilium, making the kinesin motor domain the most essential domain for SHH transduction. A more recent study proposed that KIF7 is an immobile kinesin translocated to the cilium tip after SHH binding by a complex KIF3A/Kif3B/KAP 75 suggesting that domains in the KIF7 gene essential to its function are more complex. This could explain why any mutation, even point mutation in the Coiled-coil domain or in C-terminal part of the KIF7 protein induces clinical feature as severe as when mutation led to a highly truncated protein. Another explanation is that any mutation causes unproper folding of the protein inducing its degradation by the proteasome.

KIF7 loss of function alters the cerebral cortex organization and the connectivity between the cortex and thalamus

In the mouse embryo, GLI3 plays a major role in patterning the dorsal telencephalon and in regulating the proliferation and differenciation of cortical neurons 40–42. The Kif7 knock-down that likely decreases GLI3R level leads to a decrease of apical radial progenitors and basal IP divisions, and of neuronal differenciation 44. In fibroblasts isolated from lung of patients carrying KIF7 mutation and in E11.5 Kif7 -/- whole embryo, the processing of GLI3-full length in GLI3-R is decreased and GLI-A increased leading to over-SHH signaling 3,49. Our study demonstrates that the level of GLI3-R was decreased in the developing cortex of the Kif7 -/- embryos, whose layering of principal neurons was altered and exhibited sporadic heterotopia and delayed connectivity between the cortex and the thalamus. Heterotopia have been observed in one patient 17 and we found cortical heterotopia in some Kif7 -/- brains. Cortical heterotopia is not a landmark of ciliopathy but were detected in a Joubert syndrom patient carrying a mutation in the CELSR2 gene coding for a planar cell polarity protein 76 essential for neural progenitor cell fate decision, neural migration, axon guidance and neural maturation 77,78. In the patient as in the mice model, the heterotopia are focal and may have been missed in other patients. They nevertheless emphasize the mistrafficking of principal cells in KIF7 depleted brain. Although disorganized thalamo-cortical connectivity had never been reported in patients with KIF7 mutation or in patients with ciliopathy, thalamocortical connectivity is essential for cognitive performances in young infants 79 and its alteration leads to neurodevelopemental delay and later cognitive deficits 80. Moreover, a link beween ciliary function, cortex development and connectivity with the thalamus had been previously identified in mutant mice lacking planar cell polarity proteins CELSR2 and FZD3. These mice display ciliary defects and an abnormal development of tracts between the thalamus and the cortex leading to an hypotrophy of the intermediate zone where thalamo-cortical and corticofugal axons navigate81,82.

Extra-cortical origin of SHH

Beside its major function in forebrain patterning 20,40, SHH moreover regulates the embryonic and perinatal cerebral cortex development 27,28. The origin of the morphogen in this brain structure remains obscure. In agreement with a recent set of data from Moreau et al. 83, we showed that cIN born in the MGE or POA no longer express SHH after they reach the developing cortex. In addition, the level of local Shh mRNA synthesis was extremely low in the cortex at E14.5. In contrast, SHH protein was detected in immunoposive dots distributed throughout the cortex. Dots aligned radially in the VZ and CP, and tangentially in the IZ. This cortical distribution is compatible with an extrinsic origin of SHH and transventricular delivery as described in the cerebellar ventricular zone 84. SHH is indeed present in the cerebrospinal fluid and likely secreted by the plexus choroid, since Shh mRNA is expressed in the choroid plexus of mice from E12 to E14.5 85. Accordingly, we showed here that the choroid plexus of the lateral ventricles was strongly immunopositive for SHH at E14.5. Interestingly, the CDO/BOC co-receptor - that favors SHH transport - was also detected along radially oriented structures by immunostaining in the mouse embryonic cortex (not illustrated). SHH could thus be transported away from the ventricular zone and/or meninges in vesicular structures able to diffuse accross the cortical thickness. Accordingly, it has been shown that SHH transport needs ESCRT-III member protein CHMP1A 13 and SHH related proteins are found in exosomes secreted from Caenorhabditis elegans epiderman cells 86, Drosophila wing imaginal discs 82, chick notochord cells and human cell lines 87. How SHH is transported thoughout the cortex should be further explored. It has already be shown that SHH is present in multivesicular bodies, in neurites and filipodia in the post natal hippocampus and cerebellum, and in vesicles located inside and outside of hippocampal neurons in culture 88.

Kif7 deletion leads to defects in cortical interneuron migration

The defects induced by gene mutations responsible of primary cilium dysfunction, including Kif7 deletion, have been examined in principal cells born in the dorsal telencephalon, which organize the cortical layers 89,90. However, the cIN are born ventrally and migrate to the cortex during the embryonic development. Their primary cilium is functional and can transduce local SHH signals to influence their directionality, especially the tangential to radial reorientation of their trajectories toward the cortical plate 29. In the present study, we showed that acute SHH application on control cIN slightly increased the percentage of radially migrating cIN. In contrast, cyclopamine that inhibited the SHH pathway blocked the migration of a large population of cIN in the deep cortical layersdemonstrating that endogeneous cortical SHH is likely to activate the SHH pathway in migrating cIN to control their trajectories. Since Kif7 -/- cIN showed a migratory behavior that recalled the migratory behavior of cyclopamine treated control cIN, we concluded that SHH signaling was inhibited in Kif7 -/- cIN. By consequence, we propose that Kif7 -/- cIN, which differentiate in a ventral telencephalic domain expressing high SHH level, present reduced Gli-A signaling, e.g. decreased SHH signaling. In addition to this cell autonomous defect that impairs cIN migration, the colonization of the dorsal cortex by Kif7 -/- cIN could be delayed due to the decreased dorsal expression of CXCL12 in Kif7 -/- embryos 62,91.

In conclusion, using the Kif7 -/- murine model, we show that the Kif7 deletion is responsible for a large range of developmental defects in the cerebral cortex, which include: i) alteration of the inhibitory interneurons migration by a combination of intrinsic and extrinsic factors, ii) abnormal formation of cortical layers by the cortical plate neurons and iii) major pathfinding defects in axonal projections between the cortex and the thalamus. Patients carrying mutations in the KIF7 gene are classified as ciliopathic patients and display developmental delay, ID and epilepsy. These clinical features could thus be caused by an abnormal distribution of excitatory and inhibitory cortical neurons leading to impaired excitatory/inhibitory balance, and by the development of abnormal connections in the cerebral cortex and with others brains structures. By demonstrating focal heterotopia and abnormal thalamo-cortical connectivity in the murine model, we hope to open a new field of clinical investigations using MRI and tractography to identify such defects in patients carrying mutations in the KIF7 gene.

Material and methods

Mice

E14.5 embryos of the swiss (for in situ hybridization and RNAscope) and C57/Bl6J (for sonic hedgehog immunostaining) strains were obtained from pregnant females purchased at JanvierLabs (France). Male and female wild type (WT) and Kif7 -/- animals (E14.5, E16.5 and P0) expressing TdTomato or not in the MGE-derived cIN (Nkx2.1-Cre;Rosa26-TdTomato) were generated in our animal facility by mating male Kif7 +/-; Nkx2.1-Cre;Rosa26-TdTomato and female Kif7 +/- mice to produce embryos and P0. Thirty four pregnant females were used. Kif7 +/- mice are a generous gift of Pr Chi-Chung Hui (the Provider, SickKids Hospital, Toronto, USA) and Prof Bénédicte Durand (the Transferor, CNRS, Lyon, France). Mice were genotyped as described previously 3. The day of the vaginal plug was noted E0.5. Experiments have been validated and approved by the Ethical committee Charles Darwin (C2EA-05, authorized project 02241.02) and mice were housed and mated according to European guidelines. Both male and female animals were used in this study. Sex is not considered as a biological variable in embryos.

Immunohistochemistry and imaging

Heads of embryos and P0 animals were fixed by immersion in 0.1% picric acid /4% paraformaldehyde (PAF) in phosphate buffer (PB) for 4 hours and then in 4% PAF in PB overnight. After PBS washes, brains were dissected, included in 3% type VII agarose (Sigma, A0701) and sectionned at 70 μm with a vibratome. Immunostaining was performed on free-floating sections as explain previously (Baudoin et al, 2012). Primary antibodies were goat anti SHH-Nter (1:100, R&D system AF464), goat anti Netrin G1a (NG1a) (1:100, R&D system AF1166), rabbit anti TBR1 (1:1000, Abcam ab31940), rabbit anti TBR2 (1:1000, Abcam ab23345), rabbit anti PAX6 (1:100, clone poly19013, Covance PRB-278P), rabbit anti GSH2 (1:2000, Millipore ABN162), chicken anti MAP2 (1:500, Novus, NB30213), and rat CTIP2 (1:1000, Abcam ab18465). Primary antibodies were incubated in PGT (PBS/gelatine 2g per L/TritonX-100 0.25%) with 0.1% glycine and 0.1% lysine for anti Netrin G1a antibodies and 3% BSA for anti SHH-Nter antibodies. Primary antibodies were revealed by immunofluorescence with the appropriate Alexa dye (Molecular Probes) or Cy3 (Jackson laboratories) conjugated secondary antibodies diluted in PGT (1:400). Bisbenzimide (1/5000 in PBT, Sigma) was used for nuclear counterstaining. Sections were mounted in mowiol/DABCO (25mg/mL) and imaged on a macroscope (MVX10 olympus) or an epifluorescence microscope (LEICA DM6000) using X40 and X63 immersion objectives, or a confocal microscope (Leica TCS SP5) using a x40 objective. Double immunotraning were performed when possible and images were representative of what was observed in more than 4 animals per genotype except for heterotopia. Macroscop images were treated to remove background around sections using wand tool in Adobe Photoshop 7.0 software. On epifluorescence and confocal microscop images, acquisition parameters and levels of intensity in Adobe Photoshop 7.0 software were similar for WT and Kif7 -/-. Merged images were acquired on the same section except for TBR1 with TBR2 aquired on adjacent sections. The fluorescence intensity along the depth of the cortex was assessed using the plot profile function of ImageJ under a 500 pixels wide line starting in the ventricle to the surface of the brain.

In situ hybridization

Specific antisense RNA probe for Shh, Gli1 and CxCl12 genes (gift of Marie-Catherine Tiveron) were used for in situ hybridization analyses. DIG-probes were synthesized with a labeling kit according to the manufacturer’s instructions (Roche, France). In situ hybridization was performed according to a modification of Tiveron et al 45 (see Supplementary material). Sections were mounted on glass slides, dried, dehydrated in graded ethanol solutions, cleared in xylene and coverslipped with Eukitt.

RNAscope

Embryonic E14.5 heads were cuted in cold Leibovitz medium (Invitrogen) and immersion fixed in cold 4% (w/v) paraformaldehyde (PFA) in 0.12 m phosphate buffer, pH 7.4, overnight. Brains were then cryoprotected in PFA 4% /sucrose 10 %, embedded in gelatin 7.5% /sucrose 10% at 4°C and frozen. Biological samples were kept at -80°C until coronally sectioned at 20 μm with cryostat. RNAscope experiment was performed according manufacturer’s instructions (RNAScope® Multiplex Fluorescent V2 Assay, ACDbio) after a pre-treatment to remove gelatin. Two different probes/channels were used (C1 for Shh, C2 for Lhx-6). Probes were diluted according manufacturer’s instructions (1 Vol of C2 for 50 vol of C1) and dilution of the fluorophores (Opal 520 and 570) was 1:1500. Sections were mounted in mowiol/DABCO (25mg/mL) and imaged on a confocal microscop (Leica TCS SP5) using an immersion X10 or X63 objectives on 10 µm stacks of images (1 image /µm).

DiI experiments

Tiny crystals of DiI (1,1′-dioctadecyl-3,3,3′3′-tetramethylindocarbocyanine perchlorate) were inserted in the dorsal or lateral cortex of WT and Kif7 -/-embryonic brains fixed at E14.5 by immersion in 4% PAF. Injected brains were stored in 4% PAF at room temperature in the dark for several weeks to allow DiI diffusion. Before sectioning, brains were included in 3% agar. Coronal sections 60 μm thick were prepared with a vibratome and collected individually in 4% PAF. Sections were imaged with a macroscope (MVX10 Olympus).

Whole brain clearing and imaging

E14 and E16 embryos (WT, n=5; Kif7 -/-, n=6) were collected in cold L15 medium, perfused transcardially with 4% PAF using a binocular microscope. Brains were dissected and postfixed 3 days in 4% PAF at 4°C, and stored in PB at 4°C until clearing. All buffer solutions were supplemented with 0.01% Sodium Azide (Sigma-Aldrich). Whole brain immunostaining and clearing was performed at RT under gentle shaking according to a modified version of the original protocol 46. Briefly, perfused brains were dehydrated in graded methanol solutions (Fischer Scientific) (20, 50, 80% in ddH2O, 2×100%, 1h30 each), bleached overnight in 3% hydrogen peroxide (Sigma-Aldrich) in 100% methanol and rehydrated progressively. After PBS washes, brains were blocked and permeabilized 2 days in PBGT (0.2% gelatin (Merck), 0.5% triton X-100 (Sigma-Aldrich) in PBS). Brains were incubated 3 days at 37°C in primary goat anti Netrin G1a (1:100, R&D system AF1166) and rabbit anti Tbr1 (1:1000, Abcam ab31940) antibodies. After 1 day of PBGT washes, brains were incubated 1 day at 37°C in secondary 647 donkey anti-goat (1:1000, Molecular probes,) and cy3 donkey anti-rabbit (1:1000,Jackson laboratories) antibodies.Immunolabeled brains were washed 1 day in PBS, embedded into 1.5% low-melting agarose (type VII, Sigma) in 1% ultra-pure Tris-acetate-EDTA solution, placed in 5-ml tubes (Eppendorf, 0030119452) and dehydrated 1h in each methanol baths (50%, 80% and 2×100%). Samples were then incubated for 3h in 33% methanol / 66% dichloromethane (DCM, Sigma-Aldrich), washed in 100% DCM (2×30 min) and incubated overnight (without shaking) in dibenzyl ether (DBE). Brains were stored in DBE at RT. Cleared samples were imaged on a light-sheet microscope (LaVision Biotec, x6.3 zoom magnification) equipped with a sCMOS camera (Andor Neo) and Imspector Microscope controller software. Imaris (Bitplane) was used for 3D reconstructions, snapshots and movies.

Brain slices

We dissected brains of E14.5 WT and Kif7 -/- transgenic Nkx2.1-cre, TdTomato embryos in cold L15 medium, embedded in 3% type VII agarose and sectioned coronally with a manual slicer as explained in Baudoin et al, 2012. Forebrain slices 250 µm thick were transferred in Millicell chambers and cultured for 4 hours in F12/DMEM medium supplemented with CFS 10% in a CO2 incubator (Merck Millipore) prior to pharmacological treatment and recording. Slices were incubated in drug for 2 hours before transfer in culture boxes equipped with a bottom glass coverslip for time-lapse imaging.

Pharmacological treatment

Either recombinant Mouse SHH (C25II), N-terminus (464-SH-025, R&D systems) at 0.5µg/mL final or cyclopamine (C4116, Sigma) at 2.5 µM final were added to the culture medium of slices and renewed after 12 hours. Recombinant SHH was reconstituted at 100 µg/mL in sterile PBS with BSA 0.1%. The stock solution of cyclopamine was 10 mM in DMSO and the culture medium of control experiments contained 1/4000 DMSO (vehicle). Control experiments in culture medium with and without vehicle did not differ and were analyzed together.

Videomicroscopy and image processing

Slices were imaged on an inverted microscope (Leica DMI4000) equipped with a spinning disk (Roper Scientific, USA) and a temperature-controlled chamber. Multi-position acquisition was performed with a Coolsnap HQ camera (Roper Scientific, USA) to allow the recording of the whole cortex. Images were acquired with a X20 objective (LX20, Fluotar, Leica, Germany) and a 561 nm laser (MAG Biosystems, Arizona). Z-stacks of 30 μm were acquired 50 µm away from the slice surface, with a step size of 2 μm and a time interval of 2 or 5 minutes for at least 21 hours. Acquisitions were controlled using the Metamorph software (Roper Scientific, USA). Cell trajectories were reconstructed on movies by tracking manually the cell rear with MTrakJ (ImageJ plugin, NIH, USA) and were clustered according to their position in cortical layers and to their orientation (see details in Fig. 5). The migration speed of cells and the frequency and duration of pauses were extracted from tracking data using an excel macro.

Study design

WT group was compared to Kif7 -/- group and to pharmacological treatment groups for brain slice experiments. The experimental unit was a single animal for immunohistochemistry analysis and individual cell for videomicroscopy analysis.

Statistical analysis

All data were obtained from at least three independent experiments and are presented as mean ± SEM (standard error of mean). Statistical analyses were performed with the GraphPad Prism software or R. Statistical significance of the data was evaluated using the unpaired two-tailed t test, the Mann–Whitney test, the Chi2 test or the Two-way ANOVA test followed by a post-hoc test. Data distribution was tested for normality using the D’Agostino and Pearson omnibus normality test. Values of P<0.05 were considered significant. In figures, levels of significance were expressed by * for P<0.05, ** for P<0.01, *** for P<0.001 and **** for P<0.0001.

Supporting information

eLife Assessment

This study provides convincing evidence that the Kinesin protein family member KIF7 regulates the development of the cerebral cortex and its connectivity and the specificity of Sonic Hedgehog signaling by controlling the details of Gli repressor vs activator functions. This study provides important new insights into general aspects of cortical development.

Reviewer #1 (Public review):

Summary:

This is an interesting follow-up to a paper published in Human Molecular Genetics reporting novel roles in corticogenesis of the Kif7 motor protein that can regulate the activator as well as the repressor functions of the Gli transcription factors in Shh signalling. This new work investigates how a null mutation in the Kif7 gene affects the formation of corticofugal and thalamocortical axon tracts and the migration of cortical interneurons. It demonstrates that the Kif7 null mutant embryos present with ventriculomegaly and heterotopias as observed in patients carrying KIF7 mutations. The Kif7 mutation also disrupts the connectivity between the cortex and thalamus and leads to an abnormal projection of thalamocortical axons. Moreover, cortical interneurons show migratory defects that are mirrored in cortical slices treated with the Shh inhibitor cyclopamine suggesting that the Kif7 mutation results in a down-regulation of Shh signalling. Interestingly, these defects are much less severe at later stages of corticogenesis.

Strengths/weaknesses:

The findings of this manuscript are clearly presented and are based on detailed analyses. Using a compelling set of experiments, especially the live imaging to monitor interneuron migration, the authors convincingly investigate Kif7's roles and their results support their major claims. The migratory defects in interneurons and the potential role of Shh signalling present novel findings and provide some mechanistic insights but rescue experiments would further support Kif7's role in interneuron migration. Similarly, the mechanism underlying the misprojection which has previously been reported in other cilia mutants remains unexplored. Taken together, this manuscript makes novel contributions to our understanding of the role of primary cilia in forebrain development and to the aetiology of neural symptoms in ciliopathy patients.

Reviewer #2 (Public review):

Summary:

This study investigates the role of KIF7, a ciliary kinesin involved in the Sonic Hedgehog (SHH) signaling pathway, in cortical development using Kif7 knockout mice. The researchers examined embryonic cortex development (mainly at E14.5), focusing on structural changes and neuronal migration abnormalities.

Strengths:

(1) The phenotype observed is interesting, and the findings provide neurodevelopmental insight into some of the symptoms and malformations seen in patients with KIF7 mutations.

(2) The authors assess several features of cortical development, including structural changes in layers of the developing cortex, connectivity of the cortex with the thalamus, as well as migration of cINs from CGE and MGE to the cortex.

Weaknesses:

(1) The Kif7 null does have phenotype differences from individual mutations seen in patients. It would be interesting to add more thoughts about how the null differs from these mutants in ciliary structure and SHH signaling via the cilium.

(2) The description of altered cortex development at E14.5 is perhaps rather descriptive. It would be useful to assess more closely the changes occurring in different cell types and stages. For this it seems very important to have a time course of cortical development and how the structural organization changes over time. This would be easy to assess with the addition of serial sections from the same mice. It might also be interesting to see how SHH signaling is altered in different cortical cell types over time with a SHH signaling reporter mouse.

(3) Abnormal neurodevelopmental phenotypes have been widely reported in the absence of other key genes affecting primary cilia function (Willaredt et al., J Neurosci 2008; Guo et al., Nat Commun 2015). It would be interesting to have more discussion of how the Kif7 null phenotype compares to some of these other mutants.

(4) The authors see alterations in cIN migration to the cortex and observe distinct differences in the pattern of expression of Cxcl12 as well as suggest cell-intrinsic differences within cIN in their ability to migrate. The slice culture experiments though make it a little difficult to interpret the cell intrinsic effects on cIN of loss of Kif7, as the differences in Cxcl12 patterns still exist presumably in the slice cultures. It would be useful to assess their motility in an assay where they were isolated, as well as assess transcriptional changes in cINs in vivo lacking KIF7 for expression patterns that may affect motility or other aspects of migration.

Author response:

The following is the authors’ response to the original reviews

Reviewer #1 (Public review):

Summary:

This is an interesting follow-up to a paper published in Human Molecular Genetics reporting novel roles in corticogenesis of the Kif7 motor protein that can regulate the activator as well as the repressor functions of the Gli transcription factors in Shh signalling. This new work investigates how a null mutation in the Kif7 gene affects the formation of corticofugal and thalamocortical axon tracts and the migration of cortical interneurons. It demonstrates that the Kif7 null mutant embryos present with ventriculomegaly and heterotopias as observed in patients carrying KIF7 mutations. The Kif7 mutation also disrupts the connectivity between the cortex and thalamus and leads to an abnormal projection of thalamocortical axons. Moreover, cortical interneurons show migratory defects that are mirrored in cortical slices treated with the Shh inhibitor cyclopamine suggesting that the Kif7 mutation results in a down-regulation of Shh signalling. Interestingly, these defects are much less severe at later stages of corticogenesis.

Strengths/weaknesses:

The findings of this manuscript are clearly presented and are based on detailed analyses. Using a compelling set of experiments, especially the live imaging to monitor interneuron migration, the authors convincingly investigate Kif7's roles and their results support their major claims. The migratory defects in interneurons and the potential role of Shh signalling present novel findings and provide some mechanistic insights but rescue experiments would further support Kif7's role in interneuron migration. Similarly, the mechanism underlying the misprojection which has previously been reported in other cilia mutants remains unexplored. Taken together, this manuscript makes novel contributions to our understanding of the role of primary cilia in forebrain development and to the aetiology of neural symptoms in ciliopathy patients.

We again thank Reviewer 1 for her/his positive assessment of our article. We have addressed several weaknesses identified by the reviewer, supplementing the initial results with new data, and correcting or clarifying the text where necessary. Our detailed responses to the reviewer’s recommendations appear at the end of each comment.

Reviewer #1 (Recommendations for the authors):

(1) The authors report remarkable phenotypic changes in E14.5 embryos in the projection patterns of corticofugal/thalamocortical axons and in interneuron migration, but some of those phenotypes appear much less severe at E16.5. This might be indicative of a delay in development. Does the migration of interneurons to more dorsal regions correspond to an extended Cxcl12 expression? Do interneuorons still show migratory defects at E16.5? To address a potential delay, the authors could, if feasible, repeat Tbr2/Tomato and L1 or neurofilament stainings in E18.5 embryos?

The question of a possible developmental delay in Kif7 -/- embryos is important. To document this topic, we have extended our study initially focused on embryonic stage E14.5 to earlier (E12.5) and later (E16.5, E18.5/P0) developmental stages. We added new data on E12.5 (Fig. 1, Fig. 3, Fig. S4) and E18.5 (Fig. 3, Fig. 4) embryos in the main figures, and considerably extended the data on E16.5 embryos (Fig. 1, Fig. 3). The legends of figures and the text of the result section (p5-p6) have been modified accordingly. We now describe developmental defects in Kif7 -/- embryos, which are not simple developmental delays. The sequences of thalamic axon development and cIN migration are representative of this complexity.

Thalamic axons: the pioneer projection is misrouted to the amygdala at E14.5 (Fig. 4B) whereas most Kif7 -/- thalamic axons extend to the cortex at E16.5, with a slight delay compared to WT axons (Fig. 4D). At E18.5, the Kif7 -/- thalamo-cortical projection appears rather normal in the rostral forebrain but is drastically reduced in the median and caudal forebrain (Fig. 4E). This strong decrease is confirmed by neurofilament staining performed at E18.5 which identifies a major loss of corticofugal and thalamo-cortical projections in Kif7 -/- brains (Fig. 4F).

Migrating cIN: During normal development, CXCL12 maintains cIN in their tangential pathways as they start to colonize the cortical wall (E13.5/E14.5). Then CXCL12 drops in the SVZ (Tiveron et al., 2006; Caronia-Brown and Grove, 2011) allowing wild type cIN to invade the cortical plate (Stumm et al., 2003; Li et al., 2008; Atkins et al., 2023). In Kif7 -/- embryos, CXCL12 is never expressed in the SVZ of the dorsal cortex. Therefore Kif7 -/- cIN migrate radially in the dorsal cortex instead of tangentially. We have improved our text in the result section to clarify this transient defect (p8-9).

(2) Figure 1D: The overview of the Gsh2 and Tbr2 stainings does not allow us to see details of the PSPB. The lines indicating the position of the PSPB are not helpful either. Higher magnifications are required to see whether there are subtle differences at these boundaries as observed for other cilia mutants.

We thank the reviewer for her/his question that allowed us to identify a mild default of patterning at the PSB, illustrated by high magnification pictures in the Fig. 1D and described in the result section (p5). This subtle defect of PSB patterning is consistent with previous observations in Kif7 -/- embryos (Putoux et al, 2019) and appears milder than the PSB defect in hypomorphic Gli3 Pdn mutants (PSB shifted dorsally and less well defined as illustrated in Kuschel et al, 2003 and Magnani et al., 2010).

(3) Figure 3: The authors report an interesting mis-projection of thalamocortical axons towards the amygdala. A very similar pattern has been described in Gli3 hypomorphic Pdn mutants (Magnani et al., 2010), in Rfx3, and in Inpp5e null mutant embryos (Magnani et al., 2015). These papers lend further support that this Kif7 phenotype is Gli3 dependent and should be cited in the manuscript. Moreover, the mechanism(s) underlying this mis-projection remain unexplored. Is this phenotype rescued in the previously reported Kif7/ Gli3D699 double mutants? Is there an abnormal expression of axon guidance molecules?

We deeply thank the reviewer for drawing our attention to the abnormal projection of thalamic axons to the amygdala described in the Gli3 Pdn mutant and in two ciliary mutants, Rfx3 -/- and Lnpp5e -/-. We cite these two papers (Magnani et al., 2010, 2015) in the revised manuscript (p7). In the Gli3 Pdn mutant, transplantation experiments show that a patterning defect of the ventral telencephalon (VT) underlies the mis-projection of the thalamus to the amygdala (Magnani et al, 2010). In the Rfx3 ciliary mutant, two possible mechanisms are proposed: pre-thalamus patterning defect and ectopic Netrin and Slit1 expression in the VT (Magnani et al, 2015). We do agree that understanding the mechanism of the thalamic misprojection in the Kif7 mutant would be of great interest. However, given the complexity of the putative mechanisms described in the Gli3 Pdn and Rfx3 mutants, we believe that this question deserves further investigation in a future study. Finally, the possibility that the thalamic projection defect observed in Kif7 -/- embryos could be rescued in Kif7/Gli3699 (double mutants in which Gli3R is overexpressed in the dorsal and ventral forebrain) is very unlikely. Our two main arguments are:

(1) Magnani et al (2015) did not rescue the TCA pathfinding defect in the Rfx3 -/- ciliary mutant when they overexpressed GLI3-R (see TCA description in the Rfx3/ Gli3699 double mutant, last paragraph of the result section). The authors concluded “This finding could be explained by a requirement for Gli activator and not Gli repressor function in VT {ventral telencephalon} patterning and indeed, Gli3 western blots showed that the levels of Gli3R are not altered in the VT of Rfx3 -/- embryos”.

(2) The GLI3-R/Gli3-FL ratio is decreased in the cortex of the Kif7 -/- embryos (dorsal telencephalon) as expected, whereas it is very low in the MGE of WT embryos (ventral telencephalon) and remains unaltered in the Kif7 -/- embryos (Fig. 2B).

Similarly, the analysis of Kif7 -/- cIN migratory defects leads us to conclude that Kif7 ablation impairs Gli activation function rather than Gli repressor function in the VT where cIN are generated.

(4) Figure 4: The authors should discuss the difference between Tbr2 and Cxcl12 expression which does not extend into the dorsal-most cortical SVZ.

We observed that the transient CXCL12 expression is lacking in the SVZ of the dorsal cortex of Kif7 -/- embryos at E14.5, in a region where TBR2 cells abnormally reach the cortical surface and intermingle with post-mitotic cells. A sentence in our previous version (lines 233-234) could suggest a link between the abnormal location of TBR2 expressing cells and the lack of CXCL12 expression. Having found no data in the literature to explain the absence of CXCL12 expression in the brain by an abnormal cellular environment or by a defect in transcription factor expression, we do not want to further elaborate on differences and similarities between TBR2 and CXCL12 expression patterns in the Kif7 -/- brain. We have modified our text accordingly in the result section of the revised manuscript (p8-9).

(5) Figure 5: The authors convincingly describe migratory defects of interneurons. The treatment with Shh agonist and antagonist provides some mechanistic insights but genetic or pharmacological rescue experiments would lend further support. For example, they could treat Kif7 mutant sections with Shh agonists or analyse Kif7/Gli3D699 double mutants.

We thank the reviewer for her/his positive assessment of our analysis of the cIN migration. Unfortunately, the rescue experiments proposed by the reviewer should not help to further support our conclusions. First, Kif7 ablation in cIN prevents the processing of any SHH signal in the transcriptional pathway. Second, increasing GLI3R by crossing Kif7 -/- animals with Gli3D699 mice could possibly rescue the alterations of layering in the dorsal cortex where the GLI3R/GLI-FL ratio is strongly decreased and the SHH pathway activated. Such a rescue had been previously described for corpus callosum defects (Putoux et al., 2019). However, because cIN are generated in the ventral forebrain where SHH signaling predominantly activates the formation of GLI-A and where Kif7 ablation does not alter the GLI3 ratio, GLI3R re-introduction in the basal forebrain should rather increase the migratory defects of Kif7 -/- cIN instead of producing a rescue. To further support our conclusion, we analyzed the migratory behavior of Kif7 -/- cIN in a WT cortical environment. The results illustrated in the Fig. 6A and described in page 9 of the result section confirm that the migration defects of Kif7 -/- cIN are reminiscent of an inhibition but not an activation of the transcriptional SHH pathway (same phenotype as in Kif3a ciliary mutants described in Baudoin et al, 2012).

(6) Figure 6: The authors describe the Shh mRNA and protein expression with relevance to interneuron migration. In contrast to the in situ hybridisation, the immunofluorescence analysis is not very convincing and requires further controls. The authors should at least show a no primary antibody control and, if available, could include a staining on Shh mutants. These additional controls are important as Shh protein expression in the developing cortex is highly controversial and a recent paper describes a different pattern (Manuel et al., 2022: https://journals.plos.org/plosbiology/article?id=10.1371/journal.pbio.3001563#). Moreover, it remains unclear whether the Shh protein expression is uniform within the cortex or follows lateral to medial or ventricular to pial gradients. A more thorough description and corresponding figures would be helpful.

Manuel et al. (2022) used the SHH KO (generated by Chiang et al., 1996) that develops a long proboscis to validate the rabbit anti-SHH antibodies (from Genetech) used in their study. They show a lack of SHH signal in the SHH KO. However, it is difficult to identify the cortex in this mouse line and the authors did not specify which part of the SHH protein was used to generate antibodies. We wished to use the SHH KO generated by Chiang and backcrossed on a C57B/6 line (Rash and Grove, 2007) that develops a layered neocortex at E17.5. However,

(1) the SHH KO was obtained by replacing exon2 with a PGK-neo cassette and could express a 101 aa truncated protein comprising the N-ter part of the protein, and

(2) the antibody we used, is a polyclonal N-ter antibody that targets the active SHH protein (Cys25-Gly198 part of SHH protein used as immunogen to produce the antibody). We thus thought that this labeling experiment will not give information on the specificity of the antibody, some epitopes being able to recognize the truncated protein produced in the SHH KO.

To overcome the lack of a good mutant mice to validate the SHH N-ter antibodies, we analyzed the SHH immunostaining pattern at E12.5 and compared the expression profile with previously published SHH mRNA expression patterns. The border of the third ventricle and the ZLI were strongly immunostained by SHH-Nter antibodies and these regions were shown to express SHH mRNA at E12.5-E13.5 (Kicker et al. 2004, Loulier et al., 2005, Sahara et al., 2007 and Fig. 7B1). In brain sections at E14.5, only the choroid plexus was strongly labeled and some structures showed diffused labeling. We analyzed the distribution of SHH mRNAs in the cortex using a highly sensitive technique (RNAscop) at E14.5 and showed that very few cortical cells expressed SHH mRNA and at very low level. Anti-SHH-Nter antibodies immunostained numerous bright dots throughout the cortical neuropile, which is not surprising for a diffusible factor like SHH. However, the labeling was not homogeneous and showed a ventricle to pial gradient at E12.5 and aligned distributions in the different cortical layers at E14.5. We have described the expression pattern in more detail and modified the Fig. S4 by adding an image of immunostaining performed without SHH N-ter antibody.

(7) Figure S1: The Gli3 Western blot needs to be quantified. As the authors only show one control and one mutant sample, it remains unclear how representative this blot is. In addition to Gli3R and Gli3FL, the authors should also determine the ratio of both isoforms. Are there also differences in the MGE?

We now produce results of Gli3 western blots in the cortex and MGE of several E14.5 Kif7 KO (n=4) and WT (n=4) embryos. The GLI3R/GLI3FL ratio has been determined in the cortex and in the MGE of WT and mutant embryos. Results are illustrated in the Fig. 2.

Minor points:

The authors should carefully amend the literature on Gli genes and forebrain development. For example:

(1) Line 85: Add Hasenpusch-Theil et al., 2018.

We added this reference.

(2) Line 141: Remove Magnani et al., 2010 (they characterized hypomorphic Gli3 Pdn mutants) and replace with Kuschel et al., 2003.

Since our revised figure 2 illustrates GLI3 western blots and compare GLI3R/GLI3FL ratios in the cortex and MGE of WT and Kif7-/- embryos, we no longer cite these papers in the result section.

(3) Line 380: Replace reference with Theil, 2005.

We have replaced Magnani et al, 2014 by Theil 2005 in the sentence.

(4) Line 414: Rallu et al is not an appropriate reference for this as this manuscript does not investigate the expression of a single cortical marker in Shh/Gli3 double mutants.

We removed the reference Rallu et al. in the sentence.

(5) Reference in line 355: do not use Vancouver style.

We apologize for the mistake that was corrected.

(6) Spelling: Line 447 it should read "choroid plexus"

We again apologize for the mistake that has been corrected.

Reviewer #2 (Public review):

Summary:

This study investigates the role of KIF7, a ciliary kinesin involved in the Sonic Hedgehog (SHH) signaling pathway, in cortical development using Kif7 knockout mice. The researchers examined embryonic cortex development (mainly at E14.5), focusing on structural changes and neuronal migration abnormalities.

Strengths:

(1) The phenotype observed is interesting, and the findings provide neurodevelopmental insight into some of the symptoms and malformations seen in patients with KIF7 mutations.

(2) The authors assess several features of cortical development, including structural changes in layers of the developing cortex, connectivity of the cortex with the thalamus, as well as migration of cINs from CGE and MGE to the cortex.

We greatly thank Reviewer 2 for her/his positive assessment of our work that characterize the neurodevelopmental defects induced by KIF7 ablation. We have deeply reorganized and implemented data in the figures to show changes occurring in different cortical cell types and at different stages. We have moreover corrected and clarified the text where necessary. Our detailed responses to the reviewer’s recommendations appear at the end of each comment.

Weaknesses:

(1) The Kif7 null does have phenotype differences from individual mutations seen in patients. It would be interesting to add more thoughts about how the null differs from these mutants in ciliary structure and SHH signaling via the cilium.

We are grateful to the Reviewer for recalling that Kif7 ablation alters SHH signaling within primary cilium and has a strong effect on ciliary structure. In the revised version of the manuscript, we discuss data from the literature that describe these alterations in human (Putoux et al, 2011) and in murine KIF7 depleted cells (He et al, 2015; Cheung et al., 2009; Lai et al., 2021) (discussion p13).

(2) The description of altered cortex development at E14.5 is perhaps rather descriptive. It would be useful to assess more closely the changes occurring in different cell types and stages. For this it seems very important to have a time course of cortical development and how the structural organization changes over time. This would be easy to assess with the addition of serial sections from the same. It might also be interesting to see how SHH signaling is altered in different cortical cell types over time with a SHH signaling reporter mouse.

We thank the Reviewer for her/his request that helped us to improve our description of developmental defaults in the Kif7 -/- cortex. In the revised manuscript, we have expanded our study initially focused on embryonic stage E14.5 to earlier (E12.5) and later (E16.5, E18.5 /P0) developmental stages. Instead of focusing on median forebrain sections, we have expanded our observations to rostral and caudal sections. Altogether, these new observations allow us to describe more precisely the complex developmental defects in the Kif7 -/- cortex over time, in specific cortical regions (dorsal versus lateral cortex, and rostral versus caudal levels). Figures 1, 3, 4, and S4 have been deeply edited to present new data on E12.5 (Fig. 1, Fig. 3, Fig. S4), E16.5 (Fig. 1, Fig. 3) and E18.5 (Fig. 3, Fig. 4) embryos. We have modified the legends and text in the result section (p5-6) accordingly. We agree with the Reviewer that deciphering how SHH signaling is altered in the different cortical cells over time should be highly interesting and relevant. Nevertheless, we anticipate complex analyses and consider that they should be retained for future studies.

(3) Abnormal neurodevelopmental phenotypes have been widely reported in the absence of other key genes affecting primary cilia function (Willaredt et al., J Neurosci 2008; Guo et al., Nat Commun 2015). It would be interesting to have more discussion of how the Kif7 null phenotype compares to some of these other mutants.

We agree with this Reviewer concern. In the revised manuscript, we discuss our results with regard to previous observations in other ciliary mutants. The murine cobblestone mutant described in Willaredt et al. (2008) indeed shows defects similar to those we describe in the Kif7 -/- mouse. We thank again the Reviewer for her/his helpful comment that allowed us to strengthen and better interpret our results. Guo et al (2015) did not conduct a study of ciliary mutants. Nevertheless, their characterization of cortical developmental defects following invalidation of genes involved in human ciliopathies identified cell autonomous defects in cortical progenitors and in differentiating cortical neurons, which corroborate our observations (p.15)

(4) The authors see alterations in cIN migration to the cortex and observe distinct differences in the pattern of expression of Cxcl12 as well as suggest cell-intrinsic differences within cIN in their ability to migrate. The slice culture experiments though make it a little difficult to interpret the cell intrinsic effects on cIN of loss of Kif7, as the differences in Cxcl12 patterns still exist presumably in the slice cultures. It would be useful to assess their motility in an assay where they were isolated, as well as assess transcriptional changes in cINs in vivo lacking KIF7 for expression patterns that may affect motility or other aspects of migration.

To circumvent the difference in the expression profile of CXCL12 in the dorsal cortex of WT and Kif7 -/- embryos on the migratory behavior of cIN, we compared the trajectories and dynamics of WT and Kif7 -/- cIN imaged in the lateral cortex where CXCL12 expression appears similar in WT and Kif7 -/- brains.

We moreover followed the reviewer recommendation and analyzed the migratory behavior of Kif7 -/- cIN that migrate as isolated cells on a dissociated substrate of WT cortical cells. We sincerely thank the reviewer for her/his suggestion as the results revealed an interesting and relevant ciliary phenotype in migrating Kif7 -/- cIN. This additional experiment confirms that Kif7 -/- cIN exhibit the same migratory defects as those initially characterized in the Kif3a -/- ciliary mutant. The new results are illustrated in the Fig. 6A and described in the result section (p9). We agree with the reviewer that the analysis of transcriptional changes that could affect Kif7 -/- cIN motility and migration would be very interesting to study, but this study is beyond the scope of the present article.

Version 2

Abstract

Abstract

Mutations of KIF7, a key ciliary component of Sonic hedgehog (SHH) pathway, are associated in humans with cerebral cortex malformations and clinical features suggestive of cortical dysfunction. KIF7 regulates the processing of GLI-A and GLI3-R transcription factors in a SHH-dependent manner both in humans and mice. Here, we examine the embryonic cortex development of a mouse model that lacks the expression of KIF7 (Kif7 -/-). The cortex is composed of principal neurons generated locally in the dorsal telencephalon where SHH expression is low and inhibitory interneurons (cIN) generated in the ventral telencephalon where SHH expression is high. We observe a strong impact of Kif7 deletion on the dorsal cortex development whose abnormalities resemble those of GLI3-R mutants: subplate cells are absent, the intermediate progenitor layer and cortical plate do not segregate properly, and corticofugal axons do not develop timely, leading to a delayed colonization of the telencephalon by thalamo-cortical axons. These structural defects alter the cortical distribution of cIN, which moreover exhibit intrinsic migration defects and cortical trajectories resembling those of cyclopamine-treated cIN. Our results show that Kif7 deletion impairs the cortex development in multiple ways, exhibiting opposite effects on SHH pathway activity in the developing principal neurons and inhibitory interneurons.

Introduction

The primary cilium is a tiny microtubule-based organelle present on the surface of nearly all mammalian cell types including neurons, which functions as a signaling hub and transduces several signaling pathways comprising Sonic Hedghog (SHH), Wnt, Delta/Notch and mTOR pathways 1. Primary cilium dysfunction causes pleiotropic diseases named ciliopathies. KIF7 is a ciliary kinesin responsible for the trafficking and the positive and negative regulation of the GLI transcription factors in the primary cilium of mammals 2. Loss of function of KIF7 in the mouse has shown that KIF7 regulates SHH signalling by acting downstream of Smoothened (Smo) and upstream of GLI2 and GLI3 3–7. Studies in mice established that KIF7 activity is dependent on the expression level of SHH. In the absence of SHH, KIF7 localizes to the base of the primary cilium 5. GLI factors are phosphorylated and addressed to the proteasome at the base of the primary cilium for degradation, leading to the formation of a cleaved and stable transcriptional repressor (GLI3-R). In the presence of SHH, KIF7 accumulates at the distal tip of the primary cilium 4,5 and associates with full length GLI2/3 that become transcriptional activators (GLI-A) 8.

Since 2011, ten studies have identified patients carrying mutations in the KIF7 gene responsible of ciliopathies classified according to clinical features as hydrolethalus, acrocallosal, Joubert and Greig cephalopolysyndactyly syndromes 9–18. MRI investigations revealed macrocephaly, ventricles enlargement and corpus callosum alterations. They also showed the characteristic hindbrain abnormalities observed in most ciliopathies such as molar tooth sign (MTS) and cerebellar atrophy. Patients also presented with mild but frequent cortical malformations as well as neurodevelopmental delay, intellectual disability and seizures 11,12,15,19 which indicate cortical abnormalities.

In physiological conditions, the proper activity of the cortex relies on the excitatory/inhibitory balance i.e. on the ratio, positioning and connectivity of excitatory glutamatergic neurons (principal neurons, PN) and inhibitory GABAergic interneurons (cIN), generated from progenitors in the dorsal and ventral telencephalon, respectively. The cortical neurons of dorsal and ventral origin each migrate towards the developing cortex, the so-called cortical plate (CP), according to a specific timeline allowing the establishment of proper connections in the CP.

SHH plays a central role in the forebrain patterning and differentiation. At early embryonic stage, the ventral expression of SHH orchestrates the ventro-dorsal 20,21 and medio-lateral 22 regionalization of the mouse forebrain and the differentiation of ventral cell types 20. Shh ablation performed after forebrain patterning alters the specification of distinct subgroups of GABAergic interneurons 23–26. Interestingly, conditional ablation of Shh and Smo in the embryonic cortex reduces the proliferation of dorsal progenitors 27, demonstrating a minimal SHH expression in the dorsal telencephalon, even before birth 28. Beside its role on proliferation and specification, SHH also controls the migration of interneurons to the cortical plate 29–33. In the embryonic telencephalon, GLI transcription factors mediate SHH signals in complex and specific ways. The three GLI factors identified in mammals are expressed in the mouse forebrain: GLI1 along the source of SHH in the ganglionic sulcus, and GLI2 and GLI3 dorsally to the SHH source 26,34,35. GLI1 acts uniquely as a pathway activator whereas GLI2 and GLI3 can be processed in transcriptional activators or inhibitors in the primary cilium. However, both GLI1 and GLI2 function primarily as transcriptional activators in response to SHH activity in the ventral forebrain 35. Nevertheless, GLI1 and GLI2 mutants show mild phenotypes 36,37 and GLI2 seems required to transduce high level SHH signals in mice 38. In contrast, the development of the dorsal cortex, where principal excitatory neurons differentiate, depends mainly on the expression of the GLI3-R repressor, in agreement with the low cortical expression of SHH 39,40. Following patterning, Gli3R/A ratio remains critical for specifying the fate of cortical progenitors and regulating cell cycle kinetics 41–43.

Previous studies investigated the mechanisms underlying the corpus callosum agenesis in patients with KIF7 mutation using Kif7 knock-out (Kif7 -/-) mice 44 and the consequence of KIF7 knock down on cortical neurogenesis in principal cells electroporated with Kif7 shRNA45. It remains that the influence of developmental abnormalities associated with ciliopathies on the cortical cytoarchitecture is poorly understood. Given the crucial role of KIF7 in regulating the GLI pathways and the important role of Gli3-R and GLI2/3-A in regulating the early developmental stages of the dorsal and ventral telencephalon, respectively, we examined here the cortical development from E12.5, the establishment of long distance projections with the thalamus and the migration of GABAergic interneurons (cIN) in Kif7 -/- mice. The migratory behavior of Kif7 -/- cINs was investigated by time-lapse videomicroscopy in co-cultures and in organotypic cortical slices. We also compared the migration of cINs in Kif7 -/- cortical slices and in control slices treated with pharmacological activator and inhibitor of the SHH pathway. We moreover determined the local distribution of the SHH protein in the embryonic cortex.

Our results show developmental defects leading to permanently displaced neurons, abnormal formation of cortical layers and defective cortical circuits that could be responsible for epilepsy and/or intellectual deficiency in patients carrying KIF7 mutation 9,11–18,44.

Results

Kif7 knock-out mice as a model to investigate the structural cortical defects that could lead to clinical feature in patients carrying KIF7 mutation

The KIF7 gene is located on chromosome 15 in human and encodes a 1343 aa protein containing, a kinesin motor domain and a GLI-binding domain in the N-terminal part followed by a Coiled-coil region and a Cargo domain able to bind a diverse set of cargos in the C-terminal part 46. Table 1 summarizes the clinical features associated with mutations targeting either the kinesin or GLI binding domain, the Coiled-coil domain, or the N-terminal Cargo domain. Interestingly, various mutations are associated with the same clinical picture, suggesting that KIF7 mutations, whatever their nature, could lead to protein loss of function, for example by altering the protein structure and solubility as proposed by Klejnot et Kozielski 47. All patients carrying mutation in the KIF7 gene have developmental delay (DD) and intellectual deficit (ID) associated with classical defects of ciliopathies (ventricle enlargement, macrocephaly, corpus callosum agenesis and MTS). Some patients have additional anatomical cerebral cortex defects such as poor frontal development, atrophy, pachygyria and heterotopia 9,10,13,16, which could participate not only to DD and ID but moreover to seizure as observed in a quarter of patients. In this context, we used a murine model in which the Kif7 gene had been deleted (Kif7 -/-) 3 to investigate the consequence of KIF7 loss of function on the cortex development. Kif7 -/- mice have been previously characterized as dying at birth with severe malformations, skeletal abnormalities (digits and ribs), neural tube patterning defects including exencephaly in a third of the mutants, microphtalmia, lack of olfactory bulbs and CC agenesis 3,4,44. In the present study, the KIF7 loss of function was analyzed in a mouse strain with a mixed genetic background in which exencephaly was observed at the same frequency in Kif7 -/- and control embryos. As previously reported, Kif7 -/- embryos displayed microphtalmia [illustrated at embryonic stage 14.5 (E14.5), Fig. 1A, black arrow] and polydactily (not illustrated). Interestingly, skin laxity previously reported in ciliopathic patients 48,49 was observed in all mutants (Fig. 1A, white arrow). Comparison of brains from Kif7 -/- embryos and control littermates revealed the absence of olfactory bulbs in Kif7 -/- brains (illustrated at E14.5, Fig. 1B, white arrows on ventral face view).

Kif7 deletion alters cortical development on latero-medial and rostro-caudal axes.

(A) Kif7 -/- embryos are microphtalmic (black arrow) and exhibit skin laxity (white arrow). (B) External examination of the brain reveals the lack of olfactory bulbs (white arrows, left panel) and the thinning of the dorsal telencephalon (black arrow, right panel). (C) DAPI staining of rostro-caudal series of coronal sections at E12.5 (C1), E14.5 (C2) and E16.5 (C3) illustrates the anatomical defects of Kif7 -/- embryonic brains quantified at E14.5 in C4. The ventricles of Kif7 -/- embryos are strongly enlarged (upper left graph; WT, n=4-6, Kif7 -/-, n=5-6 depending on the rostro-caudal level), their cortical thickness strongly decreased (upper right graph; (WT, n=3; Kif7 -/-, n= 4), resulting in minimal brain width (lower left graph; WT, n=5-6; n=5-6 for Kif7 -/- depending on the rostro-caudal level) and height (lower right graph; WT, n=5-6; Kif7 -/-, n=4-6 depending on the rostro-caudal level) changes. Statistical significance was tested by Two-way ANOVA or mixed model (GraphPad 8.1.0). For ventricle surface and cortex thickness, mix model reveals a genotype effect (p<0.0001). For brain width and height, no genotype effect was observed, but a significant effect of the rostro-caudal level on brain width (P=0.0441) and height (P=0.0092). (D) The pallium-subpallium boundary identified by the limit of expression of ventral (GSH2, D1, D2) and dorsal (TBR2, D3, D4) telencephalic markers is less precisely defined and slightly shifted to the ventricular angle in E13-E14 Kif7 -/- embryos compared to wild type embryos (epifluorescent low magnification, D1, D3; confocal high magnification images, D2, D4). Graphs in C4 represent the means and S.E.M. Cx, cortex; Hip, hippocampus; LGE, lateral ganglionic eminence; MGE, median ganglionic eminence; CGE, caudal ganglionic eminence; Th, thalamus.. Scale bars, 500 µm (C,D1,D3), 150 µm (D2,D4).

Kif7 invalidation alters the development of the cortex

The thinning of the dorsal cortex at E14.5 allowing the lateral ventricles to be seen through (Fig. 1B, black arrow on dorsal face view) proned us to investigate the structural organization of the forebrain over time (Fig. 1C1-3, E12.5-E16.5). Analyses performed on rostro-caudal series of frontal sections identified strong abnormalities in the cortex development. E12.5 Kif7 -/- embryos exhibited cortical wall folding and thinning in rostro-median sections (Fig. 1C1, arrowheads) and prominent diencephalon (Fig. 1C1, stars). At E14.5, cortex folding was not longer observed in Kif7 -/- embryos which exhibited enlarged lateral ventricles (compare left and right panels in Fig.1C2). The cortex remained thinned in the dorsal region and the latero-dorsal extension of the telencephalic vesicle was strongly reduced caudally. These defects were maintained in E16.5 embryos (Fig. 1C3) and the hippocampus did not develop properly (Fig. 1C3, caudal section). A quantitative analysis performed in E14.5 brains (Fig.1C4) confirmed the ventriculomegaly (upper left graph) and the cortex thinning (upper right graph) in the rostral and median sections despite unchanged total width and height of the telencephalon (Fig. 1C4, lowers graphs). KIF7 depletion in E10.5-E11.5 mouse embryos was reported to alter the dorso-ventral patterning of the medial forebrain 44. The frontier of expression of GSH2 (Fig. 1D1,D2) and TBR2 (Fig. 1D3,D4), two transcription factors expressed in the ventral and dorsal telencephalon respectively, was slightly abnormal at E13-E14 in Kif7 -/- embryos with a small shift of GSH2(+) cells toward the ventricular angle (Fig. 1D2) and a dispersion of some TBR2(+) cells at the pallium-subpallium boundary (PSB) (Fig. 1D4). This suggests a near-normal dorso-ventral patterning of the telencephalic vesicles, however with a discrete blurring of the dorsal and ventral markers at the PSB.

In Kif7 -/- embryos, the cleavage of full length GLI3 (GLI3-FL) into the transcriptional repressor GLI3 (GLI3-R) is decreased 3,4. Here, we analyzed the expression of GLI3-FL and GLI3-R in the cortex and in the medial ganglionic eminence (MGE) where cortical interneurons (cIN) are born. We showed minimal expression of GLI3-R and GLI3-FL in the MGE compared to the cortex (Fig. 2A) and a strong decrease in the cleavage of GLI3-FL in GLI3-R only in the cortex of Kif7 -/- embryos (Fig. 2A,B).

Western blot analysis on the cortex and MGE of wild type (WT) and Kif7 -/- embryos at E14.5.

(A) The clived form of GLI3 (GLI3-R at 83 KDa) and the full length GLI3 (GLI3-FL at 190 KDa) are more abundant in the cortex compared to the MGE (see actin band intensity for protein loading). (B) The ratio Gli3-R/Gli3-FL is lower in the MGE than in the cortex of WT animals and is significantly decreased only in the cortex of Kif7 -/- brains compared to control. Two way ANOVA reveals significant interaction between brain structure and genotype (WT, n=4; Kif7 -/-, n=4; P=0.002) and multiple comparisons show a statistical difference between genotype only in the cortex (****, P<0.0001)]. Graph represents the means and S.E.M.

The structural defects described above convinced us to examine the cellular organization of the cortex of Kif7 -/- embryos. At E12.5, the whole cortex is proliferative. From stage E14.5, the mouse cortex can be described as a stack of three specialized domains: i) the upper/superficial cortical layers containing post-mitotic cells generated in the proliferative zones of the cortex, with TBR1(+) post-mitotic neurons located in the cortical plate (CP) in-between the subplate (SP) and marginal zone (MZ) MAP2-expressing cells; ii) the deep proliferative layers, ventricular and subventricular zones (VZ and SVZ, respectively), located along the lateral ventricle; and iii) the intermediate zone (IZ) located between the CP and the proliferative layers, which hosts the radially and tangentially migrating neurons and growing axons. At the end of the embryonic development, the proliferative layers are strongly reduced due to neurogenesis decrease, whereas the cortical thickness had increased due to CP delamination. At E14.5, the TBR1(+) CP cells appeared more clustered in Kif7 -/- embryos than in wild type (WT) embryos where they formed a rather regular, radially organized CP (compare left and right panels in Fig. 3A1,3A2). The thickness of the upper MAP2(+) MZ layer was extremly irregular in Kif7 -/- embryos and MAP2(+) SP cells were missing in the dorsal cortex (Fig. 3A1,A2, white arrows indicate the limit of distribution of MAP2(+) SP cells). In E16.5 and E18.5 embryos, the CP was obviously thinner in the mutant as compared to the WT (Fig. 3B,C), even though CTIP(+) cells differentiated normally in V-VI layers (Fig. 3C). The CP thinning appeared even more pronounced in the caudal than in the median and rostral cortex. We then analyzed the proliferative layers focusing on TBR2(+) intermediate progenitors (IP) which normally form a dense layer in the cortical SVZ (Fig. 3D1, left). The TBR2(+) layer constantly showed an abnormal positioning in the dorsal cortex of Kif7 -/- embryos. TBR2(+) cells indeed reached the brain surface (Fig. 3D1, arrow) where they mixed up with post-mitotic TBR1(+) cells instead of keeping a minimal distance with TBR1 cells as observed in control brains (Fig. 3D2, arrow). As a consequence intermediate progenitors and post-mitotic neurons no longer segregated in the dorsal cortex of Kif7 -/- embryos where no IZ was observed (Fig. 3D2, arrow). Accordingly to the thinning of the cortex starting at E12.5, the TBR2(+) layer was thinner at all development stages analyzed (Fig. 3E,F). In contrast, the abnormal positioning of TBR2(+) cells at the brain surface in the dorsal cortex of E14.5 embryos was no longer observed at E16.5 (Fig. 3F) suggesting either a transient structural disorganization and/or developmental delay of the dorsal cortex in Kif7-/- embryos at E14.5. The TBR2 staining moreover revealed focal heterotopia in the dorsal or lateral cortex of E14.5 Kif7 -/- embryos, either at the ventricular or pial surface (Fig. S1).

Histological alterations in the developing cortex of Kif7 -/- embryos.

(A-C) Immunostaining of cortical post-mitotic layers in E14.5 (A), E16.5 (B) and E18.5 (C) in median coronal sections representative of 5 WT and Kif7 -/- embryos imaged on epifluorescence (A1,B,C) or confocal (A2) microscopes. At E14.5 (A1,A2), the TBR1(+) staining (red) of the cortical plate is more clustered in Kif7 -/- than in WT embryos, and the MAP2(+) staining (green) of the subplate is absent in the dorsal cortex of E14.5 Kif7 -/- embryos (white arrow, right column). The post-mitotic layers remain thinner in Kif7 -/- embryos at later embryonic stages as illustrated by TBR1 staining at E16.5 (B) and CTIP1 staining at E18.5 (C) that specifically labels the deeper cortical layers (V-VI). Moreover, the hippocampus is underdeveloped in the mutant (white arrow) (C). (D-F) Immunostaining of TBR2(+) proliferative layer. At E14.5 (D1,D2), the TBR2(+) layer (green) of secondary progenitors appears disorganized in the lateral cortex of the Kif7 -/- embryos (white arrowhead in D1) and reaches the brain surface in the dorsal cortex of Kif7 -/- embryo where it intermingles with post-mitotic TBR1(+) cells (D2, red) (white arrows). In E12.5 Kif7 -/- embryo (E), the TBR2(+) layer is thin and clustered however its localization in the thickness of the cortex is normal. At E16.5 stage (F), the TBR2(+) layer is normally positionned in the deeper layer of the cortex in both WT and Kif7 -/- embryos. Hip, hippocampus; V, ventricle; VZ, ventricular zone; SVZ, subventricular zone; IZ, intermediate zone; CP, cortical plate; MZ, marginal zone. Scale bars: 250 µm (A1,B,F), 100 µm (A2), 200 µm (D,E); 500 µm (C).

The loss of Kif7 alters the connectivity between the cortex and the thalamus

The intermediate zone (IZ) hosts migrating neurons and the growing cortical and thalamic projections 50. After reaching the CP by radial migration, post-mitotic neurons extended pioneer axons oriented tangentially in the IZ and directed to the PSB 51,52. We thus examined whether the structural defaults of the CP and SP and the lack of IZ in the dorsal cortex of Kif7-/- embryos did associate with developmental abnormalities of corticofugal projections. We labeled corticofugal axons by inserting small crystals of DiI in the CP of E14.5 paraformaldehyde fixed brains. After DiI had diffused along corticofugal axons, we analyzed labeled axons on coronal sections. In both control and Kif7 -/- brains, corticofugal axons extended below the CP toward the PSB (Fig. 4A). According to the latero-medial gradient of cortical development, DiI injections in the dorsal cortex of control and Kif7 -/- brains (Fig. 4A1) labeled much less axons than injections in the lateral cortex (Fig. 4A2,A3). Remarkably, the cortical bundles labeled from similar cortical regions were always much smaller and shorter in Kif7 -/- than in WT brains (compare right and left columns in Fig. 4A). Dorsal injections labeled large bundles crossing the PSB in WT brains, but only a few axons reached the PSB in Kif7 -/- brains (Fig. 4A1). The growth of corticofugal axons appeared severely hampered in the dorsal cortex of Kif7 -/- embryos. Lateral injections in WT brains labeled anterogradely corticofugal axons that reached the embryonic striatum and continued medially in the internal capsule (IC), and labeled retrogradely thalamic projections (Fig. 4A3, left panel). Lateral injections in Kif7 -/- embryos labeled anterogradely cortical axons that spread in the striatum (Fig. 4A2, right panel) and labeled a large bundle directed to the ventral pre-optic area (Fig. 4A3, right panel, white arrow), which was never observed in WT brains. Moreover, no thalamic axons were retrogradely labeled from the lateral cortex in Kif7 -/- embryos.

Kif7 deletion disrupts of the connectivity between the cortex and the thalamus.

(A) Panels illustrate the corticofugal projections labeled by DiI crystal (red dots) positioned in the dorsal (A1) and lateral (A2,A3) cortex of E14.5 WT (left) and Kif7 -/- (right) embryos on vibratome sections performed 30 days after DiI placement and imaged on a macroscope. In Kif7 -/- embryos, less axons project from the dorsal (A1) and lateral (A2) cortex to the subpallium than in WT (compare enlarged views of the projections below A1 and A2). Cortical injections in Kif7 -/- embryos do not label thalamic axons (compare left and right panels in A3) but label a ventral projection (A3, white arrow). (B) Rostro-caudal series of coronal sections (B1-B5) immunostained with anti-Netrin G1a (NG1a) antibodies compare the trajectory of thalamo-cortical axons (TCA) in a E14.5 WT (left) and in a Kif7 -/- (right) embryo imaged on a macroscope. TCA reach the pallium-subpallium boundary of the WT embryo whereas they are lost in the ventral forebrain of the mutant (B4, white arrow). (C) Representative three-dimensional reconstructions (C1, lateral; C2, horizontal; C3, coronal views) of WT and Kif7 -/- E14.5 brains immunostained as a whole with NG1a (red) and TBR1 (green) antibodies that label respectively the TCA and cortical plate cells before transparization and imaging with a light sheet microscope (see Supplementary movies S1 and S2). While all labeled TCA extend in the internal capsule (IC) and a significant proportion of them enter the cerebral cortex in the WT brain, TCA in the Kif7 -/- brain split in two bundles in the basal telencephalon. A bundle stops shortly after entering the IC (white arrowheads) whereas the second bundle extends ventrally (white arrows). (D) At E16.5, TCA are immunostained with NG1a in WT (left) and Kif7 -/- (right) embryos in whole brain with TBR1 antibodies before transparization and imaging (D1) and on coronal section with TBR2 (D2). The TCA extend in the cortex in Kif7 -/- brain, however, fiber density is reduced in the median and caudal brain compared to WT (D1, white arrowhead). TCA invade the post mitotic layers in the cortex (above TBR2 layer) to a lower extend in Kif7 -/- brain (D2). In the Kif7 -/- brain, thick bundles of NG1a(+) fibers project ventrally from the caudal telencephalon, a projection never observed in control brains (D1, white arrow). (E) At E18.5, rostro-caudal series of coronal sections (E1-E4) immunostained with NG1a antibodies compare the trajectory of TCA in WT (left) and Kif7 -/- (right) embryos imaged on a macroscope. TCA reach the dorsal cortex in both WT and Kif7 -/- embryos. However, the TCA projection in Kif7 -/- (arrowheads) become thinner in median sections (E2) and almost disappers in caudal sections (E3,E4). Fiber trajectories in the lateral striatum are abnormal (E2,E3, arrows). (F) Panels illustrate the cortical projections immunolabeled by Smi32 antibodies in E18.5 WT and Kif7 -/- brain. In Kif7 -/- brain, the number of fibers connecting the cortex with other brain structures is strongly reduced (arrow). Cx, cortex; dTh, dorsal thalamus; Hip, hippocampus; Hy, hypothalamus; St, striatum; IC, internal capsule; LGE, lateral ganglionic eminence; MGE, median ganglionic eminence; PSB, pallium-subpallium boundary; .. Scale bars: 250 µm (A,B,C,D), 500 µm (E,F).

To identify the trajectories of thalamic axons in Kif7 -/- embryos, we immunostained coronal sections of E14.5 brains with antibodies against the Netrin G1a (NG1a), an early marker of thalamocortical axons (TCA, Fig. 4B) 53. At E14.5, NG1a antibodies labeled a larger population of neurons in the dorsal thalamus of Kif7 -/- than in WT brains (Fig. 4B, stars). In control brains, TCA made a right angle turn to join the IC in the striatum. Then they extended to the PSB and some axons entered the lateral cortex (Fig. 4B2,B3, left panel). In Kif7 -/- embryos, most thalamic axons formed a thick bundle oriented ventrally in the basal forebrain (Fig. 4B4, white arrow in right panel). A minor structural defect was evidenced at the telo-diencephalic junction by Pax6 staining in Kif7 -/- brains (Fig. S2), recalling the prethalamus defect described in the ciliary Rfx3 mutant 54. Given the complex trajectory of NG1a thalamic axons, we immunostained thalamic axons in whole brains and imaged them after transparisation using a light-sheet microscope. The three-dimensional reconstruction of labeled projections confirmed that thalamic axons made a sharp turn to reach the IC and then navigated straight to the PBS in WT brains (Fig. 4C, left panels and Movie S1). In Kif7 -/- embryos, most thalamic axons stopped their course after leaving the diencephalon, and formed two short bundles: a large one oriented to the IC (Fig. 4C, right panels, arrowheads and Movie S2) and another one oriented to the amygdala, more caudally and ventrally (Fig.4C, white arrows). This last projection was never observed in WT brains. Similar analyses performed at E16.5 showed that two days later, most TCA had reached the cortex in both WT and Kif7 -/- embryos (Fig. 4D, arrowheads) despite a small delay in Kif7 -/- embryos and abnormal trajectories in the PSB region (Fig. 4D1,D2). The ectopic ventral caudal projection to the amygdala was still present at E16.5 in Kif7 -/- embryos (Fig.4D1, arrow). At E18.5, the TCA projections in the Kif7 -/- embryos appeared almost normal in the rostral cortex (Fig. 4E1) but drastically reduced in the median and caudal cortex (Fig. 4E2, E3, E4, arrowheads). In addition, a thick bundle of thalamic axons followed an abnormal trajectory at the PSB (Fig. 4E2,E3, arrow). Neurofilament staining with Smi32 antibodies confirmed the strong reduction of projections in the cortex and IC of Kif7 -/- embryos (Fig. 4F).

Kif7 invalidation alters the cortical distribution of cIN at E14.5

Because cIN are born in the basal forebrain and likely depend on contact/functional interactions with pioneer corticofugal projections for the first stages of their migration 55,56, we examined their distribution in the developing cortex of Kif7 -/- animals. cIN are generated outside of the cortex, in the medial and caudal ganglonic eminences (MGE, CGE) and preoptic area (POA) of the ventral forebrain. They enter the lateral cortex at E12.5 - E13.5, depending on mouse strains, and colonize the whole cortex by organizing two main tangential migratory streams, a superficial one in the MZ, and a deep and large one in the lower TBR2(+) IZ/SVZ 57,58. We thus analyzed the cortical distribution of tdTomato(+) MGE-derived cIN in WT and Kif7 -/- Nkx2.1-Cre;Rosa26-tdTomato transgenic embryos. An abnormal MGE cell distribution was observed in the developing cortex of Kif7 -/- embryos from E14.5 (Fig. 5 and Fig.S3). In mutant brains, both the latero-medial extent of the superficial and deep tangential migratory streams and the space separating them were significantly shortened compared to WT brains (Fig. 5A2,B2, B3). The closeness of the two streams in mutants was consistent with the reduced thickness of the corticofugal and TCA projections that navigate tangentially in the IZ (illustrated in Fig. 4). The superficial migratory stream was thinner and denser in Kif7 -/- embryos compared to WT (Fig. 5A1,5B1, right panels). Remarkably, cIN no longer colonized the dorsal cortex of Kif7 -/- embryos and the deep tangential stream stopped in the region where the layer of TBR2(+) SVZ cells switched to the cortical surface (Fig. 5A1, white arrow). The tangential progression of cIN in the developing cortex is controlled by CXCL12, a chemokine transiently expressed in TBR2(+) cells and in meninges 60,61. CXCL12 prevents the premature occurrence of the tangential to radial migration switch required for cIN to colonize the CP 59–62. In situ hybridization experiments revealed that CxCl12 mRNA was not expressed in the dorsal cortex of E14.5 Kif7 -/- embryos (Fig. 5C, black arrow). To characterize the consequences of the abnormal expression pattern of CXCL12 at E14.5 on the migratory behavior of cIN, we sliced coronally E14.5 Kif7 -/- and WT forebrains and imaged by time-lapse video-microscopy the trajectories of MGE-derived tdTomato-expressing cIN in the dorsal cortex of organotypic slices (Fig.5D). While most cIN presented tangential and oblique trajectories in WT slices, a majority of cIN exhibited radially oriented trajectories in Kif7 -/- slices (Fig.5D). CXCL12 expression in SVZ strongly decreases at E16.5 59,63 and analyses performed at E16.5 in fixed Kif 7 -/- and WT embryos showed that cIN had pursued their migration and reached the dorsal cortex (Fig.S3A) even though the tangential migration of cIN in Kif7 -/- embryos remained delayed as compared to WT embryos. Analyses performed in a surviving P0 animal (Fig. S3B,C) revealed an abnormal distribution of cIN in the CP characterized by a strong decrease of cIN density in the infragranular layers that were moreover much thinner than in control brains (see CTIP2 staining in Fig. S3C).

Abnormal cortical distribution of cIN and Cxcl12 transcript expression in E14.5 in Kif7 -/- brains.

(A) The cortical distribution of cIN is visualized in WT and Kif7 -/- mouse embryos crossed with the Nkx2.1-Cre/R26R-tdTomato strain in which MGE-derived cIN express the fluorescent marker tdTomato. Panels in A1 compare the distribution of tdTomato (+) cIN in WT and Kif7 -/- cortical sections prepared at the same rostro-caudal level and in which SVZ is immunostained with TBR2 antibodies (green). Pictures show that the deep migratory stream of cIN terminates in Kif7 -/- brains in the cortical region where the TBR2(+) layer reaches the cortical surface. Quantitative analysis of the mean length of the deep and superficial migratory streams measured from the entry in the pallium to the last detected cIN in the cortex is illustrated in graph A2. Statistical significance is assessed using Two way ANOVA [layers (WT, n=4; Kif7 -/-, n=3; ***, P=0.001) and genotype (*, P=0.0157)]. (B) Representative pictures (B1) of the deep and superficial tangential migratory streams of cIN in the lateral cortex of WT and Kif7 -/- embryos. Pictures illustrate the decreased thickness of the superficial stream, and the reduced distance between the deep-superficial streams in Kif7 -/-embryos. Graph in B2 (WT, n=4; Kif7 -/-, n=4) compares the distribution of the fluorescence intensity along a ventricle/MZ axis (see grey rectangles in B1) using the plot profile function of FIJI. Curves show no change in the distance between the ventricular wall and the deep cIN, but a significant reduction of the distance between the two streams in the Kif7 -/- cortical sections as quantified on the graph B3 [WT, n=4; Kif7 -/-, n=4; Two way ANOVA reveals a significant interaction between genotype and layer (P=0.0233) and multiple comparisons, a statistical difference between genotype only for the distance between de cIN streams (**, P=0.0051)]. (C) Panels compare the distribution of Cxcl12 mRNA in WT (left panel) and Kif7 -/- (right panel) forebrain coronal sections at E14.5. The WT section shows Cxcl12 transcript enrichment in a deep cortical layer already identified as the SVZ. In the Kif7 -/- cortical section, the expression of Cxcl12 transcripts is reduced to the lateral part of the SVZ. (D) Z-projections of 30 frames acquired during 12h in the dorsal cortex of living organoptypic forebrain slices representative of E14.5 WT and Kif7 -/- embryos with tdTomato expressing cIN. Kif7 -/- cIN were able to migrate dorsally but followed preferentially radially oriented trajectories Scale bars: 200 µm.

The abnormal distribution of cIN in the cortex of Kif7 -/- embryos reflected the structural and molecular abnormalities that we had identified in their developing cortex. We next examined if migratory defaults proper to Kif7 -/- cIN could moreover contribute to their abnormal cortical distribution.

Kif7 invalidation affects the migratory behavior of cIN in co-cultures and organotypic slices

We thus compared the migratory behavior of Kif7 -/- and WT cIN using an in vitro model previously established in the lab 64 to compare the dynamics of mutant and WT cIN migrating on a substrate of WT cortical cells (Fig.6A). WT and Kif7 -/- E14.5 MGE explants from tdTomato expressing embryos were cultured on WT dissociated cortical cells (Fig. 6A1). Fluorescent migrating MGE cells were imaged using time-lapse video-microscopy and tracked over time (Fig.6A2). Dynamic parameters (speed during moves, stops, Fig. 6A3-A6) and trajectories (Fig. 6A7) were analyzed on movies. Kif7 -/- cIN migrated slightly more rapidly than WT cIN (Fig. 6A3) mainly because the duration of their stops was significantly reduced (Fig. 6A4). The saltatory behavior of cIN was preserved (same frequency of fast nuclear movements, Fig. 6A5) despite a slight decrease of the speed of fast nuclear movements (Fig. 6A6). WT cIN migrated along pretty straight trajectories (Fig.6A2, left panel and Fig. 6A7, black curve) whereas Kif7 -/- cIN showed reduced directionality persistence over time (Fig. 6A2, right panel and Fig. 6A7, blue curve). Kif7 ablation thus affected the migratory behavior of cIN in a cell-autonomous manner and the migratory defaults of Kif7 -/- cIN evoked those of Kif3a -/- cIN whose primary cilium is non-functional (See Figure S6 in 29).

Dynamic behavior of migrating cIN in co culture and organotypic cortical slices.

(A) To prepare co-cultures, MGE explants were dissected out of telencephalic vesicles at E14.5 from WT or Kif7 -/- Nkx2.1-Cre;Rosa26-tdTomato embryos and placed on a substrate of WT E14.5 dissociated cortical cells (A1). After 24h in culture, MGE cells migrated centrifugally away from MGE explants on the substrate of cortical cells and were recorded for 7 hours.WT (n=160) and Kif7 -/- (n=126) MGE cells were tracked manually using the MTrackJ plugin allowing to analyze migratory parameters (A2). Box and whisker plots indicate the speed (A3), the time without nuclear movement (A4), the frequency (A5) and speed of nuclear movements (A6). Statistical significance is assessed by Mann Whitney or Unpaired tests. ***, P<0.001 (A4); *, P=0.0308 (A3), P=0.0123 (A5). Directionality persistence was calculated as the ratio of the cell displacement (distance between the first and last positions of the cell) on the cell trajectory (A7, scheme on the right). Graph represented the mean directionality ratios at each time point over the first 5 hours of recording for the recorded cells (A7). Statistical significance is assessed using Two way ANOVA and reveals a significant effect of time and genotype (****, P<0.0001)]. Scale bar: 50 µm (A2). (B) tdTomato(+) cIN migrating in living cortical slices from Nkx2.1-Cre;Rosa26-tdTomato embryos were tracked manually using the MTrackJ plugin and their trajectories color-coded as shown in legend (B2) to characterize their preferred direction (tangential, oblique, radial, immobile) and cortical layer localization (VZ-SVZ, IZ, CP).The picture in B1 illustrates the z-projection of trajectories reconstructed in a control slice, superimposed to the last picture of the movie. Cortical interneurons migrating tangentially in the superficial migratory stream (MZ) could not be tracked because of high density. Graphs in B2 compare the percentage of each kind of trajectory recorded in the lateral cortex of control slices (control, see also Suppl. Movie S3), Kif7 -/- slices (Kif7 -/-, see also Suppl. Movie S4), control slices treated acutely with either murine SHH (SHH, see also Suppl. Movie S5) or cyclopamine (Cyclo, see also Suppl. Movie S6). Significance of the differences between the four distributions were assessed by a Chi-square test, X2 (24, n=1224), P=0.0004998, ***). All experimental conditions differed from the control (Fisher test, P =0.0004998, ***). Slices from three WT animals in control condition or treated with drugs and from three Kif7 -/- animals were analyzed; the number of analyzed cells is indicated above bars. Schemes in B3 summarize the main results observed in each experimental condition. Trajectories are represented with the same color code as in B1, and line thickness is proportional to the percentage of cells exhibiting each type of trajectory. Immobile cells are figured by a coil. Box and whisker plots indicate the mean speed (B4), the frequency of period without nuclear movment (number/h) (B5) and the mean duration of period without nuclear movement (hour) (B6) for cIN migrating tangentially in the deep stream (red box and whisker plots, left) or to the cortical plate (green box and whisker plots, right). Statistical significance assessed by Krustkal-Wallis tests in each cluster. ****, P<0.0001; ***, B5 left P=0.0005; **, B4 right P=0.0088, B5 right, P=0.0037, B6 left P=0.0034; *, B4 left P=0.0233, B5 right P=0.0470, B6 left P=0.0134, B6 right P=0.0394. Number of analyzed cells is indicated on plots. VZ, ventricular zone; SVZ, subventricular zone; IZ, intermediate zone; CP, cortical plate; MZ, marginal zone. Scale bar: 300 µm.

We then analyzed the dynamic behavior of Kif7 -/- cIN in the mutant cortex (Fig. 6B). In WT Nkx2.1-Cre;Rosa26-tdTomato transgenic slices, fluorescent cIN migrated tangentially from the PSB to the dorsal cortex in two main streams, the MZ and the lower IZ/SVZ (Fig. 6B1, Supplementary Movie S3). All along these pathways, cIN sporadically operated a tangential to oblique/radial migration switch to colonize the CP. High cell density in the MZ prevented cell monitoring and MZ cells were discarded from analyses. In the deep stream, a large proportion of imaged cells (59.3 %) either maintained tangentially oriented trajectories (28.7 %, red trajectories in Fig. 6B1,B3, left column of Fig. 6B2) or reoriented to the CP along oblique or radial trajectories (30.6 % green trajectories in Fig. 6B1,B3, left column in Fig. 6B2). Most of these cIN reached the CP surface and remained there. A small proportion of cIN moved from the deep stream to the ventricular side of the slice (11.7%, blue trajectories in Fig. 6B1,B3, left column in Fig. 6B2). A significant proportion (15.6 %) migrated radially over the cortex thickness, inverting their polarity in the MZ or at the ventricular surface (pink trajectories in Fig. 6B1,B3, left column in Fig. 6B2). Rare cIN in the VZ, deep stream and CP moved very short distances or did not move (3.6%, dark colors in the left column of Fig. 6B2). Finally, 8 % cIN (orange) moved backward to the ventral brain and 1.7 % (grey) showed no specific directionality or layer specificity and were classified as chaotic (Fig. 6B1,B3, left column in Fig. 6B2). The majority of cIN presented a characteristic saltatory behavior, alternating stops and fast moves 64. The dynamics of cells recorded in the deep stream and/or moving to the CP is shown in Fig. 6B4-6 (left column on histograms).

The trajectories followed by cIN in the lateral/latero-dorsal cortex of Kif7 -/- slices (where CxCl12 mRNA was detected as shown in Fig. 5C) dramatically differed from those described above (compare WT and Kif7 -/- columns in Fig. 6B2, schemes in Fig.6B3 and Supplementary Movies S3 and S4). The number of cIN migrating tangentially in the deep migratory stream and/or reorienting to the CP dropped drastically (17.8 % as compared to 59.3 % in WT slices) to the benefit of immobile cells (35.7 % as compared to 3.6 % in WT slices) especially in the SVZ/lower IZ. The proportions of cIN moving to the ventricle or to the CP were similar (10.2 % and 13.9 %, respectively, as compared to 11.7 % and 30.6 % in WT slices) suggesting that the radial asymmetry of cortical slices was lost for Kif7 -/- cIN. The same proportion of cIN migrated radially across the cortical layers as in WT slices (17.2 % versus 15.6 % in WT) but the proportion of cIN with chaotic trajectories strongly increased (7.8 % versus 1.7 % in WT slices). Another major change was a significant decrease of the migration speed (excluding immobile cIN) (Fig. 6B4) with less frequent but longer stops in Kif7 -/- than in WT slices (Fig. 6B5,B6), confirming the dynamic parameters measured on co-cultures.

Kif7 ablation has been shown to activate SHH transcriptional signaling by blocking the cleavage of GLI3 in GLI3-R repressor, or conversely to inhibit SHH transcriptional signaling by preventing the formation of GLI activators (GLI1/2)3. Our western blot analysis confirmed that the GLI3-R/GLI3-FL ratio dropped in the Kif7 -/- cortex according to the expected inhibition of the GLI3 processing, a result not observed in the MGE. Since Kif7 -/- cIN are born in the MGE and migrate in the cortex where SHH signaling is activated due to an abnormal GLI3 processing, we first examined whether the application of SHH on a WT forebrain slice could mimic the migratory defaults of Kif7 -/- cIN in the mutant cortical slices. SHH application on control slices minimaly affected the trajectories of cIN (Fig. 6B2 compare WT and SHH (WT) columns). An increased proportion of cIN migrated radially across the cortical thickness (29 % compared to 15.6 % in WT slices) at the expense of cIN migrating from the deep stream to the CP (20.8 % instead of 30.6 % in WT slices). Frequent radial movements directed to the VZ (Fig. 6B2,B3, Supplementary Movie S5) did not evoke the migratory defaults of Kif7 -/- cIN in the mutant slices. Moreover, the frequency of chaotic trajectories did not increase (grey, Fig. 6B2) and the main dynamic alterations, a decreased speed (Fig.6B4) associated with an increased duration of stops (Fig. 6B6), were milder than those observed in Kif7 -/- cIN. On the contrary, the application of the SHH pathway inhibitor cyclopamine on WT cortical slices altered cIN trajectories in a way that resembled alterations in Kif7-/- slices (compare WT and cyclo (WT) columns in Fig. 6B2 and scheme in Fig. 6B3). For example, the proportion of immobile cIN and of cIN moving short distance reached 33.7 % as compared to 3.6 % in WT slices and 35.7% in Kif7-/- slices (Fig. 6B2). Cyclopamine application mainly affected the ability of cIN to move tangentially in the deep stream and to switch to the CP (Fig. 6B2,B3, Supplementary Movie S6). Nevertheless, the migration speed defects of Kif7 -/- cIN and of cyclopamine treated cIN differed (Fig. 6B4).

In conclusion, Kif7 -/- cIN exhibited intrinsic migratory defaults resembling those of cIN with a primary cilium ablation and showed cortical trajectories defaults resembling those of cyclopamine treated WT cIN. Kif7 ablation in cIN had the same effect than SHH signaling inhibition on cIN migratory behavior.

The strong influence of cyclopamine on the migration of cIN in WT cortical slices led us to determine the distribution of the endogenous activator SHH in the developing cortex.

Distribution of Shh mRNA and SHH protein in the E14.5 cortex

Shh mRNA is strongly expressed in the mantle zone of the ganglionic eminences (MGE, CGE) and POA during the time cIN differenciate 24,52 but its expression level in cIN after they reach the cortex is controversial 29,31,65. We thus reexamined Shh mRNA expression in the forebrain and SHH protein distribution in cortical layers. In situ hybridization (ISH) experiments performed at E13 confirmed the strong expression of Shh mRNA in the ventral forebrain, especially in the SVZ and mantle zone of the MGE (Fig. 7A1) in agreement with the SHH-dependent Gli1 mRNA expression in a VZ subdomain of the MGE 66. In contrast, Shh transcripts were bearly detectable in the cortex. Using RNAscope, which has higher sensitivity and spatial resolution than standard ISH, we re-examined the expression pattern of Shh mRNA at E14.5. In the ventral forebrain, RNAscope Shh transcript detection showed similar expression pattern as ISH (Fig. 7A1,A3,B1,B2). In addition, Shh mRNA and Lhx-6 mRNA, a cIN marker, were expressed in overlapping domains in the mantle zone of the MGE (Fig. 7A2-4). In the cortex, radially-oriented cells in the VZ/SVZ and tangentially oriented cells in the deepmigration stream expressed Shh mRNA at a very low level (one or two dots versus large cluster in the MGE) (Fig. 7C). Among the large number of cells expressing Lhx-6 transcripts (Fig. 7C1,C2), only 5% co-expressed Shh mRNA (Fig. 7C3).

Expression of Shh transcripts (A-C) and distribution of SHH protein (D) in the E13.5-E14.5 forebrain.

(A,B) Distribution in median (A) and caudal (B) coronal sections of Shh mRNA detected by in situ hybridization with an antisens Shh probe at E13.5 (A1,B1) and by RNAscope at E14.5 (A2-3,B2). Shh transcripts are strongly expressed in the medial ventral forebrain (SVZ and mantle zone of the MGE and septum, A1, A2), in the mantle zone of the CGE (B1,B2), in the zona limitans intrathalamica (ZLI in B1) and in the ventral midline of the 3rd ventricle (V3 in B1). RNAscope further confirmed the strong expression of Shh mRNA (A3, green) in MGE and septum regions that strongly express Lhx-6 mRNA (A2, red). Confocal observations in the SVZ and mantle zone of the MGE showed that Shh mRNA (green) is co-expressed with the Lhx6 mRNA (red) in a significant number of cIN (yellow cells in A4). (C) Confocal analyses at higher magnification of the double detection by RNAscope of Shh and Lhx-6 mRNA. Cells were identified on stacked images (Δz=1 µm) using Nomarski optic. In the lateral cortex close to the PSB (C1, z projection of 10 confocal planes) and in the dorsal cortex (C2, z projection of 10 confocal planes), a very small proportion of cells expressing Lhx-6 mRNA also express Shh mRNA (white arrows in C1,C2). Counting in the deep stream (SVZ-IZ) and in the MZ is shown in graph C3 (9-17 fields in three sections). A few progenitors in the cortical VZ express Shh mRNA at very low level (arrowhead, C1). (D) SHH-Nter and TBR2 co-immunostaining of Nkx2.1-Cre/R26R- tdTomato brain coronal sections at E14.5. Representative confocal merged stacked images (Δz=0.2 µm; 48 images) in the pallium-subpallium boundary (PSB, D1), lateral cortex (D2) and dorsal (D3) cortex revealing SHH-Nter immunostaining in blood vessels and the presence of numerous bright dots all over the cortical neuropile. On the ventricular side of the PSB and of the lateral-most part of the LGE, SHH-Nter(+) brigth elements are aligned radially. In the lateral cortical neuropile, smaller bright dots align radially in the VZ, tangentially in the SVZ- IZ, and radially in the CP. Cx, cortex; LGE, MGE and CGE, lateral, medial and caudal ganglionic eminence; CP, cortical plate; Hyp, hypothalamus; PSB, pallium-subpallium boundary; Sp, septum; Th, thalamus; V3, third ventricle; VZ, ventricular zone; SVZ, subventricular zone; IZ, intermediate zone; MZ, marginal zone. Scale bars: 500 µm (A, B), 20 µm (C), 100 µm (D).

Antibodies that recognize the N-ter part of SHH [SHH-Nter, e.g. “activated” SHH 67] strongly labeled in C57/Bl6 mouse brain the same regions known to strongly express SHH mRNA 65,68,69, such as the lateral wall of the third ventricle (Fig. S4A1) and the zona limitans intrathalamica at E12.5 (Fig. S4A2) and the choroid plexus and septum at E14.5 (Fig. S4B). SHH-Nter immunostaining was detected in the cortex at E12.5 and E14.5 (Fig. S4C,4D). To further investigate the expression of SHH in the cortical wall at E14.5, we performed immunostaining experiments in Nkx2.1-Cre/R26R-tdTomato embryos (Fig. 7D). SHH-Nter antibodies detected large bright dots that distributed radially away from the ventricle in the lateral LGE and at the ventricular angle (green dots in Fig. 7D1, upper panel). In the cortex, SHH-Nter antibodies immunostained bright tiny dots that show remarkable distribution (Fig. 7D2,D3). In the lateral (Fig. 7D2) and dorsal (Fig. 7D3) cortex, the bright SHH-ter immunopositive dots were aligned radially in the VZ and in the CP according to the radial arrangement of cortical progenitors and CP cells. In contrast, dots were aligned tangentially in the SVZ/IZ where cIN distribute among tangentially oriented growing axons. The intensity and density of positive dots appeared lower in the dorsal than in the lateral cortex suggesting a latero-dorsal gradient of distribution (compare Fig. 7D2 and 7D3). Some CP cells presented a diffuse cytoplasmic staining, which was not observed in the bottom layers. Together, our ISH and immunostaining experiments showed that the cortical SHH protein has probably an extrinsic origin as previously described at the early stage of cerebellum development 70.

Discussion

Our aim in the present paper was to better understand the developmental origin of functional abnormalities (e.g. epilepsy, cognitive deficits, …) in cortical circuits of human patients with KIF7 mutations by characterizing the developmental abnormalities of the cerebral cortex in a Kif7 -/- mouse model. We show that Kif7 ablation affects the development of the two populations of cortical neurons, cortical plate neurons and GABAergic interneurons, in specific and distinct ways, accordingly with their dorsal and ventral origin in the telencephalon. The alterations of the CP development (lack of most subplate cells in the dorsal cortex, and delayed corticofugal projection) are remiscient of defaults previously described in GLI3 null mutants that lack most subplate neurons and pioneer cortical projections 71. The abnormal distribution of Kif7 -/- cIN in the developing cortex partly reflects the structural and functional alterations of cortical layers on which cIN migrate (e.g. abnormal CXCL12 expression in the SVZ). However, we identify additional migration defaults of Kif7 -/- cIN that recall those of cIN lacking a primary cilium or of wild type cIN treated with the SHH pathway inhibitor cyclopamine, suggesting that the SHH pathway activity is inhibited in Kif7 -/- cIN. Kif7 ablation thus alters differently the SHH pathway activity in CP and cIN subpopulations, in agreement with their dorsal or ventral origin.

Kif7 knock-out mice as a model to study the effect of KIF7 mutations in human

KIF7 is a ciliary kinesin whose mutations in patients and ablation in mice alter the length rather than the structure of primary cilia. Primary cilium length was increased in patient fibroblasts 10, in MEF isolated from Kif7 -/- and in the neural tube of Kif7 -/- embryos at E10.5 723. Owing that KIF7 regulates the trafficking and processing of GLI factors, the cilium-dependent SHH signaling was altered in both patient fibroblasts 10 and in Kif7 -/- embryos 3,73 with GLI3-R decrease and GLI1/GLI2 up-regulation. These common features observed in cells from patients and in Kif7 -/- mice models, as well as a previous study of the corpus callosum agenesis in Kif7 mutant mice, which helped to dissect the mechanism at work in patients 44 allowed to propose that the Kif7 -/- mouse is a good model to study the alterations that lead to clinical features in patients. Most Kif7 -/- mice died at the end of gestation. However, the brain developed up to birth allowing to study its development.

All patients carrying mutations in the KIF7 gene on both alleles have developmental delay/intellectual disability (ID) and some have seizures. Investigations on cerebral malformation by MRI have shown that some of them have alterations in the cerebral cortex including cortical atrophy 9,12,13 and pachygyria 16. Since detecting defects in the cerebral cortex is more demanding in MRI quality images than corpus callosum agenesis or molar tooth sign, we can speculate that the number of patients with cerebral cortex defects is underscored. Malformations of cortical development was recognized as causes of ID 74 and epilepsy 75 and the cellular defects underlying these pathologies could imply defects in cell proliferation, neuronal migration and differenciation that lead to improper excitatory/inhibitory balance, connectivity into the cortex or of the cortex with other brain structures. The mechanism of action of KIF7 to transduce SHH ciliary dependent signaling has been debated. It has been accepted that SHH activates KIF7 binding to microtubules and its accumulation to the tip of the cilium, making the kinesin motor domain the most essential domain for SHH transduction. A more recent study proposed that KIF7 is an immobile kinesin translocated to the cilium tip after SHH binding by a complex KIF3A/Kif3B/KAP 76 suggesting that domains in the KIF7 protein essential to its function are more complex. This could explain why any mutation, even point mutation in the Coiled-coil domain or in C- terminal part of the KIF7 protein induces clinical feature as severe as when mutation led to a highly truncated protein. Another explanation is that any mutation causes unproper folding of the protein probably inducing its degradation by the proteasome.

KIF7 loss of function alters the cerebral cortex organization and the connectivity between the cortex and the thalamus

Our study demonstrates that in WT embryos, the levels of GLI3-FL and GLI3-R were much higher in the developing cortex than in the MGE and the cleavage of GLI3-FL in GLI3-R was more important in the cortex. Our observations are coherent with previous studies that identified a preponderant role of GLI3-R signaling to pattern the dorsal telencephalon and to regulate the proliferation and differenciation of neurons in the cortex 41,42. Moreover, the strong decrease of the cleavage of GLI3-FL in GLI3-R that we observed only in the cortex of Kif7 -/- embryos suggests that the hyper SHH signaling resulting from this decreased clivage affects the cortex, but not the MGE. The abnormal structure of the dorsal telencephalon in Kif7 -/- embryos (enlarged lateral ventricles, cortical wall folding and thinning) resembled the abnormal structure observed in ciliary mutants 54,77 and in GLI3 mutants 78–80. As in Kif7-/- embryos, a mild PBS disorganization and an abnormal differentiation of the caudal telencephalic vesicle and telo-diencephalic boundary that is dependent on SHH and GLI3 for its formation 81 were also reported in these mutants 77,78. The cellular disorganization observed in the developing cortex of Kif7 -/- embryos also recall defects of Gli3 mutants, especially the impaired preplate and SP neurons differentiation, the abnormal cortical layer formation and abnormal axonal growth 78–80. Several of these traits have been observed in the cortex of the ciliary mutant Rfx3 -/- that exhibits also a decreased GLI3-R/GLI3-FL ratio in the dorsal telencephalon 54. Interestingly, the knock down of KIF7 in E14.5 cortical cells altered several developmental stages of cortical neurons, leading to a decrease of apical radial progenitors, basal IP divisions, and to impaired neuronal polarity and migration 45. Sporadic heterotopia as we observed in the cortex of Kif7 -/- embryos, have been reported in one patient 16. Cortical heterotopia is not a landmark of ciliopathy but has been also described at E11 in the forebrain of the Ift88 hypomorphic mutant cobblestone (cbs, 77. And cellular alterations that should contribute to the formation of heterotopia, e.g. defective capacity to polarize and to migrate to the cortical plate, were described in post-mitotic neurons electroporated with shRNA libraries targeting ciliary proteins such as KIF7, BBS1, BBS10, NPHP8, ALMS1 45. Heterotopias were also detected in a Joubert syndrom patient carrying a mutation in the CELSR2 gene coding for a planar cell polarity protein 82 essential for neural progenitor cell fate decision, neural migration, axon guidance and neural maturation 83,84.

Beside delayed growth of cortical and thalamic axons, Kif7 -/- embryos exhibited connectivity defects between the thalamus and cortex, with abnormal trajectories of axon bundles in the basal forebrain, at the telo-diencephalic junction and along the PSB. To our knowledge, the ectopic projection of cortical axons to the pre-optic area had not been observed in GLI3 mutants that exhibited instead an ectopic thalamic projection to the amygdala 78,79. In the ciliary mutants Rfx3 -/-, ectopic cortical and thalamic axons were indeed observed in the basal telencephalon 54. Cortical neurons electroporated in utero with shRNAs to deplete KIF7 or ciliary proteins belonging to the BBSome, extended axons that showed abnormal fasculation and trajectories 45, demonstrating a cell-autonomous role of ciliary genes to control the axonal pathfinding. Finally, a large population of thalamic axons that had reached the cortex of Kif7 - /- embryos showed a massive elimination at E18.5 in the caudal telencephalon. This massive loss paralleled the loss of CP cells, as previously observed in Rfx3 -/- and lnpp5e -/- mutants54. Although disorganized thalamo-cortical connectivity had never been reported in patients with KIF7 mutation or in patients with ciliopathy, thalamocortical connectivity is essential for cognitive performances in young infants 85 and its alteration leads to neurodevelopemental delay and later cognitive deficits 86. Moreover, a link beween ciliary function, cortex development and connectivity with the thalamus had been previously identified in mutant mice lacking the planar cell polarity proteins CELSR2 and FZD3. These mice display ciliary defects and an abnormal development of tracts between the thalamus and the cortex leading to an hypotrophy of the intermediate zone where thalamo-cortical and corticofugal axons navigate 87,88.

Extra-cortical origin of SHH

Beside its major function in forebrain patterning 20,40, SHH regulates the embryonic and perinatal cerebral cortex development 27,28. The origin of the morphogen in this brain structure remains obscure. In agreement with a recent set of data from Moreau et al. 89, we showed that cIN born in the MGE or POA no longer express SHH after they reach the developing cortex. In addition, the level of local Shh mRNA synthesis was extremely low in the cortex at E14.5. In contrast, SHH protein was detected in immunoposive dots distributed throughout the cortex. Dots aligned radially in the VZ and CP, and tangentially in the IZ. This cortical distribution is compatible with an extrinsic origin of SHH and a transventricular delivery as described in the cerebellar ventricular zone 70. SHH is indeed present in the cerebrospinal fluid and likely secreted by the choroid plexus, since Shh mRNA is expressed in the choroid plexus of mice from E12 to E14.5 90. Accordingly, we showed here that the choroid plexus of the lateral ventricles was strongly immunopositive for SHH at E14.5. Interestingly, the CDO/BOC co-receptor - that favors SHH transport - was also detected along radially oriented structures by immunostaining in the mouse embryonic cortex (not illustrated). SHH could thus be transported away from the ventricular zone and/or meninges in vesicular structures able to diffuse accross the cortical thickness. Accordingly, it has been shown that SHH transport needs ESCRT-III member protein CHMP1A 15 and SHH related proteins are found in exosomes secreted from Caenorhabditis elegans epiderman cells 72, Drosophila wing imaginal discs 88, chick notochord cells and human cell lines 91. How SHH is transported throughout the cortex should be further explored. It has already be shown that SHH is present in multivesicular bodies, in neurites and filipodia in the postnatal hippocampus and cerebellum, and in vesicles located inside and outside of hippocampal neurons in culture 92.

Kif7 deletion leads to defects in cortical interneuron migration

The defects induced by gene mutations responsible for primary cilium dysfunction, including Kif7 deletion, have been examined in principal neurons born in the dorsal telencephalon, which organize the cortical layers 93,94. However, the cIN are born ventrally in the MGE where the expression levels of GLI3-R and GLI3-FL are extremely low in both WT and Kif 7-/- embryos. cIN leave the MGE and then migrate to the cortex during the embryonic development. Their primary cilium is functional and can transduce local SHH signals to influence their directionality, especially the tangential to radial reorientation of their trajectories toward the cortical plate 29. In the present study, we showed that the abnormal distribution of cIN in the mutant cortex of Kif7 -/- embryos resulted from cell autonomous defects and from extrinsic cortical defects, especially the transient lack of CXCL12 expression in the dorsal cortex at E14.5. Interestingly, Kif7 ablation affected the migratory behavior of cIN cultured on a WT cortical substrate in a similar way as did the ablation of Kif3a 29, another ciliary gene required for the transduction of SHH signals 95. Accordingly, cyclopamine application on WT cortical slices altered the migratory behavior of cIN in a way that mimicked the migratory behavior of Kif7 -/- cIN. We thus concluded from these observations that Kif7 ablation inhibited the Shh pathway activity in cIN. By consequence, Kif7 -/- cIN, which normally differentiate in a ventral telencephalic domain expressing high SHH level, present dynamic and trajectories alterations that mimics those of cIN lacking SHH signalling. Therefore Kif7 ablation in MGE-derived cIN does not activate SHH signalling by decreasing the GLI3-R/GLI3-FL ratio, as observed in the dorsal telencephalon. On the contrary, Kif7 ablation in MGE-derived cIN likely inhibits SHH signalling by a mechanism independent of GLI3 processing that was unchanged in the MGE of Kif7 -/-embryos (Fig.2). In addition to this cell autonomous defect that impaired cIN migration, we show that the lack of CXCL12 expression in the dorsal cortex transiently and locally modified the directionality of Kif7-/- cIN altering thereby the dorsal cortex colonization by cIN in Kif7 mutants 59,96.

Interestingly, acute SHH application on WT cortical slices slightly increased the percentage of radially migrating cIN. In contrast, cyclopamine that inhibited the SHH pathway blocked the migration of a large population of cIN in the deep cortical layers demonstrating that endogeneous cortical SHH is likely to activate the SHH pathway in migrating cIN to control their trajectories.

In conclusion, using the Kif7 -/- murine model, we show that the Kif7 deletion is responsible for a large range of developmental defects in the cerebral cortex, which include: i) alteration of the inhibitory interneurons migration by a combination of intrinsic and extrinsic factors, ii) abnormal formation of cortical layers by the cortical plate neurons and iii) major pathfinding defects in axonal projections between the cortex and the thalamus. Patients carrying mutations in the KIF7 gene are classified as ciliopathic patients and display developmental delay, ID and epilepsy. These clinical features could thus be caused by an abnormal distribution of excitatory and inhibitory cortical neurons leading to impaired excitatory/inhibitory balance, and by the development of abnormal connections in the cerebral cortex and with others brain structures. By demonstrating focal heterotopia and abnormal thalamo-cortical connectivity in the murine model, we hope to open a new field of clinical investigations using MRI and tractography to identify such defects in patients carrying mutations in the KIF7 gene.

Material and methods

Mice

E14.5 embryos of the swiss (for in situ hybridization and RNAscope) and C57/Bl6J (for sonic hedgehog immunostaining) strains were obtained from pregnant females purchased at JanvierLabs (France). Male and female wild type (WT) and Kif7 -/- animals (E12.5, E14.5, E16.5, E18.5 and P0) expressing tdTomato or not in the MGE-derived cIN (Nkx2.1- Cre;Rosa26-tdTomato) were generated in our animal facility by mating male Kif7 +/-; Nkx2.1-Cre;Rosa26-tdTomato and female Kif7 +/- mice to produce embryos and P0. Thirty four pregnant females were used. Kif7 +/- mice are a generous gift of Pr Chi-Chung Hui (the Provider, SickKids Hospital, Toronto, USA) and Prof Bénédicte Durand (the Transferor, CNRS, Lyon, France). Mice were genotyped as described previously 3. The day of the vaginal plug was noted E0.5. Experiments have been validated and approved by the Ethical committee Charles Darwin (C2EA-05, authorized project 02241.02) and mice were housed and mated according to European guidelines. Both male and female animals were used in this study. Sex is not considered as a biological variable in embryos.

Brain lysate preparation and immunoblotting

Cortex and MGE from E14.5 WT and Kif7 -/- embryos were dissected and frozen at −80°C until use. Samples were sonicated in 80µl Laemmli buffer and heated for 10 min at 37°C after adding 2.5% β-mercapto ethanol. Proteins were loaded on SDS-PAGE 7% gels (BIO-RAD) and transferred to 0.45 μm Nitrocellulose membranes (BIO-RAD). Membranes were cut to detect signals from proteins above and below 55 KDa and were blocked with 50 g/l non-fat dry milk in PBS/0.1% Tween 20 for 1h at RT, incubated with primary antibodies in the same solution O/N at 4°C, 1 h at RT with appropriate IRDye-conjugated secondary antibodies, and imaged and quantified using ChemiDocMP Imaging System (BIO-RAD). Commercial antibodies anti GLI3 (R&D Sytem-Biotechne, goat, #AF369, 1:200), anti-actin (Millipore, mice, #MAB1501, 1:10.000), Donkey-anti-goat-IR800 (Advansta, R-05781, 1:20.000), Goat-anti-mouse-StarBight Blue 700 (BIO-RA, 12004159, 1:2.500). Band intensities were quantified using gel analysis of Fiji software.

Immunohistochemistry and imaging

Heads of embryos and P0 animals were fixed by immersion in 0.1% picric acid /4% paraformaldehyde (PAF) in phosphate buffer (PB) for 4 hours and then in 4% PAF in PB overnight. After PBS washes, brains were dissected, included in 3% type VII agarose (Sigma, A0701) and sectionned at 70 μm with a vibratome. Immunostaining was performed on free-floating sections as explain previously 29. Primary antibodies were goat anti SHH-Nter (1:100, R&D system AF464), goat anti Netrin G1a (NG1a) (1:100, R&D system AF1166), rabbit anti TBR1 (1:1000, Abcam ab31940), rabbit anti TBR2 (1:1000, Abcam ab23345), rabbit anti PAX6 (1:100, clone poly19013, Covance PRB-278P), rabbit anti GSH2 (1:2000, Millipore ABN162), chicken anti MAP2 (1:500, Novus, NB30213), mouse anti SMI-32 (1:500, Covance SMI-32R), and rat CTIP2 (1:1000, Abcam ab18465). Primary antibodies were incubated in PGT (PBS/gelatine 2g per L/TritonX- 100 0.25%) with 0.1% glycine and 0.1% lysine for anti Netrin G1a antibodies and 3% BSA for anti SHH-Nter antibodies. Primary antibodies were revealed by immunofluorescence with the appropriate Alexa dye (Molecular Probes) or Cy3 (Jackson laboratories) conjugated secondary antibodies diluted in PGT (1:400). Bisbenzimide (1/5000 in PBT, Sigma) was used for nuclear counterstaining. Sections were mounted in mowiol/DABCO (25mg/mL) and imaged on a macroscope (MVX10 olympus) or an epifluorescence microscope (LEICA DM6000) using X40 and X63 immersion objectives, or a confocal microscope (Leica TCS SP5) using a x40 objective. Double immunostaining were performed when possible and images were representative of what was observed in more than 4 animals per genotype except for heterotopia. Macroscop images were treated to remove background around sections using wand tool in Adobe Photoshop 7.0 software. On epifluorescence and confocal microscop images, acquisition parameters and levels of intensity in Adobe Photoshop 7.0 software were similar for WT and Kif7 -/-. Merged images were acquired on the same section except for TBR1 with TBR2 acquired on adjacent sections. The fluorescence intensity along the depth of the cortex was assessed using the plot profile function of ImageJ under a 500 pixels wide line starting in the ventricle to the surface of the brain.

In situ hybridization

Specific antisense RNA probe for Shh, Gli1 and CxCl12 genes (gift of Marie-Catherine Tiveron) were used for in situ hybridization analyses. DIG-probes were synthesized with a labeling kit according to the manufacturer’s instructions (Roche, France). In situ hybridization was performed according to Tiveron et al 97 with few modifications (see Supplementary material). Sections were mounted on glass slides, dried, dehydrated in graded ethanol solutions, cleared in xylene and coverslipped with Eukitt.

RNAscope

Embryonic E14.5 heads were cuted in cold Leibovitz medium (Invitrogen) and immersion fixed in cold 4% (w/v) paraformaldehyde (PFA) in 0.12 m phosphate buffer, pH 7.4, overnight. Brains were then cryoprotected in PFA 4% /sucrose 10 %, embedded in gelatin 7.5% /sucrose 10% at 4°C and frozen. Biological samples were kept at −80°C until coronally sectioned at 20 μm with cryostat. RNAscope experiment was performed according manufacturer’s instructions (RNAScope® Multiplex Fluorescent V2 Assay, ACDbio) after a pre-treatment to remove gelatin. Two different probes/channels were used (C1 for Shh, C2 for Lhx-6). Probes were diluted according manufacturer’s instructions (1 Vol of C2 for 50 vol of C1) and dilution of the fluorophores (Opal 520 and 570) was 1:1500. Sections were mounted in mowiol/DABCO (25mg/mL) and imaged on a confocal microscop (Leica TCS SP5) using an immersion X10 or X63 objectives on 10 µm stacks of images (1 image /µm).

DiI experiments

Tiny crystals of DiI (1,1′-dioctadecyl-3,3,3′3′-tetramethylindocarbocyanine perchlorate) were inserted in the dorsal or lateral cortex of WT and Kif7 -/- embryonic brains fixed at E14.5 by immersion in 4% PAF. Injected brains were stored in 4% PAF at room temperature in the dark for several weeks to allow DiI diffusion. Before sectioning, brains were included in 3% agar. Coronal sections 60 μm thick were prepared with a vibratome and collected individually in 4% PAF. Sections were imaged with a macroscope (MVX10 Olympus).

Whole brain clearing and imaging

E14 and E16 embryos (WT, n=5; Kif7 -/-, n=6) were collected in cold L15 medium, perfused transcardially with 4% PAF using a binocular microscope. Brains were dissected and postfixed 3 days in 4% PAF at 4°C, and stored in PB at 4°C until clearing. All buffer solutions were supplemented with 0.01% Sodium Azide (Sigma-Aldrich). Whole brain immunostaining and clearing was performed at RT under gentle shaking according to a modified version of the original protocol 98. Briefly, perfused brains were dehydrated in graded methanol solutions (Fischer Scientific) (20, 50, 80% in ddH2O, 2×100%, 1h30 each), bleached overnight in 3% hydrogen peroxide (Sigma-Aldrich) in 100% methanol and rehydrated progressively. After PBS washes, brains were blocked and permeabilized 2 days in PBGT (0.2% gelatin (Merck), 0.5% triton X-100 (Sigma-Aldrich) in PBS). Brains were incubated 3 days at 37°C in primary goat anti Netrin G1a (1:100, R&D system AF1166) and rabbit anti TBR1 (1:1000, Abcam ab31940) antibodies. After 1 day of PBGT washes, brains were incubated 1 day at 37°C in secondary 647 donkey anti-goat (1:1000, Molecular probes,) and cy3 donkey anti-rabbit (1:1000, Jackson laboratories) antibodies. Immunolabeled brains were washed 1 day in PBS, embedded into 1.5% low-melting agarose (type VII, Sigma) in 1% ultra-pure Tris-acetate-EDTA solution, placed in 5-ml tubes (Eppendorf, 0030119452) and dehydrated 1h in each methanol baths (50%, 80% and 2×100%). Samples were then incubated for 3h in 33% methanol / 66% dichloromethane (DCM, Sigma-Aldrich), washed in 100% DCM (2x30 min) and incubated overnight (without shaking) in dibenzyl ether (DBE). Brains were stored in DBE at RT. Cleared samples were imaged on a light-sheet microscope (LaVision Biotec, x6.3 zoom magnification) equipped with a sCMOS camera (Andor Neo) and Imspector Microscope controller software. Imaris (Bitplane) was used for 3D reconstructions, snapshots and movies.

Brain slices

Brains of E14.5 WT and Kif7 -/- transgenic Nkx2.1-cre, tdTomato embryos were dissected in cold L15 medium, embedded in 3% type VII agarose and sectioned coronally with a manual slicer as explained in 29. Forebrain slices 250 µm thick were transferred in Millicell chambers and cultured for 4 hours in F12/DMEM medium supplemented with CFS 10% in a CO2 incubator (Merck Millipore) prior to pharmacological treatment and recording. Slices were incubated in drug for 2 hours before transfer in culture boxes equipped with a bottom glass coverslip for time-lapse imaging.

Pharmacological treatment

Either recombinant Mouse SHH (C25II), N-terminus (464-SH-025, R&D systems) at 0.5µg/mL final or cyclopamine (C4116, Sigma) at 2.5 µM final were added to the culture medium of slices and renewed after 12 hours. Recombinant SHH was reconstituted at 100 µg/mL in sterile PBS with BSA 0.1%. The stock solution of cyclopamine was 10 mM in DMSO and the culture medium of control experiments contained 1/4000 DMSO (vehicle). Control experiments in culture medium with and without vehicle did not differ and were analyzed together.

Co-cultures

Co-cultures were performed on polylysine/laminin-coated glass coverslips fixed to the bottom of perforated Petri dishes in order to image migrating MGE cells. Brains were collected in cold PBS at embryonic day E14.5. Cortices and MGE explants were then dissected in cold Leibovitz medium (Invitrogen). Cortices were mechanically dissociated. Dissociated cortical cells were cultured on the coated glass coverslips and left in the incubator (37L°C, 5% CO2) for 1 hours. A wound divided the substrate in two-halves on which WT and Kif7 -/- MGE explants from littermate embryos were placed (Fig. 6A). Co-cultures were cultured for 24h and imaged.

Videomicroscopy and image processing

Slices and co-cultures were imaged on an inverted microscope (Leica DMI4000) equipped with a spinning disk (Roper Scientific, USA) and a temperature-controlled chamber. Multi-position acquisition was performed with a Coolsnap HQ camera (Roper Scientific, USA) to allow the recording of the whole cortex. Images were acquired with a X20 objective (LX20, Fluotar, Leica, Germany) and a 561 nm laser (MAG Biosystems, Arizona). Z-stacks of 30 μm were acquired 50 µm away from the slice surface, with a step size of 2 μm and a time interval of 2 or 5 minutes for at least 21 hours. Acquisitions were controlled using the Metamorph software (Roper Scientific, USA). Cell trajectories were reconstructed on movies by tracking manually the cell rear with MTrakJ (ImageJ plugin, NIH, USA) and were clustered according to their position in cortical layers and to their orientation (see details in Fig. 6). The migration speed of cells, the frequency and duration of pauses were extracted from tracking data using excel macros. The directionality of cells was analysed along time using a macro from 99.

Study design

WT group was compared to Kif7 -/- group and to pharmacological treatment groups for brain slice experiments. The experimental unit was a single animal for immunohistochemistry analysis and individual cell for videomicroscopy analysis.

Statistical analysis

All data were obtained from at least three independent experiments and are presented as mean ± SEM (standard error of mean). Statistical analyses were performed with the GraphPad Prism software or R. Statistical significance of the data was evaluated using the unpaired two-tailed t test, the Mann–Whitney test, the Chi2 test or the Two-way ANOVA test followed by a post-hoc test. Data distribution was tested for normality using the D’Agostino and Pearson omnibus normality test. Values of P<0.05 were considered significant. In figures, levels of significance were expressed by * for P<0.05, ** for P<0.01, *** for P<0.001 and **** for P<0.0001.

eLife Assessment

This important study provides convincing evidence that the Kinesin protein family member KIF7 regulates the development of the cerebral cortex and its connectivity and the specificity of Sonic Hedgehog signaling by controlling the details of Gli repressor vs activator functions. This study provides important new insights into general aspects of cortical development.

Reviewer #1 (Public review):

Summary:

This is an interesting follow-up to a paper published in Human Molecular Genetics reporting novel roles in corticogenesis of the Kif7 motor protein that can regulate the activator as well as the repressor functions of the Gli transcription factors in Shh signalling. This new work investigates how a null mutation in the Kif7 gene affects the formation of corticofugal and thalamocortical axon tracts and the migration of cortical interneurons. It demonstrates that Kif7 null mutant embryos present with ventriculomegaly and heterotopias as observed in patients carrying KIF7 mutations. The Kif7 mutation also disrupts the connectivity between cortex and thalamus and leads to an abnormal projection of thalamocortical axons. Moreover, cortical interneurons show migratory defects that are mirrored in cortical slices treated with the Shh inhibitor cyclopamine suggesting that the Kif7 mutation results in a down-regulation of Shh signalling. Interestingly, these defects are much less severe at later stages of corticogenesis.

Strengths/weaknesses:

The findings of this manuscript are clearly presented and are based on detailed analyses. Using a compelling set of experiments, especially the live imaging to monitor interneuron migration, the authors convincingly investigate Kif7's roles and their results support their major claims. The migratory defects in interneurons and the potential role of Shh signalling present novel findings and provide some mechanistic insights but rescue experiments would further support Kif7's role in interneuron migration. Similarly, the mechanism underlying the misprojection which has previously been reported in other cilia mutants remains unexplored. Taken together, this manuscript makes novel contributions to our understanding of the role of primary cilia in forebrain development and to the aetiology of the neural symptons in ciliopathy patients.

Comments on revisions:

The authors addressed most of the points I raised in my original review. However, I am not convinced by the figures the authors present on Shh protein expression. The "bright tiny dots" of Shh protein in the cortex are not visible on the images in Figure 7. I wonder whether the authors could present higher magnification and/or black and white images with increased contrast.

Reviewer #2 (Public review):

Summary:

This study investigates the role of KIF7, a ciliary kinesin involved in the Sonic Hedgehog (SHH) signaling pathway, in cortical development using Kif7 knockout mice. The researchers examined embryonic cortex development (mainly at E14.5), focusing on structural changes and neuronal migration abnormalities.

Strengths:

(1) The phenotype observed is interesting, and the findings provide neurodevelopmental insight into some of the symptoms and malformations seen in patients with KIF7 mutations.

(2) The authors assess several features of cortical development, including structural changes in layers of the developing cortex, connectivity of the cortex with thalamus, as well as migration of cINs from CGE and MGE to cortex.

Weaknesses:

(1) The Kif7 null does have phenotype differences from individual mutations seen in patients. It would be interesting to add more thoughts about how the null differs from these mutants in ciliary structure and SHH signaling via the cilium.

(2) The description of altered cortex development at E14.5 is perhaps rather descriptive. It would be useful to assess more closely the changes occurring in different cell types and stages. For this it seems very important to have a time course of cortical development and how the structural organization changes over time. This would be easy to assess with the addition of serial sections from the same mice. It might also be interesting to see how SHH signaling is altered in different cortical cell types over time with a SHH signaling reporter mouse.

(3) Abnormal neurodevelopmental phenotypes have been widely reported in the absence of other key genes affecting primary cilia function (Willaredt et al., J Neurosci 2008; Guo et al., Nat Commun 2015). It would be interesting to have more discussion of how the Kif7 null phenotype compares to some of these other mutants.

(4) The authors see alterations in cIN migration to the cortex and observe distinct differences in the pattern of expression of Cxcl12 as well as suggest cell intrinsic differences within cIN in their ability to migrate. The slice culture experiments though make it a little difficult to interpret the cell intrinsic effects on cIN of loss of Kif7, as the differences in Cxcl12 patterns still exist presumably in the slice cultures. It would be useful to assess their motility in an assay where they were isolated, as well as assess transcriptional changes in cINs in vivo lacking KIF7 for expression patterns that may affect motility or other aspects of migration.

Comments on revisions:

The authors have made significant and thoughtful responses as well as experimental additions to the authors comments. Their efforts are appreciated and the manuscript is much improved.

Author response:

The following is the authors’ response to the previous reviews.

Reviewer #1 (Recommendations for the authors):

(1) I am not convinced by the figures the authors present on Shh protein expression. The "bright tiny dots" of Shh protein in the cortex are not visible on the images in Figure 7. I wonder whether the authors could present higher magnification and/or black and white images with increased contrast.

We have modified Figure 7: we now present a higher magnification and a black and white image with increased contrast to better visualize SHH (+) bright tiny dots in the lateral cortex.

(2)The manuscript also contains several typos.

We apologize for these mistakes which have all been corrected.

Version 3

Abstract

Abstract

Mutations of KIF7, a key ciliary component of Sonic hedgehog (SHH) pathway, are associated in humans with cerebral cortex malformations and clinical features suggestive of cortical dysfunction. KIF7 regulates the processing of GLI-A and GLI3-R transcription factors in a SHH-dependent manner both in humans and mice. Here, we examine the embryonic cortex development of a mouse model that lacks the expression of KIF7 (Kif7 -/-). The cortex is composed of principal neurons generated locally in the dorsal telencephalon where SHH expression is low and inhibitory interneurons (cIN) generated in the ventral telencephalon where SHH expression is high. We observe a strong impact of Kif7 deletion on the dorsal cortex development whose abnormalities resemble those of GLI3-R mutants: subplate cells are absent, the intermediate progenitor layer and cortical plate do not segregate properly, and corticofugal axons do not develop timely, leading to a delayed colonization of the telencephalon by thalamo-cortical axons. These structural defects alter the cortical distribution of cIN, which moreover exhibit intrinsic migration defects and cortical trajectories resembling those of cyclopamine-treated cIN. Our results show that Kif7 deletion impairs the cortex development in multiple ways, exhibiting opposite effects on SHH pathway activity in the developing principal neurons and inhibitory interneurons.

Introduction

The primary cilium is a tiny microtubule-based organelle present on the surface of nearly all mammalian cell types including neurons, which functions as a signaling hub and transduces several signaling pathways comprising Sonic Hedghog (SHH), Wnt, Delta/Notch and mTOR pathways 1. Primary cilium dysfunction causes pleiotropic diseases named ciliopathies. KIF7 is a ciliary kinesin responsible for the trafficking and the positive and negative regulation of the GLI transcription factors in the primary cilium of mammals 2. Loss of function of KIF7 in the mouse has shown that KIF7 regulates SHH signalling by acting downstream of Smoothened (Smo) and upstream of GLI2 and GLI3 3–7. Studies in mice established that KIF7 activity is dependent on the expression level of SHH. In the absence of SHH, KIF7 localizes to the base of the primary cilium 5. GLI factors are phosphorylated and addressed to the proteasome at the base of the primary cilium for degradation, leading to the formation of a cleaved and stable transcriptional repressor (GLI3-R). In the presence of SHH, KIF7 accumulates at the distal tip of the primary cilium 4,5 and associates with full length GLI2/3 that become transcriptional activators (GLI-A) 8.

Since 2011, ten studies have identified patients carrying mutations in the KIF7 gene responsible of ciliopathies classified according to clinical features as hydrolethalus, acrocallosal, Joubert and Greig cephalopolysyndactyly syndromes 9–18. MRI investigations revealed macrocephaly, ventricles enlargement and corpus callosum alterations. They also showed the characteristic hindbrain abnormalities observed in most ciliopathies such as molar tooth sign (MTS) and cerebellar atrophy. Patients also presented with mild but frequent cortical malformations as well as neurodevelopmental delay, intellectual disability and seizures 11,12,15,19 which indicate cortical abnormalities.

In physiological conditions, the proper activity of the cortex relies on the excitatory/inhibitory balance i.e. on the ratio, positioning and connectivity of excitatory glutamatergic neurons (principal neurons, PN) and inhibitory GABAergic interneurons (cIN), generated from progenitors in the dorsal and ventral telencephalon, respectively. The cortical neurons of dorsal and ventral origin each migrate towards the developing cortex, the so-called cortical plate (CP), according to a specific timeline allowing the establishment of proper connections in the CP. SHH plays a central role in the forebrain patterning and differentiation. At early embryonic stage, the ventral expression of SHH orchestrates the ventro-dorsal 20,21 and medio-lateral 22 regionalization of the mouse forebrain and the differentiation of ventral cell types 20. Shh ablation performed after forebrain patterning alters the specification of distinct subgroups of

GABAergic interneurons 23–26. Interestingly, conditional ablation of Shh and Smo in the embryonic cortex reduces the proliferation of dorsal progenitors 27, demonstrating a minimal SHH expression in the dorsal telencephalon, even before birth 28. Beside its role on proliferation and specification, SHH also controls the migration of interneurons to the cortical plate 29–33. In the embryonic telencephalon, GLI transcription factors mediate SHH signals in complex and specific ways. The three GLI factors identified in mammals are expressed in the mouse forebrain: GLI1 along the source of SHH in the ganglionic sulcus, and GLI2 and GLI3 dorsally to the SHH source 26,34,35. GLI1 acts uniquely as a pathway activator whereas GLI2 and GLI3 can be processed in transcriptional activators or inhibitors in the primary cilium. However, both GLI1 and GLI2 function primarily as transcriptional activators in response to SHH activity in the ventral forebrain 35. Nevertheless, GLI1 and GLI2 mutants show mild phenotypes 36,37 and GLI2 seems required to transduce high level SHH signals in mice 38. In contrast, the development of the dorsal cortex, where principal excitatory neurons differentiate, depends mainly on the expression of the GLI3-R repressor, in agreement with the low cortical expression of SHH 39,40. Following patterning, Gli3R/A ratio remains critical for specifying the fate of cortical progenitors and regulating cell cycle kinetics 41–43.

Previous studies investigated the mechanisms underlying the corpus callosum agenesis in patients with KIF7 mutation using Kif7 knock-out (Kif7 -/-) mice 44 and the consequence of KIF7 knock down on cortical neurogenesis in principal cells electroporated with Kif7 shRNA 45. It remains that the influence of developmental abnormalities associated with ciliopathies on the cortical cytoarchitecture is poorly understood. Given the crucial role of KIF7 in regulating the GLI pathways and the important role of Gli3-R and GLI2/3-A in regulating the early developmental stages of the dorsal and ventral telencephalon, respectively, we examined here the cortical development from E12.5, the establishment of long distance projections with the thalamus and the migration of GABAergic interneurons (cIN) in Kif7 -/- mice. The migratory behavior of Kif7 -/- cINs was investigated by time-lapse videomicroscopy in co-cultures and in organotypic cortical slices. We also compared the migration of cINs in Kif7 -/- cortical slices and in control slices treated with pharmacological activator and inhibitor of the SHH pathway. We moreover determined the local distribution of the SHH protein in the embryonic cortex. Our results show developmental defects leading to permanently displaced neurons, abnormal formation of cortical layers and defective cortical circuits that could be responsible for epilepsy and/or intellectual deficiency in patients carrying KIF7 mutation 9,11–18,44.

Results

Kif7 knock-out mice as a model to investigate the structural cortical defects that could lead to clinical feature in patients carrying KIF7 mutation

The KIF7 gene is located on chromosome 15 in human and encodes a 1343 aa protein containing, a kinesin motor domain and a GLI-binding domain in the N-terminal part followed by a Coiled-coil region and a Cargo domain able to bind a diverse set of cargos in the C-terminal part 46. Table 1 summarizes the clinical features associated with mutations targeting either the kinesin or GLI binding domain, the Coiled-coil domain, or the N-terminal Cargo domain. Interestingly, various mutations are associated with the same clinical picture, suggesting that KIF7 mutations, whatever their nature, could lead to protein loss of function, for example by altering the protein structure and solubility as proposed by Klejnot et Kozielski 47. All patients carrying mutation in the KIF7 gene have developmental delay (DD) and intellectual deficit (ID) associated with classical defects of ciliopathies (ventricle enlargement, macrocephaly, corpus callosum agenesis and MTS). Some patients have additional anatomical cerebral cortex defects such as poor frontal development, atrophy, pachygyria and heterotopia 9,10,13,16, which could participate not only to DD and ID but moreover to seizure as observed in a quarter of patients. In this context, we used a murine model in which the Kif7 gene had been deleted (Kif7 -/-) 3 to investigate the consequence of KIF7 loss of function on the cortex development. Kif7 -/- mice have been previously characterized as dying at birth with severe malformations, skeletal abnormalities (digits and ribs), neural tube patterning defects including exencephaly in a third of the mutants, microphtalmia, lack of olfactory bulbs and CC agenesis 3,4,44. In the present study, the KIF7 loss of function was analyzed in a mouse strain with a mixed genetic background in which exencephaly was observed at the same frequency in Kif7 -/- and control embryos. As previously reported, Kif7 -/- embryos displayed microphtalmia [illustrated at embryonic stage 14.5 (E14.5), Fig. 1A, black arrow] and polydactyly (not illustrated). Interestingly, skin laxity previously reported in ciliopathic patients 48,49 was observed in all mutants (Fig. 1A, white arrow). Comparison of brains from Kif7 -/- embryos and control littermates revealed the absence of olfactory bulbs in Kif7 -/- brains (illustrated at E14.5, Fig. 1B, white arrows on ventral face view).

Kif7 deletion alters cortical development on latero-medial and rostro-caudal axes.

(A) Kif7 -/- embryos are microphtalmic (black arrow) and exhibit skin laxity (white arrow). (B) External examination of the brain reveals the lack of olfactory bulbs (white arrows, left panel) and the thinning of the dorsal telencephalon (black arrow, right panel). (C) DAPI staining of rostro-caudal series of coronal sections at E12.5 (C1), E14.5 (C2) and E16.5 (C3) illustrates the anatomical defects of Kif7 -/- embryonic brains quantified at E14.5 in C4. The ventricles of Kif7 -/- embryos are strongly enlarged (upper left graph; WT, n=4-6, Kif7 -/-, n=5- 6 depending on the rostro-caudal level), their cortical thickness strongly decreased (upper right graph; (WT, n=3; Kif7 -/-, n= 4), resulting in minimal brain width (lower left graph; WT, n=5- 6; n=5-6 for Kif7 -/- depending on the rostro-caudal level) and height (lower right graph; WT, n=5-6; Kif7 -/-, n=4-6 depending on the rostro-caudal level) changes. Statistical significance was tested by Two-way ANOVA or mixed model (GraphPad 8.1.0). For ventricle surface and cortex thickness, mix model reveals a genotype effect (p<0.0001). For brain width and height, no genotype effect was observed, but a significant effect of the rostro-caudal level on brain width (P=0.0441) and height (P=0.0092). (D) The pallium-subpallium boundary identified by the limit of expression of ventral (GSH2, D1, D2) and dorsal (TBR2, D3, D4) telencephalic markers is less precisely defined and slightly shifted to the ventricular angle in E13-E14 Kif7 - /- embryos compared to wild type embryos (epifluorescent low magnification, D1, D3; confocal high magnification images, D2, D4). Graphs in C4 represent the means and S.E.M. Cx, cortex; Hip, hippocampus; LGE, lateral ganglionic eminence; MGE, median ganglionic eminence; CGE, caudal ganglionic eminence; Th, thalamus.. Scale bars, 500 µm (C,D1,D3), 150 µm (D2,D4).

Kif7 invalidation alters the development of the cortex

The thinning of the dorsal cortex at E14.5 allowing the lateral ventricles to be seen through (Fig. 1B, black arrow on dorsal face view) proned us to investigate the structural organization of the forebrain over time (Fig. 1C1-3, E12.5-E16.5). Analyses performed on rostro-caudal series of frontal sections identified strong abnormalities in the cortex development. E12.5 Kif7 -/- embryos exhibited cortical wall folding and thinning in rostro-median sections (Fig. 1C1, arrowheads) and prominent diencephalon (Fig. 1C1, stars). At E14.5, cortex folding was not longer observed in Kif7 -/- embryos which exhibited enlarged lateral ventricles (compare left and right panels in Fig. 1C2). The cortex remained thinned in the dorsal region and the latero- dorsal extension of the telencephalic vesicle was strongly reduced caudally. These defects were maintained in E16.5 embryos (Fig. 1C3) and the hippocampus did not develop properly (Fig. 1C3, caudal section). A quantitative analysis performed in E14.5 brains (Fig. 1C4) confirmed the ventriculomegaly (upper left graph) and the cortex thinning (upper right graph) in the rostral and median sections despite unchanged total width and height of the telencephalon (Fig. 1C4, lowers graphs). KIF7 depletion in E10.5-E11.5 mouse embryos was reported to alter the dorso-ventral patterning of the medial forebrain 44. The frontier of expression of GSH2 (Fig. 1D1,D2) and TBR2 (Fig. 1D3,D4), two transcription factors expressed in the ventral and dorsal telencephalon respectively, was slightly abnormal at E13-E14 in Kif7 -/- embryos with a small shift of GSH2(+) cells toward the ventricular angle (Fig. 1D2) and a dispersion of some TBR2(+) cells at the pallium-subpallium boundary (PSB) (Fig. 1D4). This suggests a near-normal dorso-ventral patterning of the telencephalic vesicles, however with a discrete blurring of the dorsal and ventral markers at the PSB.

In Kif7 -/- embryos, the cleavage of full length GLI3 (GLI3-FL) into the transcriptional repressor GLI3 (GLI3-R) is decreased 3,4. Here, we analyzed the expression of GLI3-FL and GLI3-R in the cortex and in the medial ganglionic eminence (MGE) where cortical interneurons (cIN) are born. We showed minimal expression of GLI3-R and GLI3-FL in the MGE compared to the cortex (Fig. 2A) and a strong decrease in the cleavage of GLI3-FL in GLI3-R only in the cortex of Kif7 -/- embryos (Fig. 2A,B).

Western blot analysis on the cortex and MGE of wild type (WT) and Kif7 -/- embryos at E14.5.

(A) The cleaved form of GLI3 (GLI3-R at 83 KDa) and the full length GLI3 (GLI3-FL at 190 KDa) are more abundant in the cortex compared to the MGE (see actin band intensity for protein loading). (B) The ratio Gli3-R/Gli3-FL is lower in the MGE than in the cortex of WT animals and is significantly decreased only in the cortex of Kif7 -/- brains compared to control. Two way ANOVA reveals significant interaction between brain structure and genotype (WT, n=4; Kif7 -/-, n=4; P=0.002) and multiple comparisons show a statistical difference between genotype only in the cortex (****, P<0.0001)]. Graph represents the means and S.E.M.

The structural defects described above convinced us to examine the cellular organization of the cortex of Kif7 -/- embryos. At E12.5, the whole cortex is proliferative. From stage E14.5, the mouse cortex can be described as a stack of three specialized domains: i) the upper/superficial cortical layers containing post-mitotic cells generated in the proliferative zones of the cortex, with TBR1(+) post-mitotic neurons located in the cortical plate (CP) in-between the subplate (SP) and marginal zone (MZ) MAP2-expressing cells; ii) the deep proliferative layers, ventricular and subventricular zones (VZ and SVZ, respectively), located along the lateral ventricle; and iii) the intermediate zone (IZ) located between the CP and the proliferative layers, which hosts the radially and tangentially migrating neurons and growing axons. At the end of the embryonic development, the proliferative layers are strongly reduced due to neurogenesis decrease, whereas the cortical thickness had increased due to CP delamination. At E14.5, the TBR1(+) CP cells appeared more clustered in Kif7 -/- embryos than in wild type (WT) embryos where they formed a rather regular, radially organized CP (compare left and right panels in Fig. 3A1,A2). The thickness of the upper MAP2(+) MZ layer was extremly irregular in Kif7 -/- embryos and MAP2(+) SP cells were missing in the dorsal cortex (Fig. 3A1,A2, white arrows indicate the limit of distribution of MAP2(+) SP cells). In E16.5 and E18.5 embryos, the CP was obviously thinner in the mutant as compared to the WT (Fig. 3B,C), even though CTIP(+) cells differentiated normally in V-VI layers (Fig. 3C). The CP thinning appeared even more pronounced in the caudal than in the median and rostral cortex. We then analyzed the proliferative layers focusing on TBR2(+) intermediate progenitors (IP) which normally form a dense layer in the cortical SVZ (Fig. 3D1, left). The TBR2(+) layer constantly showed an abnormal positioning in the dorsal cortex of Kif7 -/- embryos. TBR2(+) cells indeed reached the brain surface (Fig. 3D1, arrow) where they mixed up with post-mitotic TBR1(+) cells instead of keeping a minimal distance with TBR1 cells as observed in control brains (Fig. 3D2, arrow). As a consequence intermediate progenitors and post-mitotic neurons no longer segregated in the dorsal cortex of Kif7 -/- embryos where no IZ was observed (Fig. 3D2, arrow). Accordingly to the thinning of the cortex starting at E12.5, the TBR2(+) layer was thinner at all development stages analyzed (Fig. 3E,F). In contrast, the abnormal positioning of TBR2(+) cells at the brain surface in the dorsal cortex of E14.5 embryos was no longer observed at E16.5 (Fig. 3F) suggesting either a transient structural disorganization and/or developmental delay of the dorsal cortex in Kif7-/- embryos at E14.5. The TBR2 staining moreover revealed focal heterotopia in the dorsal or lateral cortex of E14.5 Kif7 -/- embryos, either at the ventricular or pial surface (Fig. S1).

Histological alterations in the developing cortex of Kif7 -/- embryos.

(A-C) Immunostaining of cortical post-mitotic layers in E14.5 (A), E16.5 (B) and E18.5 (C) in median coronal sections representative of 5 WT and Kif7 -/- embryos imaged on epifluorescence (A1,B,C) or confocal (A2) microscopes. At E14.5 (A1,A2), the TBR1(+) staining (red) of the cortical plate is more clustered in Kif7 -/- than in WT embryos, and the MAP2(+) staining (green) of the subplate is absent in the dorsal cortex of E14.5 Kif7 -/- embryos (white arrow, right column). The post-mitotic layers remain thinner in Kif7 -/- embryos at later embryonic stages as illustrated by TBR1 staining at E16.5 (B) and CTIP1 staining at E18.5 (C) that specifically labels the deeper cortical layers (V-VI). Moreover, the hippocampus is underdeveloped in the mutant (white arrow) (C). (D-F) Immunostaining of TBR2(+) proliferative layer. At E14.5 (D1,D2), the TBR2(+) layer (green) of secondary progenitors appears disorganized in the lateral cortex of the Kif7 -/- embryos (white arrowhead in D1) and reaches the brain surface in the dorsal cortex of Kif7 -/- embryo where it intermingles with post-mitotic TBR1(+) cells (D2, red) (white arrows). In E12.5 Kif7 -/- embryo (E), the TBR2(+) layer is thin and clustered however its localization in the thickness of the cortex is normal. At E16.5 stage (F), the TBR2(+) layer is normally positionned in the deeper layer of the cortex in both WT and Kif7 -/- embryos. Hip, hippocampus ; V, ventricle; VZ, ventricular zone; SVZ, subventricular zone; IZ, intermediate zone; CP, cortical plate ; MZ, marginal zone. Scale bars: 250 µm (A1,B,F), 100 µm (A2), 200 µm (D,E); 500 µm (C).

The loss of Kif7 alters the connectivity between the cortex and the thalamus

The intermediate zone (IZ) hosts migrating neurons and the growing cortical and thalamic projections 50. After reaching the CP by radial migration, post-mitotic neurons extended pioneer axons oriented tangentially in the IZ and directed to the PSB 51,52. We thus examined whether the structural defects of the CP and SP and the lack of IZ in the dorsal cortex of Kif7 -/- embryos did associate with developmental abnormalities of corticofugal projections. We labeled corticofugal axons by inserting small crystals of DiI in the CP of E14.5 paraformaldehyde fixed brains. After DiI had diffused along corticofugal axons, we analyzed labeled axons on coronal sections. In both control and Kif7 -/- brains, corticofugal axons extended below the CP toward the PSB (Fig. 4A). According to the latero-medial gradient of cortical development, DiI injections in the dorsal cortex of control and Kif7 -/- brains (Fig. 4A1) labeled much less axons than injections in the lateral cortex (Fig. 4A2,A3). Remarkably, the cortical bundles labeled from similar cortical regions were always much smaller and shorter in Kif7 -/- than in WT brains (compare right and left columns in Fig. 4A). Dorsal injections labeled large bundles crossing the PSB in WT brains, but only a few axons reached the PSB in Kif7 -/- brains (Fig. 4A1). The growth of corticofugal axons appeared severely hampered in the dorsal cortex of Kif7 -/- embryos. Lateral injections in WT brains labeled anterogradely corticofugal axons that reached the embryonic striatum and continued medially in the internal capsule (IC), and labeled retrogradely thalamic projections (Fig. 4A3, left panel). Lateral injections in Kif7 -/- embryos labeled anterogradely cortical axons that spread in the striatum (Fig. 4A2, right panel) and labeled a large bundle directed to the ventral pre-optic area (Fig. 4A3, right panel, white arrow), which was never observed in WT brains. Moreover, no thalamic axons were retrogradely labeled from the lateral cortex in Kif7 -/- embryos.

Kif7 deletion disrupts of the connectivity between the cortex and the thalamus.

(A) Panels illustrate the corticofugal projections labeled by DiI crystal (red dots) positioned in the dorsal (A1) and lateral (A2,A3) cortex of E14.5 WT (left) and Kif7 -/- (right) embryos on vibratome sections performed 30 days after DiI placement and imaged on a macroscope. In Kif7 -/- embryos, less axons project from the dorsal (A1) and lateral (A2) cortex to the subpallium than in WT (compare enlarged views of the projections below A1 and A2). Cortical injections in Kif7 -/- embryos do not label thalamic axons (compare left and right panels in A3) but label a ventral projection (A3, white arrow). (B) Rostro-caudal series of coronal sections (B1-B5) immunostained with anti-Netrin G1a (NG1a) antibodies compare the trajectory of thalamo-cortical axons (TCA) in a E14.5 WT (left) and in a Kif7 -/- (right) embryo imaged on a macroscope. TCA reach the pallium-subpallium boundary of the WT embryo whereas they are lost in the ventral forebrain of the mutant (B4, white arrow). (C) Representative three-dimensional reconstructions (C1, lateral; C2, horizontal; C3, coronal views) of WT and Kif7 -/- E14.5 brains immunostained as a whole with NG1a (red) and TBR1 (green) antibodies that label respectively the TCA and cortical plate cells before transparization and imaging with a light sheet microscope (see Supplementary movies S1 and S2). While all labeled TCA extend in the internal capsule (IC) and a significant proportion of them enter the cerebral cortex in the WT brain, TCA in the Kif7 -/- brain split in two bundles in the basal telencephalon. A bundle stops shortly after entering the IC (white arrowheads) whereas the second bundle extends ventrally (white arrows). (D) At E16.5, TCA are immunostained with NG1a in WT (left) and Kif7 -/- (right) embryos in whole brain with TBR1 antibodies before transparization and imaging (D1) and on coronal section with TBR2 (D2). The TCA extend in the cortex in Kif7 -/- brain, however, fiber density is reduced in the median and caudal brain compared to WT (D1, white arrowhead). TCA invade the post mitotic layers in the cortex (above TBR2 layer) to a lower extend in Kif7 -/- brain (D2). In the Kif7 -/- brain, thick bundles of NG1a(+) fibers project ventrally from the caudal telencephalon, a projection never observed in control brains (D1, white arrow). (E) At E18.5, rostro-caudal series of coronal sections (E1-E4) immunostained with NG1a antibodies compare the trajectory of TCA in WT (left) and Kif7 -/- (right) embryos imaged on a macroscope. TCA reach the dorsal cortex in both WT and Kif7 -/- embryos. However, the TCA projection in Kif7 -/- (arrowheads) become thinner in median sections (E2) and almost disappers in caudal sections (E3,E4). Fiber trajectories in the lateral striatum are abnormal (E2,E3, arrows). (F) Panels illustrate the cortical projections immunolabeled by Smi32 antibodies in E18.5 WT and Kif7 -/- brain. In Kif7 -/- brain, the number of fibers connecting the cortex with other brain structures is strongly reduced (arrow). Cx, cortex; dTh, dorsal thalamus; Hip, hippocampus; Hy, hypothalamus; St, striatum; IC, internal capsule; LGE, lateral ganglionic eminence; MGE, median ganglionic eminence; PSB, pallium-subpallium boundary; .. Scale bars: 250 µm (A,B,C,D), 500 µm (E,F).

To identify the trajectories of thalamic axons in Kif7 -/- embryos, we immunostained coronal sections of E14.5 brains with antibodies against the Netrin G1a (NG1a), an early marker of thalamocortical axons (TCA, Fig. 4B) 53. At E14.5, NG1a antibodies labeled a larger population of neurons in the dorsal thalamus of Kif7 -/- than in WT brains (Fig. 4B, stars). In control brains, TCA made a right angle turn to join the IC in the striatum. Then they extended to the PSB and some axons entered the lateral cortex (Fig. 4B2,B3, left panel). In Kif7 -/- embryos, most thalamic axons formed a thick bundle oriented ventrally in the basal forebrain (Fig. 4B4, white arrow in right panel). A minor structural defect was evidenced at the telo-diencephalic junction by Pax6 staining in Kif7 -/- brains (Fig. S2), recalling the prethalamus defect described in the ciliary Rfx3 mutant 54. Given the complex trajectory of NG1a thalamic axons, we immunostained thalamic axons in whole brains and imaged them after transparisation using a light-sheet microscope. The three-dimensional reconstruction of labeled projections confirmed that thalamic axons made a sharp turn to reach the IC and then navigated straight to the PBS in WT brains (Fig. 4C, left panels and Movie S1). In Kif7 -/- embryos, most thalamic axons stopped their course after leaving the diencephalon, and formed two short bundles: a large one oriented to the IC (Fig. 4C, right panels, arrowheads and Movie S2) and another one oriented to the amygdala, more caudally and ventrally (Fig. 4C, white arrows). This last projection was never observed in WT brains. Similar analyses performed at E16.5 showed that two days later, most TCA had reached the cortex in both WT and Kif7 -/- embryos (Fig. 4D, arrowheads) despite a small delay in Kif7 -/- embryos and abnormal trajectories in the PSB region (Fig. 4D1,D2). The ectopic ventral caudal projection to the amygdala was still present at E16.5 in Kif7 -/- embryos (Fig. 4D1, arrow). At E18.5, the TCA projections in the Kif7 -/- embryos appeared almost normal in the rostral cortex (Fig. 4E1) but drastically reduced in the median and caudal cortex (Fig. 4E2,E3,E4, arrowheads). In addition, a thick bundle of thalamic axons followed an abnormal trajectory at the PSB (Fig. 4E2,E3, arrow). Neurofilament staining with Smi32 antibodies confirmed the strong reduction of projections in the cortex and IC of Kif7 -/- embryos (Fig. 4F).

Kif7 invalidation alters the cortical distribution of cIN at E14.5

Because cIN are born in the basal forebrain and likely depend on contact/functional interactions with pioneer corticofugal projections for the first stages of their migration 55,56, we examined their distribution in the developing cortex of Kif7 -/- animals. cIN are generated outside of the cortex, in the medial and caudal ganglonic eminences (MGE, CGE) and preoptic area (POA) of the ventral forebrain. They enter the lateral cortex at E12.5 - E13.5, depending on mouse strains, and colonize the whole cortex by organizing two main tangential migratory streams, a superficial one in the MZ, and a deep and large one in the lower TBR2(+) IZ/SVZ 57,58. We thus analyzed the cortical distribution of tdTomato(+) MGE-derived cIN in WT and Kif7 -/- Nkx2.1- Cre;Rosa26-tdTomato transgenic embryos. An abnormal MGE cell distribution was observed in the developing cortex of Kif7 -/- embryos from E14.5 (Fig. 5 and Fig. S3). In mutant brains, both the latero-medial extent of the superficial and deep tangential migratory streams and the space separating them were significantly shortened compared to WT brains (Fig. 5A2,B2,B3). The closeness of the two streams in mutants was consistent with the reduced thickness of the corticofugal and TCA projections that navigate tangentially in the IZ (illustrated in Fig. 4). The superficial migratory stream was thinner and denser in Kif7 -/- embryos compared to WT (Fig. 5A1,B1, right panels). Remarkably, cIN no longer colonized the dorsal cortex of Kif7 -/- embryos and the deep tangential stream stopped in the region where the layer of TBR2(+) SVZ cells switched to the cortical surface (Fig. 5A1, white arrow). The tangential progression of cIN in the developing cortex is controlled by CXCL12, a chemokine transiently expressed in TBR2(+) cells and in meninges 60,61. CXCL12 prevents the premature occurrence of the tangential to radial migration switch required for cIN to colonize the CP 59–62. In situ hybridization experiments revealed that CxCl12 mRNA was not expressed in the dorsal cortex of E14.5 Kif7 -/- embryos (Fig. 5C, black arrow). To characterize the consequences of the abnormal expression pattern of CXCL12 at E14.5 on the migratory behavior of cIN, we sliced coronally E14.5 Kif7 -/- and WT forebrains and imaged by time-lapse video-microscopy the trajectories of MGE-derived tdTomato-expressing cIN in the dorsal cortex of organotypic slices (Fig. 5D). While most cIN presented tangential and oblique trajectories in WT slices, a majority of cIN exhibited radially oriented trajectories in Kif7 -/- slices (Fig. 5D). CXCL12 expression in SVZ strongly decreases at E16.5 59,63 and analyses performed at E16.5 in fixed Kif 7 -/- and WT embryos showed that cIN had pursued their migration and reached the dorsal cortex (Fig. S3A) even though the tangential migration of cIN in Kif7 -/- embryos remained delayed as compared to WT embryos. Analyses performed in a surviving P0 animal (Fig. S3B,C) revealed an abnormal distribution of cIN in the CP characterized by a strong decrease of cIN density in the infragranular layers that were moreover much thinner than in control brains (see CTIP2 staining in Fig. S3C).

Abnormal cortical distribution of cIN and Cxcl12 transcript expression in E14.5 in Kif7 -/- brains.

(A) The cortical distribution of cIN is visualized in WT and Kif7 -/- mouse embryos crossed with the Nkx2.1-Cre/R26R-tdTomato strain in which MGE-derived cIN express the fluorescent marker tdTomato. Panels in A1 compare the distribution of tdTomato (+) cIN in WT and Kif7 -/- cortical sections prepared at the same rostro-caudal level and in which SVZ is immunostained with TBR2 antibodies (green). Pictures show that the deep migratory stream of cIN terminates in Kif7 -/- brains in the cortical region where the TBR2(+) layer reaches the cortical surface. Quantitative analysis of the mean length of the deep and superficial migratory streams measured from the entry in the pallium to the last detected cIN in the cortex is illustrated in graph A2. Statistical significance is assessed using Two way ANOVA [layers (WT, n=4; Kif7 -/-, n=3; ***, P=0.001) and genotype (*, P=0.0157)]. (B) Representative pictures (B1) of the deep and superficial tangential migratory streams of cIN in the lateral cortex of WT and Kif7 -/- embryos. Pictures illustrate the decreased thickness of the superficial stream, and the reduced distance between the deep-superficial streams in Kif7 -/- embryos. Graph in B2 (WT, n=4; Kif7 -/-, n=4) compares the distribution of the fluorescence intensity along a ventricle/MZ axis (see grey rectangles in B1) using the plot profile function of FIJI. Curves show no change in the distance between the ventricular wall and the deep cIN, but a significant reduction of the distance between the two streams in the Kif7 -/- cortical sections as quantified on the graph B3 [WT, n=4; Kif7 -/-, n=4; Two way ANOVA reveals a significant interaction between genotype and layer (P=0.0233) and multiple comparisons, a statistical difference between genotype only for the distance between de cIN streams (**, P=0.0051)]. (C) Panels compare the distribution of Cxcl12 mRNA in WT (left panel) and Kif7 -/- (right panel) forebrain coronal sections at E14.5. The WT section shows Cxcl12 transcript enrichment in a deep cortical layer already identified as the SVZ. In the Kif7 -/- cortical section, the expression of Cxcl12 transcripts is reduced to the lateral part of the SVZ. (D) Z-projections of 30 frames acquired during 12h in the dorsal cortex of living organoptypic forebrain slices representative of E14.5 WT and Kif7 -/- embryos with tdTomato expressing cIN. Kif7 -/- cIN were able to migrate dorsally but followed preferentially radially oriented trajectories Scale bars: 200 µm.

The abnormal distribution of cIN in the cortex of Kif7 -/- embryos reflected the structural and molecular abnormalities that we had identified in their developing cortex. We next examined if migratory defaults proper to Kif7 -/- cIN could moreover contribute to their abnormal cortical distribution.

Kif7 invalidation affects the migratory behavior of cIN in co-cultures and organotypic slices

We thus compared the migratory behavior of Kif7 -/- and WT cIN using an in vitro model previously established in the lab 64 to compare the dynamics of mutant and WT cIN migrating on a substrate of WT cortical cells (Fig. 6A). WT and Kif7 -/- E14.5 MGE explants from tdTomato expressing embryos were cultured on WT dissociated cortical cells (Fig. 6A1). Fluorescent migrating MGE cells were imaged using time-lapse video-microscopy and tracked over time (Fig. 6A2). Dynamic parameters (speed during moves, stops, Fig. 6A3-A6) and trajectories (Fig. 6A7) were analyzed on movies. Kif7 -/- cIN migrated slightly more rapidly than WT cIN (Fig. 6A3) mainly because the duration of their stops was significantly reduced (Fig. 6A4). The saltatory behavior of cIN was preserved (same frequency of fast nuclear movements, Fig. 6A5) despite a slight decrease of the speed of fast nuclear movements (Fig. 6A6). WT cIN migrated along pretty straight trajectories (Fig. 6A2, left panel and Fig. 6A7, black curve) whereas Kif7 -/- cIN showed reduced directionality persistence over time (Fig. 6A2, right panel and Fig. 6A7, blue curve). Kif7 ablation thus affected the migratory behavior of cIN in a cell-autonomous manner and the migratory defaults of Kif7 -/- cIN evoked those of Kif3a -/- cIN whose primary cilium is non-functional (See Figure S6 in 29).

Dynamic behavior of migrating cIN in co culture and organotypic cortical slices.

(A) To prepare co-cultures, MGE explants were dissected out of telencephalic vesicles at E14.5 from WT or Kif7 -/- Nkx2.1-Cre;Rosa26-tdTomato embryos and placed on a substrate of WT E14.5 dissociated cortical cells (A1). After 24h in culture, MGE cells migrated centrifugally away from MGE explants on the substrate of cortical cells and were recorded for 7 hours.WT (n=160) and Kif7 -/- (n=126) MGE cells were tracked manually using the MTrackJ plugin allowing to analyze migratory parameters (A2). Box and whisker plots indicate the speed (A3), the time without nuclear movement (A4), the frequency (A5) and speed of nuclear movements (A6). Statistical significance is assessed by Mann Whitney or Unpaired tests. ***, P<0.001 (A4); *, P=0.0308 (A3), P=0.0123 (A5). Directionality persistence was calculated as the ratio of the cell displacement (distance between the first and last positions of the cell) on the cell trajectory (A7, scheme on the right). Graph represented the mean directionality ratios at each time point over the first 5 hours of recording for the recorded cells (A7). Statistical significance is assessed using Two way ANOVA and reveals a significant effect of time and genotype (****, P<0.0001)]. Scale bar: 50 µm (A2). (B) tdTomato(+) cIN migrating in living cortical slices from Nkx2.1-Cre;Rosa26-tdTomato embryos were tracked manually using the MTrackJ plugin and their trajectories color-coded as shown in legend (B2) to characterize their preferred direction (tangential, oblique, radial, immobile) and cortical layer localization (VZ-SVZ, IZ, CP).The picture in B1 illustrates the z-projection of trajectories reconstructed in a control slice, superimposed to the last picture of the movie. Cortical interneurons migrating tangentially in the superficial migratory stream (MZ) could not be tracked because of high density. Graphs in B2 compare the percentage of each kind of trajectory recorded in the lateral cortex of control slices (control, see also Suppl. Movie S3), Kif7 -/- slices (Kif7 -/-, see also Suppl. Movie S4), control slices treated acutely with either murine SHH (SHH, see also Suppl. Movie S5) or cyclopamine (Cyclo, see also Suppl. Movie S6). Significance of the differences between the four distributions were assessed by a Chi-square test, X2 (24, n=1224), P=0.0004998, ***). All experimental conditions differed from the control (Fisher test, P =0.0004998, ***). Slices from three WT animals in control condition or treated with drugs and from three Kif7 -/- animals were analyzed; the number of analyzed cells is indicated above bars. Schemes in B3 summarize the main results observed in each experimental condition. Trajectories are represented with the same color code as in B1, and line thickness is proportional to the percentage of cells exhibiting each type of trajectory. Immobile cells are figured by a coil. Box and whisker plots indicate the mean speed (B4), the frequency of period without nuclear movment (number/h) (B5) and the mean duration of period without nuclear movement (hour) (B6) for cIN migrating tangentially in the deep stream (red box and whisker plots, left) or to the cortical plate (green box and whisker plots, right). Statistical significance assessed by Krustkal-Wallis tests in each cluster. ****, P<0.0001; ***, B5 left P=0.0005; **, B4 right P=0.0088, B5 right, P=0.0037, B6 left P=0.0034; *, B4 left P=0.0233, B5 right P=0.0470, B6 left P=0.0134, B6 right P=0.0394. Number of analyzed cells is indicated on plots. VZ, ventricular zone; SVZ, subventricular zone; IZ, intermediate zone; CP, cortical plate; MZ, marginal zone. Scale bar: 300 µm.

We then analyzed the dynamic behavior of Kif7 -/- cIN in the mutant cortex (Fig. 6B). In WT Nkx2.1-Cre;Rosa26-tdTomato transgenic slices, fluorescent cIN migrated tangentially from the PSB to the dorsal cortex in two main streams, the MZ and the lower IZ/SVZ (Fig. 6B1, Supplementary Movie S3). All along these pathways, cIN sporadically operated a tangential to oblique/radial migration switch to colonize the CP. High cell density in the MZ prevented cell monitoring and MZ cells were discarded from analyses. In the deep stream, a large proportion of imaged cells (59.3 %) either maintained tangentially oriented trajectories (28.7 %, red trajectories in Fig. 6B1,B3, left column of Fig. 6B2) or reoriented to the CP along oblique or radial trajectories (30.6 % green trajectories in Fig. 6B1,B3, left column in Fig. 6B2). Most of these cIN reached the CP surface and remained there. A small proportion of cIN moved from the deep stream to the ventricular side of the slice (11.7%, blue trajectories in Fig. 6B1,B3, left column in Fig. 6B2). A significant proportion (15.6 %) migrated radially over the cortex thickness, inverting their polarity in the MZ or at the ventricular surface (pink trajectories in Fig. 6B1,B3, left column in Fig. 6B2). Rare cIN in the VZ, deep stream and CP moved very short distances or did not move (3.6%, dark colors in the left column of Fig. 6B2). Finally, 8 % cIN (orange) moved backward to the ventral brain and 1.7 % (grey) showed no specific directionality or layer specificity and were classified as chaotic (Fig. 6B1,B3, left column in Fig. 6B2). The majority of cIN presented a characteristic saltatory behavior, alternating stops and fast moves 64. The dynamics of cells recorded in the deep stream and/or moving to the CP is shown in Fig. 6B4-6 (left column on histograms).

The trajectories followed by cIN in the lateral/latero-dorsal cortex of Kif7 -/- slices (where CxCl12 mRNA was detected as shown in Fig. 5C) dramatically differed from those described above (compare WT and Kif7 -/- columns in Fig. 6B2, schemes in Fig. 6B3 and Supplementary Movies S3 and S4). The number of cIN migrating tangentially in the deep migratory stream and/or reorienting to the CP dropped drastically (17.8 % as compared to 59.3 % in WT slices) to the benefit of immobile cells (35.7 % as compared to 3.6 % in WT slices) especially in the SVZ/lower IZ. The proportions of cIN moving to the ventricle or to the CP were similar (10.2

% and 13.9 %, respectively, as compared to 11.7 % and 30.6 % in WT slices) suggesting that the radial asymmetry of cortical slices was lost for Kif7 -/- cIN. The same proportion of cIN migrated radially across the cortical layers as in WT slices (17.2 % versus 15.6 % in WT) but the proportion of cIN with chaotic trajectories strongly increased (7.8 % versus 1.7 % in WT slices). Another major change was a significant decrease of the migration speed (excluding immobile cIN) (Fig. 6B4) with less frequent but longer stops in Kif7 -/- than in WT slices (Fig. 6B5,B6), confirming the dynamic parameters measured on co-cultures.

Kif7 ablation has been shown to activate SHH transcriptional signaling by blocking the cleavage of GLI3 in GLI3-R repressor, or conversely to inhibit SHH transcriptional signaling by preventing the formation of GLI activators (GLI1/2)3. Our western blot analysis confirmed that the GLI3-R/GLI3-FL ratio dropped in the Kif7 -/- cortex according to the expected inhibition of the GLI3 processing, a result not observed in the MGE. Since Kif7 -/- cIN are born in the MGE and migrate in the cortex where SHH signaling is activated due to an abnormal GLI3 processing, we first examined whether the application of SHH on a WT forebrain slice could mimic the migratory defaults of Kif7 -/- cIN in the mutant cortical slices. SHH application on control slices minimaly affected the trajectories of cIN (Fig. 6B2 compare WT and SHH (WT) columns). An increased proportion of cIN migrated radially across the cortical thickness (29 % compared to 15.6 % in WT slices) at the expense of cIN migrating from the deep stream to the CP (20.8 % instead of 30.6 % in WT slices). Frequent radial movements directed to the VZ (Fig. 6B2,B3, Supplementary Movie S5) did not evoke the migratory defaults of Kif7 -/- cIN in the mutant slices. Moreover, the frequency of chaotic trajectories did not increase (grey, Fig. 6B2) and the main dynamic alterations, a decreased speed (Fig. 6B4) associated with an increased duration of stops (Fig. 6B6), were milder than those observed in Kif7 -/- cIN. On the contrary, the application of the SHH pathway inhibitor cyclopamine on WT cortical slices altered cIN trajectories in a way that resembled alterations in Kif7-/- slices (compare WT and cyclo (WT) columns in Fig. 6B2 and scheme in Fig. 6B3). For example, the proportion of immobile cIN and of cIN moving short distance reached 33.7 % as compared to 3.6 % in WT slices and 35.7% in Kif7-/- slices (Fig. 6B2). Cyclopamine application mainly affected the ability of cIN to move tangentially in the deep stream and to switch to the CP (Fig. 6B2,B3, Supplementary Movie S6). Nevertheless, the migration speed defects of Kif7 -/- cIN and of cyclopamine treated cIN differed (Fig. 6B4).

In conclusion, Kif7 -/- cIN exhibited intrinsic migratory defaults resembling those of cIN with a primary cilium ablation and showed cortical trajectories defaults resembling those of cyclopamine treated WT cIN. Kif7 ablation in cIN had the same effect than SHH signaling inhibition on cIN migratory behavior.

The strong influence of cyclopamine on the migration of cIN in WT cortical slices led us to determine the distribution of the endogenous activator SHH in the developing cortex.

Distribution of Shh mRNA and SHH protein in the E14.5 cortex

Shh mRNA is strongly expressed in the mantle zone of the ganglionic eminences (MGE, CGE) and POA during the time cIN differenciate 24,52 but its expression level in cIN after they reach the cortex is controversial 29,31,65. We thus reexamined Shh mRNA expression in the forebrain and SHH protein distribution in cortical layers. In situ hybridization (ISH) experiments performed at E13 confirmed the strong expression of Shh mRNA in the ventral forebrain, especially in the SVZ and mantle zone of the MGE (Fig. 7A1) in agreement with the SHH- dependent Gli1 mRNA expression in a VZ subdomain of the MGE 66. In contrast, Shh transcripts were bearly detectable in the cortex. Using RNAscope, which has higher sensitivity and spatial resolution than standard ISH, we re-examined the expression pattern of Shh mRNA at E14.5. In the ventral forebrain, RNAscope Shh transcript detection showed similar expression pattern as ISH (Fig. 7A1,A3,B1,B2). In addition, Shh mRNA and Lhx-6 mRNA, a cIN marker, were expressed in overlapping domains in the mantle zone of the MGE (Fig. 7A2-4). In the cortex, radially-oriented cells in the VZ/SVZ and tangentially oriented cells in the deepmigration stream expressed Shh mRNA at a very low level (one or two dots versus large cluster in the MGE) (Fig. 7C). Among the large number of cells expressing Lhx-6 transcripts (Fig. 7C1,C2), only 5% co-expressed Shh mRNA (Fig. 7C3).

Expression of Shh transcripts (A-C) and distribution of SHH protein (D) in the E13.5-E14.5 forebrain.

(A,B) Distribution in median (A) and caudal (B) coronal sections of Shh mRNA detected by in situ hybridization with an antisens Shh probe at E13.5 (A1,B1) and by RNAscope at E14.5 (A2-3,B2). Shh transcripts are strongly expressed in the medial ventral forebrain (SVZ and mantle zone of the MGE and septum, A1, A2), in the mantle zone of the CGE (B1,B2), in the zona limitans intrathalamica (ZLI in B1) and in the ventral midline of the 3rd ventricle (V3 in B1). RNAscope further confirmed the strong expression of Shh mRNA (A3, green) in MGE and septum regions that strongly express Lhx-6 mRNA (A2, red). Confocal observations in the SVZ and mantle zone of the MGE showed that Shh mRNA (green) is co- expressed with the Lhx6 mRNA (red) in a significant number of cIN (yellow cells in A4). (C) Confocal analyses at higher magnification of the double detection by RNAscope of Shh and Lhx-6 mRNA. Cells were identified on stacked images (Δz=1 µm) using Nomarski optic. In the lateral cortex close to the PSB (C1, z projection of 10 confocal planes) and in the dorsal cortex (C2, z projection of 10 confocal planes), a very small proportion of cells expressing Lhx-6 mRNA also express Shh mRNA (white arrows in C1,C2). Counting in the deep stream (SVZ- IZ) and in the MZ is shown in graph C3 (9-17 fields in three sections). A few progenitors in the cortical VZ express Shh mRNA at very low level (arrowhead, C1). (D) SHH-Nter and TBR2 co-immunostaining of Nkx2.1-Cre/R26R-tdTomato brain coronal sections at E14.5. Representative confocal merged stacked images (Δz=0.2 µm; 48 images) in the pallium- subpallium boundary (PSB, D1), lateral cortex (D2) and dorsal (D3) cortex revealing SHH-Nter immunostaining in blood vessels and the presence of numerous bright dots all over the cortical neuropile. On the ventricular side of the PSB and of the lateral-most part of the LGE, SHH- Nter(+) brigth elements are aligned radially. In the lateral cortical neuropile, smaller bright dots align radially in the VZ, tangentially in the SVZ-IZ, and radially in the CP (see higher magnification on the right panel in which SHH-Nter immunostaining is shown in white and the contrast is increased). Cx, cortex; LGE, MGE and CGE, lateral, medial and caudal ganglionic eminence; CP, cortical plate; Hyp, hypothalamus; PSB, pallium-subpallium boundary; Sp, septum; Th, thalamus; V3, third ventricle; VZ, ventricular zone; SVZ, subventricular zone; IZ, intermediate zone; MZ, marginal zone. Scale bars: 500 µm (A, B), 20 µm (C), 100 µm (D).

Antibodies that recognize the N-ter part of SHH [SHH-Nter, e.g. “activated” SHH 67] strongly labeled in C57/Bl6 mouse brain the same regions known to strongly express SHH mRNA 65,68,69, such as the lateral wall of the third ventricle (Fig. S4A1C) and the zona limitans intrathalamica at E12.5 (Fig. S4A2) and the choroid plexus and septum at E14.5 (Fig. S4B). SHH-Nter immunostaining was detected in the cortex at E12.5 and E14.5 (Fig. S4C,D). To further investigate the expression of SHH in the cortical wall at E14.5, we performed immunostaining experiments in Nkx2.1-Cre/R26R-tdTomato embryos (Fig. 7D). SHH-Nter antibodies detected large bright dots that distributed radially away from the ventricle in the lateral LGE and at the ventricular angle (green dots in Fig. 7D1, upper panel). In the cortex, SHH-Nter antibodies immunostained bright tiny dots that show remarkable distribution (Fig. 7D2,D3). In the lateral (Fig. 7D2) and dorsal (Fig. 7D3) cortex, the bright SHH-ter immunopositive dots were aligned radially in the VZ and in the CP according to the radial arrangement of cortical progenitors and CP cells. In contrast, dots were aligned tangentially in the SVZ/IZ where cIN distribute among tangentially oriented growing axons. The intensity and density of positive dots appeared lower in the dorsal than in the lateral cortex suggesting a latero-dorsal gradient of distribution (compare Fig. 7D2,D3). Some CP cells presented a diffuse cytoplasmic staining, which was not observed in the bottom layers. Together, our ISH and immunostaining experiments showed that the cortical SHH protein has probably an extrinsic origin as previously described at the early stage of cerebellum development 70.

Discussion

Our aim in the present paper was to better understand the developmental origin of functional abnormalities (e.g. epilepsy, cognitive deficits, …) in cortical circuits of human patients with KIF7 mutations by characterizing the developmental abnormalities of the cerebral cortex in a Kif7 -/- mouse model. We show that Kif7 ablation affects the development of the two populations of cortical neurons, cortical plate neurons and GABAergic interneurons, in specific and distinct ways, accordingly with their dorsal and ventral origin in the telencephalon. The alterations of the CP development (lack of most subplate cells in the dorsal cortex, and delayed corticofugal projection) are remiscient of defaults previously described in GLI3 null mutants that lack most subplate neurons and pioneer cortical projections 71. The abnormal distribution of Kif7 -/- cIN in the developing cortex partly reflects the structural and functional alterations of cortical layers on which cIN migrate (e.g. abnormal CXCL12 expression in the SVZ). However, we identify additional migration defaults of Kif7 -/- cIN that recall those of cIN lacking a primary cilium or of wild type cIN treated with the SHH pathway inhibitor cyclopamine, suggesting that the SHH pathway activity is inhibited in Kif7 -/- cIN. Kif7 ablation thus alters differently the SHH pathway activity in CP and cIN subpopulations, in agreement with their dorsal or ventral origin.

Kif7 knock-out mice as a model to study the effect of KIF7 mutations in human

KIF7 is a ciliary kinesin whose mutations in patients and ablation in mice alter the length rather than the structure of primary cilia. Primary cilium length was increased in patient fibroblasts 10, in MEF isolated from Kif7 -/- and in the neural tube of Kif7 -/- embryos at E10.5 73. Owing that KIF7 regulates the trafficking and processing of GLI factors, the cilium-dependent SHH signaling was altered in both patient fibroblasts 10 and in Kif7 -/- embryos 3,73 with GLI3-R decrease and GLI1/GLI2 up-regulation. These common features observed in cells from patients and in Kif7 -/- mice models, as well as a previous study of the corpus callosum agenesis in Kif7 mutant mice, which helped to dissect the mechanism at work in patients 44 allowed to propose that the Kif7 -/- mouse is a good model to study the alterations that lead to clinical features in patients. Most Kif7 -/- mice died at the end of gestation. However, the brain developed up to birth allowing to study its development.

All patients carrying mutations in the KIF7 gene on both alleles have developmental delay/intellectual disability (ID) and some have seizures. Investigations on cerebral malformation by MRI have shown that some of them have alterations in the cerebral cortex including cortical atrophy 9,12,13 and pachygyria 16. Since detecting defects in the cerebral cortex is more demanding in MRI quality images than corpus callosum agenesis or molar tooth sign, we can speculate that the number of patients with cerebral cortex defects is underscored. Malformations of cortical development was recognized as causes of ID 74 and epilepsy 75 and the cellular defects underlying these pathologies could imply defects in cell proliferation, neuronal migration and differenciation that lead to improper excitatory/inhibitory balance, connectivity into the cortex or of the cortex with other brain structures. The mechanism of action of KIF7 to transduce SHH ciliary dependent signaling has been debated. It has been accepted that SHH activates KIF7 binding to microtubules and its accumulation to the tip of the cilium, making the kinesin motor domain the most essential domain for SHH transduction. A more recent study proposed that KIF7 is an immobile kinesin translocated to the cilium tip after SHH binding by a complex KIF3A/Kif3B/KAP 76 suggesting that domains in the KIF7 protein essential to its function are more complex. This could explain why any mutation, even point mutation in the Coiled-coil domain or in C-terminal part of the KIF7 protein induces clinical feature as severe as when mutation led to a highly truncated protein. Another explanation is that any mutation causes unproper folding of the protein probably inducing its degradation by the proteasome.

KIF7 loss of function alters the cerebral cortex organization and the connectivity between the cortex and the thalamus

Our study demonstrates that in WT embryos, the levels of GLI3-FL and GLI3-R were much higher in the developing cortex than in the MGE and the cleavage of GLI3-FL in GLI3-R was more important in the cortex. Our observations are coherent with previous studies that identified a preponderant role of GLI3-R signaling to pattern the dorsal telencephalon and to regulate the proliferation and differenciation of neurons in the cortex 41,42. Moreover, the strong decrease of the cleavage of GLI3-FL in GLI3-R that we observed only in the cortex of Kif7 -/- embryos suggests that the hyper SHH signaling resulting from this decreased cleavage affects the cortex, but not the MGE. The abnormal structure of the dorsal telencephalon in Kif7 -/- embryos (enlarged lateral ventricles, cortical wall folding and thinning) resembled the abnormal structure observed in ciliary mutants 54,77 and in GLI3 mutants 78–80. As in Kif7-/- embryos, a mild PBS disorganization and an abnormal differentiation of the caudal telencephalic vesicle and telo- diencephalic boundary that is dependent on SHH and GLI3 for its formation 81 were also reported in these mutants 77,78. The cellular disorganization observed in the developing cortex of Kif7 -/- embryos also recall defects of Gli3 mutants, especially the impaired preplate and SP neurons differentiation, the abnormal cortical layer formation and abnormal axonal growth 78–80. Several of these traits have been observed in the cortex of the ciliary mutant Rfx3 -/- that exhibits also a decreased GLI3-R/GLI3-FL ratio in the dorsal telencephalon 54. Interestingly, the knock down of KIF7 in E14.5 cortical cells altered several developmental stages of cortical neurons, leading to a decrease of apical radial progenitors, basal IP divisions, and to impaired neuronal polarity and migration 45. Sporadic heterotopia as we observed in the cortex of Kif7 -/- embryos, have been reported in one patient 16. Cortical heterotopia is not a landmark of ciliopathy but has been also described at E11 in the forebrain of the Ift88 hypomorphic mutant cobblestone (cbs, 77). And cellular alterations that should contribute to the formation of heterotopia, e.g. defective capacity to polarize and to migrate to the cortical plate, were described in post-mitotic neurons electroporated with shRNA libraries targeting ciliary proteins such as KIF7, BBS1, BBS10, NPHP8, ALMS1 45. Heterotopias were also detected in a Joubert syndrom patient carrying a mutation in the CELSR2 gene coding for a planar cell polarity protein 82 essential for neural progenitor cell fate decision, neural migration, axon guidance and neural maturation 83,84.

Beside delayed growth of cortical and thalamic axons, Kif7 -/- embryos exhibited connectivity defects between the thalamus and cortex, with abnormal trajectories of axon bundles in the basal forebrain, at the telo-diencephalic junction and along the PSB. To our knowledge, the ectopic projection of cortical axons to the pre-optic area had not been observed in GLI3 mutants that exhibited instead an ectopic thalamic projection to the amygdala 78,79. In the ciliary mutants Rfx3 -/-, ectopic cortical and thalamic axons were indeed observed in the basal telencephalon 54. Cortical neurons electroporated in utero with shRNAs to deplete KIF7 or ciliary proteins belonging to the BBSome, extended axons that showed abnormal fasculation and trajectories 45, demonstrating a cell-autonomous role of ciliary genes to control the axonal pathfinding. Finally, a large population of thalamic axons that had reached the cortex of Kif7 -/- embryos showed a massive elimination at E18.5 in the caudal telencephalon. This massive loss paralleled the loss of CP cells, as previously observed in Rfx3 -/- and lnpp5e -/- mutants 54. Although disorganized thalamo-cortical connectivity had never been reported in patients with KIF7 mutation or in patients with ciliopathy, thalamocortical connectivity is essential for cognitive performances in young infants 85 and its alteration leads to neurodevelopemental delay and later cognitive deficits 86. Moreover, a link beween ciliary function, cortex development and connectivity with the thalamus had been previously identified in mutant mice lacking the planar cell polarity proteins CELSR2 and FZD3. These mice display ciliary defects and an abnormal development of tracts between the thalamus and the cortex leading to an hypotrophy of the intermediate zone where thalamo-cortical and corticofugal axons navigate 87,88.

Extra-cortical origin of SHH

Beside its major function in forebrain patterning 20,40, SHH regulates the embryonic and perinatal cerebral cortex development 27,28. The origin of the morphogen in this brain structure remains obscure. In agreement with a recent set of data from Moreau et al. 89, we showed that cIN born in the MGE or POA no longer express SHH after they reach the developing cortex. In addition, the level of local Shh mRNA synthesis was extremely low in the cortex at E14.5. In contrast, SHH protein was detected in immunoposive dots distributed throughout the cortex. Dots aligned radially in the VZ and CP, and tangentially in the IZ. This cortical distribution is compatible with an extrinsic origin of SHH and a transventricular delivery as described in the cerebellar ventricular zone 70. SHH is indeed present in the cerebrospinal fluid and likely secreted by the choroid plexus, since Shh mRNA is expressed in the choroid plexus of mice from E12 to E14.5 90. Accordingly, we showed here that the choroid plexus of the lateral ventricles was strongly immunopositive for SHH at E14.5. Interestingly, the CDO/BOC co- receptor - that favors SHH transport - was also detected along radially oriented structures by immunostaining in the mouse embryonic cortex (not illustrated). SHH could thus be transported away from the ventricular zone and/or meninges in vesicular structures able to diffuse accross the cortical thickness. Accordingly, it has been shown that SHH transport needs ESCRT-III member protein CHMP1A 15 and SHH related proteins are found in exosomes secreted from Caenorhabditis elegans epiderman cells 72, Drosophila wing imaginal discs 88, chick notochord cells and human cell lines 91. How SHH is transported throughout the cortex should be further explored. It has already be shown that SHH is present in multivesicular bodies, in neurites and filipodia in the postnatal hippocampus and cerebellum, and in vesicles located inside and outside of hippocampal neurons in culture 92.

Kif7 deletion leads to defects in cortical interneuron migration

The defects induced by gene mutations responsible for primary cilium dysfunction, including Kif7 deletion, have been examined in principal neurons born in the dorsal telencephalon, which organize the cortical layers 93,94. However, the cIN are born ventrally in the MGE where the expression levels of GLI3-R and GLI3-FL are extremely low in both WT and Kif 7-/- embryos. cIN leave the MGE and then migrate to the cortex during the embryonic development. Their primary cilium is functional and can transduce local SHH signals to influence their directionality, especially the tangential to radial reorientation of their trajectories toward the cortical plate 29. In the present study, we showed that the abnormal distribution of cIN in the mutant cortex of Kif7 -/- embryos resulted from cell autonomous defects and from extrinsic cortical defects, especially the transient lack of CXCL12 expression in the dorsal cortex at E14.5. Interestingly, Kif7 ablation affected the migratory behavior of cIN cultured on a WT cortical substrate in a similar way as did the ablation of Kif3a 29, another ciliary gene required for the transduction of SHH signals 95. Accordingly, cyclopamine application on WT cortical slices altered the migratory behavior of cIN in a way that mimicked the migratory behavior of Kif7 -/- cIN. We thus concluded from these observations that Kif7 ablation inhibited the Shh pathway activity in cIN. By consequence, Kif7 -/- cIN, which normally differentiate in a ventral telencephalic domain expressing high SHH level, present dynamic and trajectories alterations that mimics those of cIN lacking SHH signalling. Therefore Kif7 ablation in MGE-derived cIN does not activate SHH signalling by decreasing the GLI3-R/GLI3-FL ratio, as observed in the dorsal telencephalon. On the contrary, Kif7 ablation in MGE-derived cIN likely inhibits SHH signalling by a mechanism independent of GLI3 processing that was unchanged in the MGE of Kif7 -/-embryos (Fig. 2). In addition to this cell autonomous defect that impaired cIN migration, we show that the lack of CXCL12 expression in the dorsal cortex transiently and locally modified the directionality of Kif7-/- cIN altering thereby the dorsal cortex colonization by cIN in Kif7 mutants 59,96.

Interestingly, acute SHH application on WT cortical slices slightly increased the percentage of radially migrating cIN. In contrast, cyclopamine that inhibited the SHH pathway blocked the migration of a large population of cIN in the deep cortical layers demonstrating that endogeneous cortical SHH is likely to activate the SHH pathway in migrating cIN to control their trajectories.

In conclusion, using the Kif7 -/- murine model, we show that the Kif7 deletion is responsible for a large range of developmental defects in the cerebral cortex, which include: i) alteration of the inhibitory interneurons migration by a combination of intrinsic and extrinsic factors, ii) abnormal formation of cortical layers by the cortical plate neurons and iii) major pathfinding defects in axonal projections between the cortex and the thalamus. Patients carrying mutations in the KIF7 gene are classified as ciliopathic patients and display developmental delay, ID and epilepsy. These clinical features could thus be caused by an abnormal distribution of excitatory and inhibitory cortical neurons leading to impaired excitatory/inhibitory balance, and by the development of abnormal connections in the cerebral cortex and with others brain structures. By demonstrating focal heterotopia and abnormal thalamo-cortical connectivity in the murine model, we hope to open a new field of clinical investigations using MRI and tractography to identify such defects in patients carrying mutations in the KIF7 gene.

Material and methods

Mice

E14.5 embryos of the swiss (for in situ hybridization and RNAscope) and C57/Bl6J (for sonic hedgehog immunostaining) strains were obtained from pregnant females purchased at JanvierLabs (France). Male and female wild type (WT) and Kif7 -/- animals (E12.5, E14.5, E16.5, E18.5 and P0) expressing tdTomato or not in the MGE-derived cIN (Nkx2.1- Cre;Rosa26-tdTomato) were generated in our animal facility by mating male Kif7 +/-; Nkx2.1- Cre;Rosa26-tdTomato and female Kif7 +/- mice to produce embryos and P0. Thirty four pregnant females were used. Kif7 +/- mice are a generous gift of Pr Chi-Chung Hui (the Provider, SickKids Hospital, Toronto, USA) and Prof Bénédicte Durand (the Transferor, CNRS, Lyon, France). Mice were genotyped as described previously 3. The day of the vaginal plug was noted E0.5. Experiments have been validated and approved by the Ethical committee Charles Darwin (C2EA-05, authorized project 02241.02) and mice were housed and mated according to European guidelines. Both male and female animals were used in this study. Sex is not considered as a biological variable in embryos.

Brain lysate preparation and immunoblotting

Cortex and MGE from E14.5 WT and Kif7 -/- embryos were dissected and frozen at -80°C until use. Samples were sonicated in 80µl Laemmli buffer and heated for 10 min at 37°C after adding 2.5% β-mercapto ethanol. Proteins were loaded on SDS-PAGE 7% gels (BIO-RAD) and transferred to 0.45 μm Nitrocellulose membranes (BIO-RAD). Membranes were cut to detect signals from proteins above and below 55 KDa and were blocked with 50 g/l non-fat dry milk in PBS/0.1% Tween 20 for 1h at RT, incubated with primary antibodies in the same solution O/N at 4°C, 1 h at RT with appropriate IRDye-conjugated secondary antibodies, and imaged and quantified using ChemiDocMP Imaging System (BIO-RAD). Commercial antibodies anti GLI3 (R&D Sytem-Biotechne, goat, #AF369, 1:200), anti-actin (Millipore, mice, #MAB1501, 1:10.000), Donkey-anti-goat-IR800 (Advansta, R-05781, 1:20.000), Goat-anti-mouse-StarBight Blue 700 (BIO-RA, 12004159, 1:2.500). Band intensities were quantified using gel analysis of Fiji software.

Immunohistochemistry and imaging

Heads of embryos and P0 animals were fixed by immersion in 0.1% picric acid /4% paraformaldehyde (PAF) in phosphate buffer (PB) for 4 hours and then in 4% PAF in PB overnight. After PBS washes, brains were dissected, included in 3% type VII agarose (Sigma, A0701) and sectionned at 70 μm with a vibratome. Immunostaining was performed on free-floating sections as explain previously 29 . Primary antibodies were goat anti SHH-Nter (1:100, R&D system AF464), goat anti Netrin G1a (NG1a) (1:100, R&D system AF1166), rabbit anti TBR1 (1:1000, Abcam ab31940), rabbit anti TBR2 (1:1000, Abcam ab23345), rabbit anti PAX6 (1:100, clone poly19013, Covance PRB-278P), rabbit anti GSH2 (1:2000, Millipore ABN162), chicken anti MAP2 (1:500, Novus, NB30213), mouse anti SMI-32 (1:500, Covance SMI-32R), and rat CTIP2 (1:1000, Abcam ab18465). Primary antibodies were incubated in PGT (PBS/gelatine 2g per L/TritonX-100 0.25%) with 0.1% glycine and 0.1% lysine for anti Netrin G1a antibodies and 3% BSA for anti SHH-Nter antibodies. Primary antibodies were revealed by immunofluorescence with the appropriate Alexa dye (Molecular Probes) or Cy3 (Jackson laboratories) conjugated secondary antibodies diluted in PGT (1:400). Bisbenzimide (1/5000 in PBT, Sigma) was used for nuclear counterstaining. Sections were mounted in mowiol/DABCO (25mg/mL) and imaged on a macroscope (MVX10 olympus) or an epifluorescence microscope (LEICA DM6000) using X40 and X63 immersion objectives, or a confocal microscope (Leica TCS SP5) using a x40 objective. Double immunostaining were performed when possible and images were representative of what was observed in more than 4 animals per genotype except for heterotopia. Macroscop images were treated to remove background around sections using wand tool in Adobe Photoshop 7.0 software. On epifluorescence and confocal microscop images, acquisition parameters and levels of intensity in Adobe Photoshop 7.0 software were similar for WT and Kif7 -/-. Merged images were acquired on the same section except for TBR1 with TBR2 acquired on adjacent sections. The fluorescence intensity along the depth of the cortex was assessed using the plot profile function of ImageJ under a 500 pixels wide line starting in the ventricle to the surface of the brain.

In situ hybridization

Specific antisense RNA probe for Shh, Gli1 and CxCl12 genes (gift of Marie-Catherine Tiveron) were used for in situ hybridization analyses. DIG-probes were synthesized with a labeling kit according to the manufacturer’s instructions (Roche, France). In situ hybridization was performed according to Tiveron et al 97 with few modifications (see Supplementary material). Sections were mounted on glass slides, dried, dehydrated in graded ethanol solutions, cleared in xylene and coverslipped with Eukitt.

RNAscope

Embryonic E14.5 heads were cuted in cold Leibovitz medium (Invitrogen) and immersion fixed in cold 4% (w/v) paraformaldehyde (PFA) in 0.12 m phosphate buffer, pH 7.4, overnight. Brains were then cryoprotected in PFA 4% /sucrose 10 %, embedded in gelatin 7.5% /sucrose 10% at 4°C and frozen. Biological samples were kept at −80°C until coronally sectioned at 20 μm with cryostat. RNAscope experiment was performed according manufacturer’s instructions (RNAScope® Multiplex Fluorescent V2 Assay, ACDbio) after a pre-treatment to remove gelatin. Two different probes/channels were used (C1 for Shh, C2 for Lhx-6). Probes were diluted according manufacturer’s instructions (1 Vol of C2 for 50 vol of C1) and dilution of the fluorophores (Opal 520 and 570) was 1:1500. Sections were mounted in mowiol/DABCO (25mg/mL) and imaged on a confocal microscop (Leica TCS SP5) using an immersion X10 or X63 objectives on 10 µm stacks of images (1 image /µm).

DiI experiments

Tiny crystals of DiI (1,1′-dioctadecyl-3,3,3′3′-tetramethylindocarbocyanine perchlorate) were inserted in the dorsal or lateral cortex of WT and Kif7 -/- embryonic brains fixed at E14.5 by immersion in 4% PAF. Injected brains were stored in 4% PAF at room temperature in the dark for several weeks to allow DiI diffusion. Before sectioning, brains were included in 3% agar. Coronal sections 60 μm thick were prepared with a vibratome and collected individually in 4% PAF. Sections were imaged with a macroscope (MVX10 Olympus).

Whole brain clearing and imaging

E14 and E16 embryos (WT, n=5; Kif7 -/-, n=6) were collected in cold L15 medium, perfused transcardially with 4% PAF using a binocular microscope. Brains were dissected and postfixed 3 days in 4% PAF at 4°C, and stored in PB at 4°C until clearing. All buffer solutions were supplemented with 0.01% Sodium Azide (Sigma- Aldrich). Whole brain immunostaining and clearing was performed at RT under gentle shaking according to a modified version of the original protocol 98. Briefly, perfused brains were dehydrated in graded methanol solutions (Fischer Scientific) (20, 50, 80% in ddH2O, 2 × 100%, 1h30 each), bleached overnight in 3% hydrogen peroxide (Sigma-Aldrich) in 100% methanol and rehydrated progressively. After PBS washes, brains were blocked and permeabilized 2 days in PBGT (0.2% gelatin (Merck), 0.5% triton X-100 (Sigma-Aldrich) in PBS). Brains were incubated 3 days at 37°C in primary goat anti Netrin G1a (1:100, R&D system AF1166) and rabbit anti TBR1 (1:1000, Abcam ab31940) antibodies. After 1 day of PBGT washes, brains were incubated 1 day at 37°C in secondary 647 donkey anti-goat (1:1000, Molecular probes,) and cy3 donkey anti-rabbit (1:1000, Jackson laboratories) antibodies. Immunolabeled brains were washed 1 day in PBS, embedded into 1.5% low-melting agarose (type VII, Sigma) in 1% ultra-pure Tris-acetate-EDTA solution, placed in 5-ml tubes (Eppendorf, 0030119452) and dehydrated 1h in each methanol baths (50%, 80% and 2 × 100%). Samples were then incubated for 3h in 33% methanol / 66% dichloromethane (DCM, Sigma-Aldrich), washed in 100% DCM (2×30 min) and incubated overnight (without shaking) in dibenzyl ether (DBE). Brains were stored in DBE at RT. Cleared samples were imaged on a light-sheet microscope (LaVision Biotec, x6.3 zoom magnification) equipped with a sCMOS camera (Andor Neo) and Imspector Microscope controller software. Imaris (Bitplane) was used for 3D reconstructions, snapshots and movies.

Brain slices

Brains of E14.5 WT and Kif7 -/- transgenic Nkx2.1-cre, tdTomato embryos were dissected in cold L15 medium, embedded in 3% type VII agarose and sectioned coronally with a manual slicer as explained in 29. Forebrain slices 250 µm thick were transferred in Millicell chambers and cultured for 4 hours in F12/DMEM medium supplemented with CFS 10% in a CO2 incubator (Merck Millipore) prior to pharmacological treatment and recording. Slices were incubated in drug for 2 hours before transfer in culture boxes equipped with a bottom glass coverslip for time-lapse imaging.

Pharmacological treatment

Either recombinant Mouse SHH (C25II), N-terminus (464-SH- 025, R&D systems) at 0.5µg/mL final or cyclopamine (C4116, Sigma) at 2.5 µM final were added to the culture medium of slices and renewed after 12 hours. Recombinant SHH was reconstituted at 100 µg/mL in sterile PBS with BSA 0.1%. The stock solution of cyclopamine was 10 mM in DMSO and the culture medium of control experiments contained 1/4000 DMSO (vehicle). Control experiments in culture medium with and without vehicle did not differ and were analyzed together.

Co-cultures

Co-cultures were performed on polylysine/laminin-coated glass coverslips fixed to the bottom of perforated Petri dishes in order to image migrating MGE cells. Brains were collected in cold PBS at embryonic day E14.5. Cortices and MGE explants were then dissected in cold Leibovitz medium (Invitrogen). Cortices were mechanically dissociated. Dissociated cortical cells were cultured on the coated glass coverslips and left in the incubator (37 °C, 5% CO2) for 1 hours. A wound divided the substrate in two-halves on which WT and Kif7 -/- MGE explants from littermate embryos were placed (Fig. 6A). Co-cultures were cultured for 24h and imaged.

Videomicroscopy and image processing

Slices and co-cultures were imaged on an inverted microscope (Leica DMI4000) equipped with a spinning disk (Roper Scientific, USA) and a temperature-controlled chamber. Multi-position acquisition was performed with a Coolsnap HQ camera (Roper Scientific, USA) to allow the recording of the whole cortex. Images were acquired with a X20 objective (LX20, Fluotar, Leica, Germany) and a 561 nm laser (MAG Biosystems, Arizona). Z-stacks of 30 μm were acquired 50 µm away from the slice surface, with a step size of 2 μm and a time interval of 2 or 5 minutes for at least 21 hours. Acquisitions were controlled using the Metamorph software (Roper Scientific, USA). Cell trajectories were reconstructed on movies by tracking manually the cell rear with MTrakJ (ImageJ plugin, NIH, USA) and were clustered according to their position in cortical layers and to their orientation (see details in Fig. 6). The migration speed of cells, the frequency and duration of pauses were extracted from tracking data using excel macros. The directionality of cells was analysed along time using a macro from 99.

Study design

WT group was compared to Kif7 -/- group and to pharmacological treatment groups for brain slice experiments. The experimental unit was a single animal for immunohistochemistry analysis and individual cell for videomicroscopy analysis.

Statistical analysis

All data were obtained from at least three independent experiments and are presented as mean ± SEM (standard error of mean). Statistical analyses were performed with the GraphPad Prism software or R. Statistical significance of the data was evaluated using the unpaired two-tailed t test, the Mann–Whitney test, the Chi2 test or the Two-way ANOVA test followed by a post-hoc test. Data distribution was tested for normality using the D’Agostino and Pearson omnibus normality test. Values of P<0.05 were considered significant. In figures, levels of significance were expressed by * for P<0.05, ** for P<0.01, *** for P<0.001 and **** for P<0.0001.